### 2021-04-07 Sleep Review Paper v1.Rmd
Uploaded first draft of R markdown

---
title             : "Study Sample Characteristics in the Sleep and Chronobiology Research: A Systematic Review"
shorttitle        : "Demographics analysis of sleep and chronobiology research"

author: 
  - name          : "Selma Tir"
    affiliation   : "1,2"
    email         : "selma.tir@ndcn.ox.ac.uk"
    role:
      - wrote the original draft protocol, reviewed and scored articles, reviewed and edited the manuscript
  - name          : "Rhiannon White"
    affiliation   : "1,3"
    email         : "r.white.5@warwick.ac.uk"
    role:         
      - reviewed and scored articles, reviewed and edited the manuscript
  - name          : "Manuel Spitschan"
    affiliation   : "1,4,5"
    corresponding : yes    # Define only one corresponding author
    email         : "manuel.spitschan@psy.ox.ac.uk"
    role:
      - wrote the original draft protocol, reviewed and edited the manuscript

affiliation:
  - id            : "1"
    institution   : "Department of Experimental Psychology, University of Oxford, Anna Watts Building, Radcliffe Observatory Quarter, Oxford, OX2 6GG, United Kingdom"
  - id            : "2"
    institution   : "Nuffield Department of Clinical Neurosciences, University of Oxford"
  - id            : "3"
    institution   : "Warwick Medical School, University of Warwick, United Kingdom"
  - id            : "4"
    institution   : "Centre for Chronobiology, Psychiatric Hospital of the University of Basel, Switzerland"
  - id            : "5"
    institution   : "Transfaculty Research Platform Molecular and Cognitive Neurosciences, University of Basel, Switzerland"

authornote: |
  This systematic review is funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).

abstract: |
  Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. In this systematic review, we evaluated the study sample characteristics in the published sleep and chronobiology research over the past 40 years.

  Articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals, identified by their five-year Impact Factor, were found through MEDLINE. 6,777 articles were initially included for screening. Inclusion requirements included conducting original research, reporting human data, and recruiting volunteers. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0=not reported), and any reported aggregate summary statistics for these variables were recorded. Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed including any of the demographic variables as covariates.

  ~75% of screened articles met inclusion criteria. While >90% of studies reported age or sex, all other variables were reported in <10% of cases. We found that sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s. Overall, ~75% of studies recruited both male and female participants. Of studies recruiting a single sex, ~50% all-female studies focused on a sex-dependent feature, compared to <5% in all-male studies.

  In this comprehensive review, we found that the majority of studies report at least sex or age, while many other important variables are typically not reported. Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate disaggregated data analyses.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Study Sample Characteristics, Demographics, Sex, Systematic Review"
wordcount         : "X"

bibliography      : ["r-references.bib"]

floatsintext      : no
figurelist        : yes
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : yes

documentclass     : "apa6"
classoption       : "jou"
output            : papaja::apa6_pdf
---

```{r setup, include=FALSE}
# Version control
## R version 4.0.4 (2021-02-15)
## R Studio version 1.4.1106
## Macintosh; Intel Mac OS X 10.13.6

# Load packages
library("papaja")
library(corrplot)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggExtra)
library(ggsci)
library(graphics)
library(gridExtra)
library(knitr)
library(latticeExtra)
library(mosaic)
library(pracma)
library(readr)
library(readxl)
library(reshape2)
library(stats) 
library(stringr)
library(tibble)
library(tidyr)
library(lemon)
r_refs("r-references.bib")

# Set working directory
setwd("~/Sleep Review")

# Download scoring sheet 
data.original <- read_excel("Sleep Review Scoring Sheet Test.xlsx")
```

# Introduction

  Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. In this systematic review, we evaluated the study sample characteristics in the published sleep and chronobiology research over the past 40 years.

# Methods

  Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals  were found through MEDLINE by MS. The eight journals were selected based on their five-year Impact Factor, and included Journal of Pineal Research, Sleep, Journal of Sleep Research, Sleep Medicine, Journal of Clinical Medicine, Journal of Biological Rhythms, Behavioral Sleep Medicine, and Chronobiology International. 6,777 articles were initially sampled for screening. Inclusion requirements included conducting original research in the English language, reporting human data, and recruiting volunteers. As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, erratums and corrigendums were excluded. All articles were reviewed for eligibility and coded by ST. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded. Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.

```{r Inclusion & Exclusion, echo=FALSE, message=FALSE, warning=FALSE}
# Filter reviewed articles
data.original <- data.original %>%
  filter(article_fulltext_available_yn == 1)

# Separate included and excluded articles
data.included <- data.original %>%
  filter(article_included_yn == 1)

data.excluded <- data.original %>%
  filter(article_included_yn == 0)

# Check that the number of included and excluded articles is equal to the total (return TRUE if yes)
check.article.number <- (as.numeric(nrow(data.included)) + as.numeric(nrow(data.excluded))) == as.numeric(nrow(data.original))

# Compute percentage of included vs excluded articles, and round to two decimals
percent.included <- round(as.numeric(nrow(data.included))/as.numeric(nrow(data.original)) * 100, 2)
percent.excluded <- round(as.numeric(nrow(data.excluded))/as.numeric(nrow(data.original)) * 100, 2)

# Compute percentage of articles that were excluded for other reasons
data.excluded.other <- data.excluded %>%
  filter(!is.na(article_excluded_reason_other))
percent.excluded.other <- round(nrow(data.excluded.other)/nrow(data.excluded) * 100, 2)

# Compute percentage of excluded articles by reason for exclusion
data.eligibility <- data.original
data.eligibility$article_excluded_reason <- tolower(data.eligibility$article_excluded_reason)

data.eligibility <- data.eligibility %>%
  mutate(eligibility=if_else(!is.na(article_excluded_reason_other), "Not included for other reasons", 
                          if_else(article_included_yn == 1, "Included",
                                  if_else(article_excluded_reason == "not recruiting volunteers", 
                                          "Not recruiting volunteers",
                                          if_else(article_excluded_reason == "not an original research article",
                                                  "Not an original research article",
                                                  if_else(article_excluded_reason == "not reporting human data",
                                                          "Not reporting human data",
                                                          "NA"))))))

data.eligibility <- data.eligibility %>%
  count(eligibility, article_year) %>%
  ungroup()

data.eligibility <- data.eligibility %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(eligibility) %>%
  mutate(per=round(100*n/total, 2))

data.eligibility$article_year <- as.numeric(data.eligibility$article_year)
data.eligibility$per <- as.numeric(data.eligibility$per)

# Stacked bar chart of excluded articles by reason for exclusion across the years
figure1B <- ggplot(data.eligibility, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(eligibility, levels=c("Not an original research article",
                                                 "Not reporting human data",
                                                 "Not included for other reasons",
                                                 "Not recruiting volunteers",
                                                 "NA",
                                                 "Included")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Eligibility") +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_manual(values=c("#29AF7FFF","#482677FF","#D8DE29FF","#FDE725FF","#2D708EFF")) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12), legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))
```

```{r Funding, include=FALSE}
# Select articles that report funding
data.funding.name <- data.included %>%
  filter(funding_reported_yn == 1)

# Compute percentage of articles that report funding
percent.funding.name <- round(nrow(data.funding.name)/as.numeric(nrow(data.included)) * 100, 2)

# Select articles that report funding number
data.funding.num <- data.included %>%
  filter(funding_refnum_yn == 1)

# Compute percentage of articles that report funding number, out of those that report funding
percent.funding.num <- round(nrow(data.funding.num)/as.numeric(nrow(data.funding.name)) * 100, 2)

list.funding <- tolower(data.included$funding_names)

# Split the category combinations and convert to lower case letters
list.funding <- as.list(strsplit(list.funding, ","))
list.funding <- unlist(list.funding)

# Count number of occurrences for each category
table.list.funding <- as.data.frame(table(list.funding))
n.funding.nih <- sum(str_count(list.funding, "nih"), str_count(list.funding, "national institute"))

# Compute percentage of articles that report funding from the National Institutes in the US
data.funding.NI <- data.funding.name %>%
  filter(grepl("NI", funding_names) | grepl("National Institute", funding_names))

percent.funding.NI <- round(nrow(data.funding.NI)/as.numeric(nrow(data.included)) * 100, 2)

# Select articles that report funding number and funding from NIH in the US
data.funding.NI.num <- data.funding.name %>%
  filter(grepl("NI", funding_names) | grepl("National Institute", funding_names)) %>%
  filter(funding_refnum_yn == 1)

percent.funding.NI.num <- round(nrow(data.funding.NI.num)/as.numeric(nrow(data.funding.NI)) * 100, 2)
```

```{r Study Country, echo=FALSE, message=FALSE, warning=FALSE}
# Set country code to upper case
data.included$study_country_values <- toupper(data.included$study_country_values)

# Separate articles for which country is reported explicitly vs inferred
data.country.reported <- data.included %>%
  filter(study_country_reported_explicitly_yn == 1)

data.country.inferred <- data.included %>%
  filter(study_country_reported_inferred_yn == 1)

# Compute percentage of articles where country is reported explicitly vs inferred
percent.country.reported <- round(nrow(data.country.reported)/nrow(data.included) * 100, 2)
percent.country.inferred <- round(nrow(data.country.inferred)/nrow(data.included) * 100, 2)

# Separate articles that report one or multiple countries of study
country.single <- data.included %>%
  filter(!grepl(",", study_country_values)) %>%
  group_by(study_country_values) %>%
  tally() %>%
  mutate(n=round((n/as.numeric(nrow(data.included))*100),1))

country.multiple <- data.included %>%
  filter(grepl(",", study_country_values)) %>%
  group_by(study_country_values) %>%
  tally() %>%
  mutate(n=round((n/as.numeric(nrow(data.included))*100),1))

# Compute percentage of articles where multiple countries of study are reported
percent.country.multiple <- round(nrow(country.multiple)/nrow(data.included) * 100, 2)

# Examine countries that are the most represented overall for studies that report a single country
country.single.maj <- country.single %>%
  filter(n>4)

pc.country.single.maj <- ggplot(data=country.single.maj, aes(x="", y=n, fill=study_country_values)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  geom_text(aes(label=paste0(n, "%")), position=position_stack(vjust=0.5)) +
  ggtitle("Countries most represented overall") +
  labs(fill="Country") +
  scale_fill_brewer(palette="RdBu") +
  theme_void()

# Compute number of articles for the 8 most represented countries by year of publication
# Select articles where a single country is reported or inferred
data.country <- data.included %>%
  filter(!grepl(",", study_country_values)) %>%
  count(study_country_values, article_year) %>%
  ungroup()

# we checked what were the 8 most represented countries overall in the pie chart above

# Change all countries that are not the 8 most represented countries to 'Other'
data.country$study_country_values[data.country$study_country_values != "AUS" &
                                    data.country$study_country_values != "CAN" &
                                    data.country$study_country_values != "DEU" &
                                    data.country$study_country_values != "FRA" &
                                    data.country$study_country_values != "GBR" &
                                    data.country$study_country_values != "USA" &
                                    data.country$study_country_values != "ITA" &
                                    data.country$study_country_values != "JPN"] <- "Other"

# Sum rows with 'Other' by year
data.country <- aggregate(cbind(n) ~ study_country_values + article_year, data=data.country, FUN=sum)

# Compute proportion of articles per country by year of publication
data.country <- data.country %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(study_country_values) %>%
  mutate(per=round(100*n/total, 2))

data.country$article_year <- as.numeric(data.country$article_year)
data.country$per <- as.numeric(data.country$per)

# Stacked bar chart of 8 most represented study countries by year of publication
figure2B <- ggplot(data.country, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(study_country_values, levels=c("Other","AUS","CAN","DEU","FRA",
                                                          "GBR","ITA","JPN","USA")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Country") +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_viridis_d(direction=-1) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12), legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))
```

```{r Number of sub-studies, echo=FALSE, message=FALSE, warning=FALSE}
# Compute percentage of articles per number of sub-studies
num.study <- data.included %>%
  group_by(num_substudies) %>%
  tally() %>%
  mutate(num_substudies=as.factor(num_substudies),
         n=round((n/as.numeric(nrow(data.included))*100),1))

percent.study.single <- num.study$n[1]
percent.study.multiple <- 1 - num.study$n[1]

### The rest  of the analysis will be conducted on single studies
# Filter single studies
data <- data.included %>%
  filter(num_substudies == 1)
```

  Data were coded in an Excel Spreadsheet and analyzed in R Studio (version 4.0.4). Reporting of funding, geographical location, and number of sub-studies for each article were investigated for the sample of articles that passed all eligibility criteria. The rest of the analysis was conducted on the subset of articles that reported a single study (`r round(percent.study.single)`% of articles), i.e. a single sample population.

```{r Sample size, echo=FALSE, message=FALSE, warning=FALSE}
# Compute percentage of studies that report sample size
data.ss <- data %>%
  filter(sample_size_bycategory_reported_yn == 1)

percent.ss <- round(nrow(data.ss)/nrow(data) * 100, 2)

# Violin plot of sample size by year of publication with log10 scaling
data.ss$article_year <- as.factor(data.ss$article_year)
data.ss$sample_size_total_reported_values <- as.numeric(data.ss$sample_size_total_reported_values)

figure3 <- ggplot(data.ss, aes(x=article_year, y=sample_size_total_reported_values)) +
  geom_violin(fill="#440154FF") +
  theme_apa() +
  scale_y_continuous(trans='log10', breaks=c(10,1000,100000), labels=c("10","1,000","100,000"), expand=c(0,0)) +
  scale_x_discrete(breaks=c("1979","1989","1999","2009","2019")) +
  xlab("Year of publication") +
  ylab("Sample size, log-10 scaling") +
  annotation_logticks(sides="l") +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
```

```{r Age, echo=FALSE, message=FALSE, warning=FALSE}
# Select articles that report age
data.age <- data.included %>%
  filter(age_reported_mean_yn == 1 | age_reported_median_yn == 1 | age_reported_min_yn | age_reported_max_yn | age_reported_other_yn == 1 | age_reported_by_sex_yn == 1)

# Compute percentage of articles where age is reported
percent.age <- round(nrow(data.age)/nrow(data.included) * 100, 2)

## Correlate the reporting of age variables by year
# select year and age columns
data.age.year <- data.age %>%
  select(article_year, age_reported_mean_yn, age_reported_sd_yn, age_reported_median_yn, 
         age_reported_min_yn, age_reported_max_yn, age_reported_iqr_yn, age_reported_other_yn)

# Replace 0 by -1
data.age.year$age_reported_mean_yn[data.age.year$age_reported_mean_yn == 0] <- -1
data.age.year$age_reported_sd_yn[data.age.year$age_reported_sd_yn == 0] <- -1
data.age.year$age_reported_median_yn[data.age.year$age_reported_median_yn == 0] <- -1
data.age.year$age_reported_min_yn[data.age.year$age_reported_min_yn == 0] <- -1
data.age.year$age_reported_max_yn[data.age.year$age_reported_max_yn == 0] <- -1
data.age.year$age_reported_iqr_yn[data.age.year$age_reported_iqr_yn == 0] <- -1
data.age.year$age_reported_other_yn[data.age.year$age_reported_other_yn == 0] <- -1

# Replace NAs
data.age.year[is.na(data.age.year)] <- -1

# Group by year and compute the average for each sex variable
age.corr.year <- data.age.year %>%
  group_by(article_year) %>%
  summarise(mean=mean(age_reported_mean_yn), sd=mean(age_reported_sd_yn), median=mean(age_reported_median_yn),
            min=mean(age_reported_min_yn), max=mean(age_reported_max_yn), iqr=mean(age_reported_iqr_yn),
            other=mean(age_reported_other_yn))  %>%
  ungroup()

age.corr.year <- age.corr.year %>%
  group_by(article_year) %>%
  mutate(total=((mean+sd+median+min+max+iqr+other)/7))

# Reshape data
age.corr.year <- melt(age.corr.year,id.vars='article_year',measure.vars=c('mean','sd','median','min','max','iqr','other','total'))

# Heat map of reporting of age variables by year of publication
figure4A <- ggplot(data=age.corr.year, aes(x=variable, y=article_year, fill=value)) +
  geom_tile() +
  theme_apa() +
  ylab(label="Year") +
  xlab("Age variable") +
  scale_fill_viridis_c(end=0.6, direction=-1, labels=c("0.0","0.25","0.5","0.75","1")) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8), legend.position="none",
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10),
        legend.title=element_blank())

## Correlate the reporting of age variables by journal
# Select journal and age columns
data.age.journal <- data.age %>%
  select(article_journal, age_reported_mean_yn, age_reported_sd_yn, age_reported_median_yn, 
         age_reported_min_yn, age_reported_max_yn, age_reported_iqr_yn, age_reported_other_yn)

# Replace 0 by -1
data.age.journal$age_reported_mean_yn[data.age.journal$age_reported_mean_yn == 0] <- -1
data.age.journal$age_reported_sd_yn[data.age.journal$age_reported_sd_yn == 0] <- -1
data.age.journal$age_reported_median_yn[data.age.journal$age_reported_median_yn == 0] <- -1
data.age.journal$age_reported_min_yn[data.age.journal$age_reported_min_yn == 0] <- -1
data.age.journal$age_reported_max_yn[data.age.journal$age_reported_max_yn == 0] <- -1
data.age.journal$age_reported_iqr_yn[data.age.journal$age_reported_iqr_yn == 0] <- -1
data.age.journal$age_reported_other_yn[data.age.journal$age_reported_other_yn == 0] <- -1

# Replace NAs
data.age.journal[is.na(data.age.journal)] <- -1

# Group by journal and compute the average for each sex variable
age.corr.journal <- data.age.journal %>%
  group_by(article_journal) %>%
  summarise(mean=mean(age_reported_mean_yn), sd=mean(age_reported_sd_yn), median=mean(age_reported_median_yn),
            min=mean(age_reported_min_yn), max=mean(age_reported_max_yn), iqr=mean(age_reported_iqr_yn),
            other=mean(age_reported_other_yn)) %>%
  ungroup()

age.corr.journal <- age.corr.journal %>%
  group_by(article_journal) %>%
  mutate(total=((mean+sd+median+min+max+iqr+other)/7))

# Reshape data
age.corr.journal <- melt(age.corr.journal,id.vars='article_journal',
                         measure.vars=c('mean','sd','median','min','max','iqr','other','total'))

# Heat map of reporting of age variables by journal of publication
figure4B <- ggplot(data=age.corr.journal, aes(x=variable, y=article_journal, fill=value)) +
  geom_tile() +
  theme_apa() +
  ylab(label="Journal") +
  xlab("Age variable") +
  scale_fill_viridis_c(end=0.6, direction=-1, labels=c("0.0","0.25","0.5","0.75","1")) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8), legend.position="right",
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10),
        legend.title=element_blank())

## Regression for age by year of publication
data.age.mean <- data.age %>%
  filter(age_reported_mean_yn == 1) %>%
  filter(!is.na(as.numeric(age_reported_mean_value)))

data.age.mean$age_reported_mean_value <- as.numeric(data.age.mean$age_reported_mean_value)

figure4C <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
  geom_smooth(method="lm", color="#440154FF") +
  theme_apa() +
  ylab("Mean age") +
  xlab("Year") +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(limits=c(20,60), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8)) 
```

```{r Sex, echo=FALSE, message=FALSE, warning=FALSE}
# Select sex variables       
data.sex <- data.included %>%
  select(article_year, sex_reported_yn, 
         sex_reported_num_male, sex_reported_percent_male,
         sex_reported_num_female,sex_reported_percent_female)

# Compute whether males or females are reported at all
data.sex$male <- ifelse(is.na(data.sex$sex_reported_num_male) & is.na(data.sex$sex_reported_percent_male),
                             "No",
                             "Yes")
data.sex$male <- as.factor(data.sex$male)
data.sex$female <- ifelse(is.na(data.sex$sex_reported_num_female) & is.na(data.sex$sex_reported_percent_female),
                             "No",
                             "Yes")
data.sex$female <- as.factor(data.sex$female)

# Compute whether articles report males, females, both or unspecified
data.sex <- data.sex %>%
  mutate(report=if_else(male == "Yes" & female == "Yes", "Both Sexes", 
                          if_else(male == "Yes" & female == "No", "Male",
                                  if_else(male == "No" & female == "Yes", "Female", "Unspecified"))))

# Count number of articles for each sex condition by year of publication
data.sex <- data.sex %>%
  count(report, article_year) %>%
  ungroup()

# Compute proportion of articles per sex category by year of publication
data.sex <- data.sex %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(report, add=TRUE) %>%
  mutate(per=round(100*n/total,2))

data.sex$article_year <- as.numeric(data.sex$article_year)
data.sex$per <- as.numeric(data.sex$per)

# Stacked bar chart of sex reporting by year of publication
figure5A <- ggplot(data.sex, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(report, levels=c("Female",
                                            "Unspecified",
                                            "Both Sexes",
                                            "Male")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Sex") +
  scale_fill_manual(values=c("#482677FF","#D8DE29FF","#29AF7FFF","#2D708EFF")) +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

# Select articles for which sex is reported
data.sex.reported <- data.included %>%
  filter(sex_reported_yn == 1) %>%
  mutate(sample_size_total_reported_values=as.numeric(sample_size_total_reported_values),
         sex_reported_num_male=as.numeric(sex_reported_num_male),
         sex_reported_percent_male=as.numeric(sex_reported_percent_male),
         sex_reported_num_female=as.numeric(sex_reported_num_female),
         sex_reported_percent_female=as.numeric(sex_reported_percent_female))

# Select articles for which intersex is reported
data.intersex <- data.included %>%
  filter(sex_reported_intersex_yn == 1)

# compute percentage of articles where sex and intersex is reported
percent.sex <- round(nrow(data.sex.reported)/nrow(data.included) * 100, 2)
percent.intersex <- round(nrow(data.intersex)/nrow(data.included) * 100, 2)

# Compute whether males or females are recruited in the articles that report  sex
data.sex.reported$male <- ifelse(is.na(data.sex.reported$sex_reported_num_male) &
                                   is.na(data.sex.reported$sex_reported_percent_male),
                             "No",
                             "Yes")
data.sex.reported$male <- as.factor(data.sex.reported$male)
data.sex.reported$female <- ifelse(is.na(data.sex.reported$sex_reported_num_female) &
                                     is.na(data.sex.reported$sex_reported_percent_female),
                             "No",
                             "Yes")
data.sex.reported$female <- as.factor(data.sex.reported$female)

# Compute whether articles report males, females or both
data.sex.reported <- data.sex.reported %>%
  mutate(report=if_else(male == "Yes" & female == "Yes", "Both Sexes", 
                          if_else(male == "Yes" & female == "No", "Male",
                                  if_else(male == "No" & female == "Yes", "Female", "NA"))))

# Compute percentage of articles that only recruit males or females
data.sex.male <- data.sex.reported %>%
  filter(report=="Male")

percent.sex.male <- round(nrow(data.sex.male)/nrow(data.sex.reported) * 100, 2)

data.sex.female <- data.sex.reported %>%
  filter(report=="Female")

percent.sex.female <- round(nrow(data.sex.female)/nrow(data.sex.reported) * 100, 2)

# Compute percentage of articles reporting on a sex-dependent feature
data.sex.male.focus <- data.sex.male %>%
  filter(sample_focus_sex_dependent_yn == 1)

percent.sex.male.focus <- round(nrow(data.sex.male.focus)/nrow(data.sex.male) * 100, 2)

data.sex.female.focus <- data.sex.female %>%
  filter(sample_focus_sex_dependent_yn == 1)

percent.sex.female.focus <- round(nrow(data.sex.female.focus)/nrow(data.sex.female) * 100, 2)

# Compute percentage of studies reporting age by sex
data.sex.age <- data.included %>%
  filter(age_reported_by_sex_yn == 1)

percent.sex.age <- round(nrow(data.sex.age)/nrow(data.included) * 100, 2)
```

```{r Other demographics, echo=FALSE, message=FALSE, warning=FALSE}
# Select single studies and other demographic variables
data.demo <- data.included %>%
  select(article_year, ethnicity_reported_yn, education_reported_mean_yn, education_reported_yn,
         profession_reported_yn, ses_reported_yn)

# Compute the reporting of the different demographic variables
data.demo <- data.demo %>%
  mutate(education=if_else(education_reported_mean_yn == 0 & education_reported_yn == 0, "No", "Yes"),
         ethnicity=if_else(ethnicity_reported_yn == 1, "Yes", "No"),
         profession=if_else(profession_reported_yn == 1, "Yes", "No"),
         ses=if_else(ses_reported_yn == 1, "Yes", "No"))

# Count number of articles for each demographic variable by year of publication
data.demo.education <- data.demo %>%
  count(education, article_year) %>%
  ungroup()

data.demo.ethnicity <- data.demo %>%
  count(ethnicity, article_year) %>%
  ungroup()

data.demo.profession <- data.demo %>%
  count(profession, article_year) %>%
  ungroup()

data.demo.ses <- data.demo %>%
  count(ses, article_year) %>%
  ungroup()

# Compute proportion of articles by year of publication
data.demo.education <- data.demo.education %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(education, add=TRUE) %>%
  mutate(per=round(100*n/total,2),
         variable="edu") %>%
  select(article_year, variable, education, per)
colnames(data.demo.education)[3] <- "report"

data.demo.ethnicity <- data.demo.ethnicity %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(ethnicity, add=TRUE) %>%
  mutate(per=round(100*n/total,2),
         variable="ethni") %>%
  select(article_year, variable, ethnicity, per)
colnames(data.demo.ethnicity)[3] <- "report"

data.demo.profession <- data.demo.profession %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(profession, add=TRUE) %>%
  mutate(per=round(100*n/total,2),
         variable="prof") %>%
  select(article_year, variable, profession, per)
colnames(data.demo.profession)[3] <- "report"

data.demo.ses <- data.demo.ses %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(ses, add=TRUE) %>%
  mutate(per=round(100*n/total,2),
         variable="ses") %>%
  select(article_year, variable, ses, per)
colnames(data.demo.ses)[3] <- "report"

# Merge back the datasets together
merge1 <- rbind(data.demo.education, data.demo.ethnicity)
merge2 <- rbind(merge1, data.demo.profession)
data.demo <- rbind(merge2, data.demo.ses)

data.demo$article_year <- as.numeric(data.demo$article_year)
data.demo$per <- as.numeric(data.demo$per)

# Stacked bar chart  of reporting of other demographic variables by year of publication
figure6.bars <- ggplot(data.demo, aes(x=variable, y=per, fill=report)) +
  geom_bar(stat="identity") +
  facet_grid(~article_year) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Variable reported") +
  scale_fill_manual(values=c("#482677FF","#D8DE29FF")) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"),
        axis.text.x=element_text(size=5, angle=45, hjust=1),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10),
        panel.spacing=unit(0.1, "lines"))

data.demo.reported <- data.demo %>%
  filter(report == "Yes")

figure6.points <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
  geom_point(size=2.5) +
  geom_line(linetype="dashed") +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  scale_color_viridis_d(name="Variable", labels=c("Education","Ethnicity","Profession","Socio-economic status")) +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(limits=c(0,100), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

# Compute percentage of articles that report education
data.education <- data.included %>%
  filter(education_reported_mean_yn == 1 | education_reported_yn == 1)

percent.education <- round(nrow(data.education)/nrow(data.included) * 100, 2)

# Compute distribution of the number of categories reported for education
data.education$education_reported_numcategory <- as.numeric(data.education$education_reported_numcategory)

figure7A <- ggplot(data=data.education, aes(x=education_reported_numcategory)) +
  geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
  geom_vline(aes(xintercept=mean(education_reported_numcategory, na.rm=TRUE)), color="#D8DE29FF", size=1) +
  theme_apa() +
  ylab("Density") +
  xlab("Education") +
  scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))

# Compute percentage of articles that report ethnicity
data.ethnicity <- data.included %>%
  filter(ethnicity_reported_yn == 1)

percent.ethnicity <- round(nrow(data.ethnicity)/nrow(data.included) * 100, 2)

# Compute distribution of the number of categories reported for ethnicity
data.ethnicity$ethnicity_reported_numcategory <- as.numeric(data.ethnicity$ethnicity_reported_numcategory)

figure7B <- ggplot(data=data.ethnicity, aes(x=ethnicity_reported_numcategory)) +
  geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
  geom_vline(aes(xintercept=mean(ethnicity_reported_numcategory, na.rm=TRUE)), color="#D8DE29FF", size=1) +
  theme_apa() +
  ylab("Density") +
  xlab("Ethnicity") +
  scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))

# Compute percentage of articles that report profession
data.profession <- data.included %>%
  filter(profession_reported_yn == 1)

percent.profession <- round(nrow(data.profession)/nrow(data.included) * 100, 2)

# Compute distribution of the number of categories reported for profession
data.profession$profession_reported_numcategory <- as.numeric(data.profession$profession_reported_numcategory)

figure7C <- ggplot(data=data.profession, aes(x=profession_reported_numcategory)) +
  geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
  geom_vline(aes(xintercept=mean(profession_reported_numcategory, na.rm=TRUE)), color="#D8DE29FF", size=1) +
  theme_apa() +
  ylab("Density") +
  xlab("Profession") +
  scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))

# Compute percentage of articles that report ses
data.ses <- data.included %>%
  filter(ses_reported_yn ==1)

percent.ses <- round(nrow(data.ses)/nrow(data.included) * 100, 2)

# Compute distribution of the number of categories reported for ses
data.ses$ses_reported_numcategory <- as.numeric(data.ses$ses_reported_numcategory)

figure7D <- ggplot(data=data.ses, aes(x=ses_reported_numcategory)) +
  geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
  geom_vline(aes(xintercept=mean(ses_reported_numcategory, na.rm=TRUE)), color="#D8DE29FF", size=1) +
  theme_apa() +
  ylab("Density") +
  xlab("SES") +
  scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8)) 
```

```{r Focus, echo=FALSE, message=FALSE, warning=FALSE}
## Sex feature
# Compute percentage of articles focusing on a sex dependent feature
data.focus.sex <- data.included %>%
  filter(sample_focus_sex_dependent_yn == 1)

percent.focus.sex <- round(nrow(data.focus.sex)/nrow(data.included) * 100, 2)

# Compute number of articles per sex dependent feature 
sex.dependent <- data.focus.sex %>%
  group_by(sample_focus_sex_dependent) %>%
  tally() %>%
  mutate(n=round((n/nrow(data.focus.sex)*100),1))

# Pie chart of the reported sex dependent features 
pc.sex.dependent <- ggplot(data=sex.dependent, aes(x="", y=n, fill=sample_focus_sex_dependent)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  geom_text(aes(label=paste0(n, "%")), position=position_stack(vjust=0.5)) +
  ggtitle("Articles focused on a sex dependent feature") +
  labs(fill="Sex dependent feature") +
  scale_fill_brewer(palette="RdBu") +
  theme_apa()

## Clinical feature
# Compute percentage of articles focusing on a clinical feature
data.focus.clinical <- data.included %>%
  filter(sample_focus_clinical_yn == 1)

percent.focus.clinical <- round(nrow(data.focus.clinical)/nrow(data.included) * 100, 2)

# Compute number of articles per clinical feature 
clinical <- data.focus.clinical %>%
  group_by(sample_focus_clinical) %>%
  tally() %>%
  mutate(n=round((n/nrow(data.focus.clinical)*100),1))

# Pie chart of the reported clinical features 
pc.clinical <- ggplot(data=clinical, aes(x="", y=n, fill=sample_focus_clinical)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  geom_text(aes(label=paste0(n, "%")), position=position_stack(vjust=0.5)) +
  ggtitle("Articles focused on a clinical feature") +
  labs(fill="Clinical feature") +
  scale_fill_brewer(palette="RdBu") +
  theme_void()

## Twins
# Compute percentage of articles focusing on twins
data.focus.twins <- data.included %>%
  filter(sample_focus_twins == 1)

percent.focus.twins <- round(nrow(data.focus.twins)/nrow(data.included) * 100, 2)

## University students
# Compute percentage of articles focusing on university students
data.focus.students <- data.included %>%
  filter(sample_focus_university_students == 1)

percent.focus.students <- round(nrow(data.focus.students)/nrow(data.included) * 100, 2)

## Shift workers
# Compute percentage of articles focusing on shift workers
data.focus.shift <- data.included %>%
  filter(sample_focus_shift_workers == 1)

percent.focus.shift <- round(nrow(data.focus.shift)/nrow(data.included) * 100, 2)

## Pregnant women
# Compute percentage of articles focusing on pregnant women
data.focus.pregnant <- data.included %>%
  filter(sample_focus_pregnant_women == 1)

percent.focus.pregnant <- round(nrow(data.focus.pregnant)/nrow(data.included) * 100, 2)
```

```{r Disaggregation, echo=FALSE, warning=FALSE, message=FALSE}
# Select articles that report no difference in results based on age vs use age as a covariate
data.diff.age <- data.included %>%
  filter(analysis_no_difference_age == 1)

data.dis.age <- data.included %>%
  filter(analysis_disaggregation_age == 1)

# Select articles that report no difference in results based on sex vs use sex as a covariate
data.diff.sex <- data.included %>%
  filter(analysis_no_difference_sex == 1)

data.dis.sex <- data.included %>%
  filter(analysis_disaggregation_sex == 1)

# Select articles that report no difference in results based on ethnicity vs use ethnicity as a covariate
data.diff.ethnicity <- data.included %>%
  filter(analysis_no_difference_ethnicity == 1)

data.dis.ethnicity <- data.included %>%
  filter(analysis_disaggregation_ethnicity == 1)

# Select articles that report no difference in results based on education vs use education as a covariate
data.diff.education <- data.included %>%
  filter(analysis_no_difference_education == 1)

data.dis.education <- data.included %>%
  filter(analysis_disaggregation_education == 1)

# Select articles that report no difference in results based on profession vs use profession as a covariate
data.diff.profession <- data.included %>%
  filter(analysis_no_difference_profession == 1)

data.dis.profession <- data.included %>%
  filter(analysis_disaggregation_profession == 1)

# Select articles that report no difference in results based on ses vs use ses as a covariate
data.diff.ses <- data.included %>%
  filter(analysis_no_difference_ses == 1)

data.dis.ses <- data.included %>%
  filter(analysis_disaggregation_ses == 1)

# Select single studies and disaggregation variables
data.analysis <- data.included %>%
  select(article_year, analysis_disaggregation_age, analysis_disaggregation_sex, analysis_disaggregation_ethnicity, 
         analysis_disaggregation_education, analysis_disaggregation_profession, analysis_disaggregation_ses)

# Compute whether articles use ddemographic variables in the analysis
data.analysis <- data.analysis %>%
  mutate(report=if_else(analysis_disaggregation_age == 1 & analysis_disaggregation_sex == 0 & 
                            analysis_disaggregation_ethnicity == 0 & analysis_disaggregation_education == 0 &
                            analysis_disaggregation_profession == 0 & analysis_disaggregation_ses == 0 , "Age", 
                          if_else(analysis_disaggregation_age == 0 & analysis_disaggregation_sex == 1 & 
                            analysis_disaggregation_ethnicity == 0 & analysis_disaggregation_education == 0 &
                            analysis_disaggregation_profession == 0 & analysis_disaggregation_ses == 0 , "Sex",
                          if_else(analysis_disaggregation_age == 1 & analysis_disaggregation_sex == 1 & 
                            analysis_disaggregation_ethnicity == 0 & analysis_disaggregation_education == 0 &
                            analysis_disaggregation_profession == 0 & analysis_disaggregation_ses == 0 , "Age & Sex",
                          if_else(analysis_disaggregation_age == 1 & analysis_disaggregation_sex == 0 & (
                            analysis_disaggregation_ethnicity == 1 | analysis_disaggregation_education == 1 |
                            analysis_disaggregation_profession == 1 | analysis_disaggregation_ses == 1), "Age +1",
                          if_else(analysis_disaggregation_age == 0 & analysis_disaggregation_sex == 1 & (
                            analysis_disaggregation_ethnicity == 1 | analysis_disaggregation_education == 1 |
                            analysis_disaggregation_profession == 1 | analysis_disaggregation_ses == 1), "Sex +1",
                          if_else(analysis_disaggregation_age == 1 & analysis_disaggregation_sex == 1 & (
                            analysis_disaggregation_ethnicity == 1 | analysis_disaggregation_education == 1 |
                            analysis_disaggregation_profession == 1 | analysis_disaggregation_ses == 1), "Age & Sex +1",
                          if_else((analysis_disaggregation_age == 0 | analysis_disaggregation_sex == 0) & (
                            analysis_disaggregation_ethnicity == 1 | analysis_disaggregation_education == 1 |
                            analysis_disaggregation_profession == 1 | analysis_disaggregation_ses == 1), "Other demographics",
                            "No disaggregation", ))))))))

# Count the number of articles for each condition by year of publication
data.analysis <- data.analysis %>%
  count(report, article_year) %>%
  ungroup()

# Compute the proportion of articles per condition by year of publication
data.analysis <- data.analysis %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(report, add=TRUE) %>%
  mutate(per=round(100*n/total,2))

data.analysis$article_year <- as.numeric(data.analysis$article_year)
data.analysis$per <- as.numeric(data.analysis$per)

# Stacked bar chart of the use of demographic variables in the analysis by year of publication
figure8 <- ggplot(data.analysis, aes(x=article_year, y=per)) +
  geom_col(aes(fill=report)) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Variable reported") +
  scale_fill_viridis_d() +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=14), axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))
```

# Results

### Inclusion & Exclusion

  `r round(percent.included)`% of the sampled articles met eligibility criteria during the review.

```{r, fig.cap="Inclusion and exclusion of articles.", warning=FALSE, message=FALSE}
figure1B
```

### Funding

  Funding sources were reported by `r round(percent.funding.name)`% of studies, while funding number was also reported in `r round(percent.funding.num)`% of these cases. Overall, funding by the United States' National Institutes of Health (NIH) represented `r round(percent.funding.NI)`% of the reported funding agencies. `r round(percent.funding.NI.num)`% of the studies funded by the NIH also reported funding number.

### Geographical location

  `r round(100-percent.country.multiple)`% of articles were conducted in a single country. The geographical location of the study was explicitly reported in `r round(percent.country.reported)`% of studies. The country of study was inferred for the rest of the sampled articles. Inference was primarily based on the first author's affiliation. Overall, `r nrow(country.single)` countries were represented. Figure 2B shows the distribution of study location across time with the eight most represented countries.

```{r, fig.cap="Geographical location of the studies. The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
figure2B
```

### Sample size

  Sample size was reported in `r round(percent.ss)`% of studies, while `r round(percent.study.single)`% investigated a single sampled population.

```{r, fig.cap="Sample size of the recruited volunteers. Numbers are computed with a log-10 scale.", warning=FALSE, message=FALSE}
figure3
```

### Age

  `r round(percent.age)`% of articles reported a variable describing age. Figure 4A and 4B show the use of various age variables by year and journal of publication. Overall, the mean age of the study populations was `r round(mean(data.age.mean$age_reported_mean_value))` years old.

```{r, warning=FALSE, message=FALSE}
figure4A
figure4B
```

```{r, fig.cap="Reporting of age variables. Correlation between the reporting of the mean, standrad deviation of the mean, median, minimum, maximum, interquartile range and other age variables by A. year of publication and B. journal of publication. C. Mean age by year of publication.", message=FALSE, warning=FALSE}
figure4C

#grid.arrange(figure4A, figure4C, figure4B, nrow=2)
```

### Sex

  Sex was reported in `r round(percent.sex)`% of the studies. Figure 5 displays the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants. `r round(percent.sex.male)`% of the studies reporting sex only recruited male participants, while `r round(percent.sex.female)`% only employed females. Out of the studies focusing on a single gender, `r round(percent.sex.male.focus)`% of the male studies focused on a sex dependent feature, while `r round(percent.sex.female.focus)`% of the female studies did. `r round(percent.sex.age)`% of studies reported age by sex.

```{r, fig.cap="Sex inclusion. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
figure5A
```

###  Ethnicity, education, profession and socio-economic status

  Other demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status. Figure 6 shows the distribution of this reporting across the years. Figures 7 show the number of categories reported for each variable.

```{r, fig.cap="Reporting of education, ethinicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
#figure6.bars
figure6.points
```

```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
grid.arrange(figure7A, figure7B, figure7C, figure7D, nrow=2)
```

### Study focus

  `r round(percent.focus.sex)`% of articles focused on a sex dependent feature, while `r round(percent.focus.clinical)`% investigated a clinical feature. `r round(percent.focus.twins)`% of studies focused on twins, `r round(percent.focus.pregnant)`% on pregnant women, `r round(percent.focus.shift)`% on shift workers and `r round(percent.focus.students)`% on university students.

### Analysis disaggregation

  A

```{r, fig.cap="Use of study population characteristics as variables in the analysis."}
figure8
```

# Discussion

# Conclusion

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

### 2021-04-15 Sleep Review Paper v1.Rmd
Updates

@@
 # Set working directory
-setwd("~/Sleep Review")
+setwd("~/Projects/sleep_circadian_demographics_data")
 
 # Download scoring sheet 
-data.original <- read_excel("Sleep Review Scoring Sheet Test.xlsx")
+data.original <- read_excel("scoring_rhiannon.xlsx")
 ```

### 2021-04-21 Demographics Review R markdown.Rmd
R markdown update
R markdown was updated with the corrected excel scoring sheet.

---
title             : "Study Sample Characteristics in the Sleep and Chronobiology Research: A Systematic Review"
shorttitle        : "Demographics analysis of sleep and chronobiology research"

author: 
  - name          : "Selma Tir"
    affiliation   : "1,2"
    email         : "selma.tir@ndcn.ox.ac.uk"
    role:
      - wrote the original draft protocol, analyzed data, reviewed and edited the manuscript
  - name          : "Rhiannon White"
    affiliation   : "1,3"
    email         : "r.white.5@warwick.ac.uk"
    role:         
      - reviewed and scored articles, reviewed and edited the manuscript
  - name          : "Manuel Spitschan"
    affiliation   : "1,4,5"
    corresponding : yes    # Define only one corresponding author
    email         : "manuel.spitschan@psy.ox.ac.uk"
    role:
      - wrote the original draft protocol, reviewed and edited the manuscript

affiliation:
  - id            : "1"
    institution   : "Department of Experimental Psychology, University of Oxford, Anna Watts Building, Radcliffe Observatory Quarter, Oxford, OX2 6GG, United Kingdom"
  - id            : "2"
    institution   : "Nuffield Department of Clinical Neurosciences, University of Oxford"
  - id            : "3"
    institution   : "Warwick Medical School, University of Warwick, United Kingdom"
  - id            : "4"
    institution   : "Centre for Chronobiology, Psychiatric Hospital of the University of Basel, Switzerland"
  - id            : "5"
    institution   : "Transfaculty Research Platform Molecular and Cognitive Neurosciences, University of Basel, Switzerland"

authornote: |
  This systematic review is funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
abstract: |
  Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. In this systematic review, we evaluated the study sample characteristics in the published sleep and chronobiology research over the past 40 years.

  Articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals, identified by their five-year Impact Factor, were found through MEDLINE. 6,777 articles were identified and a random sample of 20% was initially included for screening. Inclusion requirements included conducting original research, reporting human data, and recruiting volunteers. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0=not reported), and any reported aggregate summary statistics for these variables were recorded. Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed including any of the demographic variables as covariates.

  ~75% of screened articles met inclusion criteria. While >90% of studies reported age or sex, all other variables were reported in <10% of cases. We found that sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s. Overall, ~75% of studies recruited both male and female participants. Of studies recruiting a single sex, ~50% all-female studies focused on a sex-dependent feature, compared to <5% in all-male studies.

  In this comprehensive review, we found that the majority of studies report at least sex or age, while many other important variables are typically not reported. Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate disaggregated data analyses.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Study Sample Characteristics, Demographics, Sex, Systematic Review"
wordcount         : "X"

bibliography      : ["r-references.bib"]

floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "jou"
output            : papaja::apa6_pdf
---

```{r setup, include=FALSE}
# Version control
## R version 4.0.5 (2021-03-31)
## R Studio version 1.4.1106
## Microsoft Windows version 10.0.19041.329

# Load packages
library(papaja)
library(corrplot)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(ggExtra)
library(ggsci)
library(graphics)
library(gridExtra)
library(knitr)
library(latticeExtra)
library(mosaic)
library(pracma)
library(readr)
library(readxl)
library(reshape2)
library(stats) 
library(stringr)
library(tibble)
library(tidyr)
library(lemon)

r_refs("r-references.bib")
```

```{r Data, include=FALSE}
## Download data sheet from Github repository
## Change path if needed

# Set working directory
setwd("~/Demographics Review")

# Download scoring sheet 
data.original <- read_excel("scoring_rhiannon.xlsx")
```

# Introduction

  Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. In this systematic review, we evaluated the study sample characteristics in the published sleep and chronobiology research over the past 40 years.

# Methods

  Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals  were found through MEDLINE by MS. The eight journals were selected based on their five-year Impact Factor, and included Journal of Pineal Research, Sleep, Journal of Sleep Research, Sleep Medicine, Journal of Clinical Medicine, Journal of Biological Rhythms, Behavioral Sleep Medicine, and Chronobiology International. 6,777 articles were identified and a random sample of 20% was initially selected for screening. Inclusion requirements included conducting original research in the English language, reporting human data, and recruiting volunteers. As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, erratums and corrigendums were excluded. All articles were reviewed for eligibility and coded by ST. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded. Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.

```{r Inclusion & Exclusion, echo=FALSE, message=FALSE, warning=FALSE}
# Filter reviewed articles
data.original <- data.original %>%
  filter(article_fulltext_available_yn == 1)

# Compute total number of reviewed articles
n.original <- as.numeric(nrow(data.original))

# Separate included and excluded articles
data.included <- data.original %>%
  filter(article_included_yn == 1)

data.excluded <- data.original %>%
  filter(article_included_yn == 0)

# Check number of included and excluded articles is equal to total (return TRUE if yes)
check.inclusion <- (as.numeric(nrow(data.included)) + as.numeric(nrow(data.excluded))) == as.numeric(nrow(data.original))

# Compute percentage of included vs excluded articles, round to two decimals
percent.included <- round(as.numeric(nrow(data.included))/as.numeric(nrow(data.original)) * 100, 2)
percent.excluded <- round(as.numeric(nrow(data.excluded))/as.numeric(nrow(data.original)) * 100, 2)

# Stacked bar chart of included articles by journal and year of publication
figure1A <- ggplot(data.included, aes(x=article_year)) +
  geom_bar(stat="count", aes(fill=article_journal)) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Number of included articles") +
  labs(fill="Journal") +
  ggtitle("A") +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_viridis_d() +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12), legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10))

# Compute percentage of articles excluded for other reasons
data.excluded.other <- data.excluded %>%
  filter(!is.na(article_excluded_reason_other))
percent.excluded.other <- round(nrow(data.excluded.other)/nrow(data.excluded) * 100, 2)

# Compute percentage of excluded articles by reason for exclusion
data.eligibility <- data.original
data.eligibility$article_excluded_reason <- tolower(data.eligibility$article_excluded_reason) 
data.eligibility <- data.eligibility %>%
  mutate(eligibility=if_else(!is.na(article_excluded_reason_other), "Not included for other reasons", 
                          if_else(article_included_yn == 1, "Included",
                                  if_else(article_excluded_reason == "not recruiting volunteers", 
                                          "Not recruiting volunteers",
                                          if_else(article_excluded_reason == "not an original research article",
                                                  "Not an original research article",
                                                  if_else(article_excluded_reason == "not reporting human data",
                                                          "Not reporting human data",
                                                          "NA"))))))

data.eligibility <- data.eligibility %>%
  count(eligibility, article_year) %>%
  ungroup()

data.eligibility <- data.eligibility %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(eligibility) %>%
  mutate(per=round(100*n/total,3))

data.eligibility$article_year <- as.numeric(data.eligibility$article_year)
data.eligibility$per <- as.numeric(data.eligibility$per)

# Stacked bar chart of excluded articles by reason for exclusion and year of publication
figure1B <- ggplot(data.eligibility, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(eligibility, levels=c("Not an original research article",
                                                 "Not reporting human data",
                                                 "Not included for other reasons",
                                                 "Not recruiting volunteers",
                                                 "Included")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Eligibility") +
  ggtitle("B") +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_manual(values=c("#29AF7FFF","#482677FF","#D8DE29FF","#2D708EFF")) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12), legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10))
```

```{r Funding, include=FALSE}
# Select articles that report funding
data.funding.name <- data.included %>%
  filter(funding_reported_yn == 1)

# Compute percentage of articles that report funding
percent.funding.name <- round(nrow(data.funding.name)/as.numeric(nrow(data.included)) * 100, 2)

# Select articles that report funding number
data.funding.num <- data.included %>%
  filter(funding_refnum_yn == 1)

# Compute percentage of articles that report funding number, out of those that report funding
percent.funding.num <- round(nrow(data.funding.num)/as.numeric(nrow(data.funding.name)) * 100, 2)
list.funding <- tolower(data.included$funding_names)

# Split the category combinations and convert to lower case letters
list.funding <- as.list(strsplit(list.funding, ","))
list.funding <- unlist(list.funding)

# Count number of occurrences for each category
table.list.funding <- as.data.frame(table(list.funding))
n.funding.nih <- sum(str_count(list.funding, "nih"), str_count(list.funding, "national institute"))

# Compute percentage of articles that report funding from the National Institutes in the US
data.funding.NI <- data.funding.name %>%
  filter(grepl("NI", funding_names) | grepl("National Institute", funding_names))
percent.funding.NI <- round(nrow(data.funding.NI)/as.numeric(nrow(data.included)) * 100, 2)

# Select articles that report funding number and funding from National Institutes in the US
data.funding.NI.num <- data.funding.name %>%
  filter(grepl("NI", funding_names) | grepl("National Institute", funding_names)) %>%
  filter(funding_refnum_yn == 1)
percent.funding.NI.num <- round(nrow(data.funding.NI.num)/as.numeric(nrow(data.funding.NI)) * 100, 2)
```

```{r Study Country, echo=FALSE, message=FALSE, warning=FALSE}
# Set country code to upper case
data.included$study_country_values <- toupper(data.included$study_country_values)

# Separate articles for which country is reported explicitly vs inferred
data.country.reported <- data.included %>%
  filter(study_country_reported_explicitly_yn == 1)

data.country.inferred <- data.included %>%
  filter(study_country_reported_inferred_yn == 1)

# Compute percentage of articles where country is reported explicitly vs inferred
percent.country.reported <- round(nrow(data.country.reported)/nrow(data.included) * 100, 2)
percent.country.inferred <- round(nrow(data.country.inferred)/nrow(data.included) * 100, 2)

# Separate articles that report one or multiple countries of study
# Compute percentage of articles per study country
country.single <- data.included %>%
  filter(!grepl(",", study_country_values)) %>%
  group_by(study_country_values) %>%
  tally() %>%
  mutate(n=round((n/as.numeric(nrow(data.included))*100),1))

country.multiple <- data.included %>%
  filter(grepl(",", study_country_values)) %>%
  group_by(study_country_values) %>%
  tally() %>%
  mutate(n=round((n/as.numeric(nrow(data.included))*100),1))

# Compute percentage of articles where multiple countries of study are reported
percent.country.multiple <- round(nrow(country.multiple)/nrow(data.included) * 100, 2)

# Compute number of articles for the 8 most represented countries by year of publication
# Check in dataset "country.single" which countries are the 8 most represented

# Select articles where a single country is reported or inferred
data.country <- data.included %>%
  filter(!grepl(",", study_country_values)) %>%
  count(study_country_values, article_year) %>%
  ungroup()

# Change all countries that are not the 8 most represented countries to 'Other'
data.country$study_country_values[data.country$study_country_values != "AUS" &
                                    data.country$study_country_values != "CAN" &
                                    data.country$study_country_values != "DEU" &
                                    data.country$study_country_values != "FRA" &
                                    data.country$study_country_values != "GBR" &
                                    data.country$study_country_values != "USA" &
                                    data.country$study_country_values != "ITA" &
                                    data.country$study_country_values != "JPN"] <- "Other"

# Sum rows with 'Other' by year
data.country <- aggregate(cbind(n) ~ study_country_values + article_year, data=data.country, FUN=sum)

# Compute proportion of articles per country by year of publication
data.country <- data.country %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(study_country_values) %>%
  mutate(per=round(100*n/total,3))
data.country$article_year <- as.numeric(data.country$article_year)
data.country$per <- as.numeric(data.country$per)

# Stacked bar chart of 8 most represented study countries by year of publication
figure2B <- ggplot(data.country, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(study_country_values, 
                           levels=c("Other","AUS","CAN","DEU","FRA",
                                    "GBR","ITA","JPN","USA")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Country") +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_viridis_d(direction=-1) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12), legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10))
```

```{r Number of sub-studies, echo=FALSE, message=FALSE, warning=FALSE}
# Compute percentage of articles per number of sub-studies
num.study <- data.included %>%
  group_by(num_substudies) %>%
  tally() %>%
  mutate(num_substudies=as.factor(num_substudies),
         n=round((n/as.numeric(nrow(data.included))*100),1))
percent.study.single <- num.study$n[1]
percent.study.multiple <- 1 - num.study$n[1]

### The rest  of the analysis will be conducted on included single studies
# Filter single studies
data <- data.included %>%
  filter(num_substudies == 1)
```

  Data were coded in an Excel Spreadsheet and analyzed in R Studio (version 4.0.5). Reporting of funding, geographical location, and number of sub-studies for each article were investigated for the sample of articles that passed all eligibility criteria. The rest of the analysis was conducted on the subset of articles that reported a single study (`r round(percent.study.single)`% of articles), i.e. a single sample population.

```{r Sample size, echo=FALSE, message=FALSE, warning=FALSE}
# Compute percentage of studies that report sample size
data.ss <- data %>%
  filter(sample_size_bycategory_reported_yn == 1)
percent.ss <- round(nrow(data.ss)/nrow(data) * 100, 2)

# Violin plot of sample size by year of publication with log10 scaling
data.ss$article_year <- as.factor(data.ss$article_year)
data.ss$sample_size_total_reported_values <- as.numeric(data.ss$sample_size_total_reported_values)

figure3 <- ggplot(data.ss, aes(x=article_year, y=sample_size_total_reported_values)) +
  geom_violin(fill="#440154FF") +
  theme_apa() +
  scale_y_continuous(trans='log10', breaks=c(10,1000,100000), 
                     labels=c("10","1,000","100,000")) +
  scale_x_discrete(breaks=c("1979","1989","1999","2009","2019")) +
  xlab("Year of publication") +
  ylab("Sample size, log-10 scaling") +
  annotation_logticks(sides="l") +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
```

```{r Age, echo=FALSE, message=FALSE, warning=FALSE}
# Select articles reporting age
data.age <- data %>%
  filter(age_reported_mean_yn == 1 | age_reported_median_yn == 1 | age_reported_min_yn | age_reported_max_yn | age_reported_other_yn == 1 | age_reported_by_sex_yn == 1)

# Compute percentage of articles reporting age
percent.age <- round(nrow(data.age)/nrow(data) * 100, 2)

## Correlate reporting of age variables by year
# select year and age columns
data.age.year <- data.age %>%
  select(article_year, age_reported_mean_yn, age_reported_sd_yn, age_reported_median_yn, 
         age_reported_min_yn, age_reported_max_yn, age_reported_iqr_yn, age_reported_other_yn)

# Replace 0 by -1
data.age.year$age_reported_mean_yn[data.age.year$age_reported_mean_yn == 0] <- -1
data.age.year$age_reported_sd_yn[data.age.year$age_reported_sd_yn == 0] <- -1
data.age.year$age_reported_median_yn[data.age.year$age_reported_median_yn == 0] <- -1
data.age.year$age_reported_min_yn[data.age.year$age_reported_min_yn == 0] <- -1
data.age.year$age_reported_max_yn[data.age.year$age_reported_max_yn == 0] <- -1
data.age.year$age_reported_iqr_yn[data.age.year$age_reported_iqr_yn == 0] <- -1
data.age.year$age_reported_other_yn[data.age.year$age_reported_other_yn == 0] <- -1

# Replace NAs
data.age.year[is.na(data.age.year)] <- -1

# Group by year and compute average for each sex variable
age.corr.year <- data.age.year %>%
  group_by(article_year) %>%
  summarise(mean=mean(age_reported_mean_yn), 
            sd=mean(age_reported_sd_yn), median=mean(age_reported_median_yn),
            min=mean(age_reported_min_yn), max=mean(age_reported_max_yn),
            iqr=mean(age_reported_iqr_yn),
            other=mean(age_reported_other_yn))  %>%
  ungroup()

age.corr.year <- age.corr.year %>%
  group_by(article_year) %>%
  mutate(total=((mean+sd+median+min+max+iqr+other)/7))

# Reshape data
age.corr.year <- melt(age.corr.year,id.vars='article_year',
                      measure.vars=c('mean','sd','median','min',
                                     'max','iqr','other','total'))

# Heat map of reporting of age variables by year of publication
figure4A <- ggplot(data=age.corr.year, aes(x=variable, y=article_year, fill=value)) +
  geom_tile() +
  theme_apa() +
  ylab(label="Year") +
  xlab("Age variable") +
  scale_fill_viridis_c(end=0.6, direction=-1, labels=c("0.0","0.25","0.5","0.75","1")) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8), 
        legend.position="none", legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_blank())

## Correlate reporting of age variables by journal
# Select journal and age columns
data.age.journal <- data.age %>%
  select(article_journal, age_reported_mean_yn, age_reported_sd_yn, age_reported_median_yn, 
         age_reported_min_yn, age_reported_max_yn, age_reported_iqr_yn, age_reported_other_yn)

# Replace 0 by -1
data.age.journal$age_reported_mean_yn[data.age.journal$age_reported_mean_yn == 0] <- -1
data.age.journal$age_reported_sd_yn[data.age.journal$age_reported_sd_yn == 0] <- -1
data.age.journal$age_reported_median_yn[data.age.journal$age_reported_median_yn == 0] <- -1
data.age.journal$age_reported_min_yn[data.age.journal$age_reported_min_yn == 0] <- -1
data.age.journal$age_reported_max_yn[data.age.journal$age_reported_max_yn == 0] <- -1
data.age.journal$age_reported_iqr_yn[data.age.journal$age_reported_iqr_yn == 0] <- -1
data.age.journal$age_reported_other_yn[data.age.journal$age_reported_other_yn == 0] <- -1

# Replace NAs
data.age.journal[is.na(data.age.journal)] <- -1

# Group by journal and compute average for each sex variable
age.corr.journal <- data.age.journal %>%
  group_by(article_journal) %>%
  summarise(mean=mean(age_reported_mean_yn), sd=mean(age_reported_sd_yn),
            median=mean(age_reported_median_yn),min=mean(age_reported_min_yn),
            max=mean(age_reported_max_yn), iqr=mean(age_reported_iqr_yn),
            other=mean(age_reported_other_yn)) %>%
  ungroup()

age.corr.journal <- age.corr.journal %>%
  group_by(article_journal) %>%
  mutate(total=((mean+sd+median+min+max+iqr+other)/7))

# Reshape data
age.corr.journal <- melt(age.corr.journal,id.vars='article_journal',
                         measure.vars=c('mean','sd','median','min',
                                        'max','iqr','other','total'))

# Heat map of reporting of age variables by journal of publication
figure4B <- ggplot(data=age.corr.journal, 
                   aes(x=variable, y=article_journal, fill=value)) +
  geom_tile() +
  theme_apa() +
  ylab(label="Journal") +
  xlab("Age variable") +
  scale_fill_viridis_c(end=0.6, direction=-1, labels=c("0.0","0.25","0.5","0.75","1")) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.position="right", legend.margin=margin(0,0,0,0),
        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_blank())

## Regression for age by year of publication
# filter articles that report mean age in years
data.age.mean <- data.age %>%
  filter(age_reported_mean_yn == 1) %>%
  filter(!is.na(as.numeric(age_reported_mean_value))) %>%
  select(article_pmid, article_year, age_reported_mean_value)

# filter articles that report mean age in weeks or months
data.age.mean.not.years <- data.age %>%
  filter(age_reported_mean_yn == 1) %>%
  filter(is.na(as.numeric(age_reported_mean_value))) %>%
  select(article_pmid, article_year, age_reported_mean_value)

# Change age from weeks to years, with 1 month = 4.3 weeks, and from months to years
# age not in years is entered with "months" or "weeks"
data.age.mean.not.years <- data.age.mean.not.years %>%
  separate(age_reported_mean_value, 
           into=c("age_reported_mean_value", "text"),
           sep=" ")
data.age.mean.not.years$age_reported_mean_value <- as.numeric(data.age.mean.not.years$age_reported_mean_value)

data.age.mean.not.years <- data.age.mean.not.years %>%
  mutate(age_reported_mean_value=ifelse(text=="weeks", 
                                        age_reported_mean_value / 4.3 /12,
                                        age_reported_mean_value / 12)) %>%
  select(article_pmid, article_year, age_reported_mean_value)# else case is in months

# Merge back mean age datasets
data.age.mean <- rbind(data.age.mean, data.age.mean.not.years)

data.age.mean$age_reported_mean_value <- as.numeric(data.age.mean$age_reported_mean_value)

figure4C <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
  geom_smooth(method="lm", color="#440154FF") +
  theme_apa() +
  ylab("Mean age") +
  xlab("Year") +
  scale_x_continuous(limits=c(1979,2019),breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(limits=c(20,60), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
```

```{r Sex, echo=FALSE, message=FALSE, warning=FALSE}
# Select sex variables       
data.sex <- data %>%
  select(article_year, sex_reported_yn,
         sex_reported_num_male, sex_reported_percent_male,
         sex_reported_num_female,sex_reported_percent_female)

# Compute whether males or females are reported at all
data.sex$male <- ifelse(is.na(data.sex$sex_reported_num_male) & is.na(data.sex$sex_reported_percent_male),
                             "No",
                             "Yes")
data.sex$male <- as.factor(data.sex$male)

data.sex$female <- ifelse(is.na(data.sex$sex_reported_num_female) & is.na(data.sex$sex_reported_percent_female),
                             "No",
                             "Yes")
data.sex$female <- as.factor(data.sex$female)

# Compute whether articles report males, females, both or unspecified
data.sex <- data.sex %>%
  mutate(report=if_else(male == "Yes" & female == "Yes", "Both Sexes", 
                          if_else(male == "Yes" & female == "No", "Male",
                                  if_else(male == "No" & female == "Yes", "Female", "Unspecified"))))

# Count number of articles for each sex condition by year of publication
data.sex <- data.sex %>%
  count(report, article_year) %>%
  ungroup()

# Compute proportion of articles per sex category by year of publication
data.sex <- data.sex %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(report, add=TRUE) %>%
  mutate(per=round(100*n/total,3))
data.sex$article_year <- as.numeric(data.sex$article_year)
data.sex$per <- as.numeric(data.sex$per)

# Stacked bar chart of sex reporting by year of publication
figure5A <- ggplot(data.sex, aes(x=article_year, y=per)) +
  geom_col(aes(fill=factor(report, levels=c("Unspecified",
                                            "Female",
                                            "Both Sexes",
                                            "Male")))) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Sex") +
  scale_fill_manual(values=c("#482677FF","#D8DE29FF","#29AF7FFF","#2D708EFF")) +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

# Select articles for which sex is reported
data.sex.reported <- data %>%
  filter(sex_reported_yn == 1) %>%
  mutate(sample_size_total_reported_values=as.numeric(sample_size_total_reported_values),
         sex_reported_num_male=as.numeric(sex_reported_num_male),
         sex_reported_percent_male=as.numeric(sex_reported_percent_male),
         sex_reported_num_female=as.numeric(sex_reported_num_female),
         sex_reported_percent_female=as.numeric(sex_reported_percent_female))

# Select articles for which intersex is reported
data.intersex <- data %>%
  filter(sex_reported_intersex_yn == 1)

# compute percentage of articles where sex and intersex is reported
percent.sex <- round(nrow(data.sex.reported)/nrow(data) * 100, 2)
percent.intersex <- round(nrow(data.intersex)/nrow(data) * 100, 2)

# Compute whether males or females are recruited in the articles that report  sex
data.sex.reported$male <- ifelse(is.na(data.sex.reported$sex_reported_num_male) &
                                   is.na(data.sex.reported$sex_reported_percent_male),
                             "No",
                             "Yes")
data.sex.reported$male <- as.factor(data.sex.reported$male)

data.sex.reported$female <- ifelse(is.na(data.sex.reported$sex_reported_num_female) &
                                     is.na(data.sex.reported$sex_reported_percent_female),
                             "No",
                             "Yes")
data.sex.reported$female <- as.factor(data.sex.reported$female)

# Compute whether articles report males, females or both
data.sex.reported <- data.sex.reported %>%
  mutate(report=if_else(male == "Yes" & female == "Yes", "Both Sexes", 
                          if_else(male == "Yes" & female == "No", "Male",
                                  if_else(male == "No" & female == "Yes", "Female", "NA"))))

# Compute percentage of articles that only recruit males or females
data.sex.male <- data.sex.reported %>%
  filter(report=="Male")
percent.sex.male <- round(nrow(data.sex.male)/nrow(data.sex.reported) * 100, 2)

data.sex.female <- data.sex.reported %>%
  filter(report=="Female")
percent.sex.female <- round(nrow(data.sex.female)/nrow(data.sex.reported) * 100, 2)

# Compute percentage of articles reporting on a sex-dependent feature
data.sex.male.focus <- data.sex.male %>%
  filter(sample_focus_sex_dependent_yn == 1)
percent.sex.male.focus <- round(nrow(data.sex.male.focus)/nrow(data.sex.male) * 100, 2)

data.sex.female.focus <- data.sex.female %>%
  filter(sample_focus_sex_dependent_yn == 1)
percent.sex.female.focus <- round(nrow(data.sex.female.focus)/nrow(data.sex.female) * 100, 2)

# Compute percentage of studies reporting age by sex
data.sex.age <- data %>%
  filter(age_reported_by_sex_yn == 1)
percent.sex.age <- round(nrow(data.sex.age)/nrow(data) * 100, 2)
```

```{r Other demographics, echo=FALSE, message=FALSE, warning=FALSE}
# Select single studies and other demographic variables
data.demo <- data %>%
  select(article_year, ethnicity_reported_yn, 
         education_reported_mean_yn, education_reported_yn,
         profession_reported_yn, ses_reported_yn)

# Compute the reporting of the different demographic variables
data.demo <- data.demo %>%
  mutate(education=if_else(education_reported_mean_yn == 0 & education_reported_yn == 0, "No", "Yes"),
         ethnicity=if_else(ethnicity_reported_yn == 1, "Yes", "No"),
         profession=if_else(profession_reported_yn == 1, "Yes", "No"),
         ses=if_else(ses_reported_yn == 1, "Yes", "No"))

# Count number of articles for each demographic variable by year of publication
data.demo.education <- data.demo %>%
  count(education, article_year) %>%
  ungroup()
data.demo.ethnicity <- data.demo %>%
  count(ethnicity, article_year) %>%
  ungroup()
data.demo.profession <- data.demo %>%
  count(profession, article_year) %>%
  ungroup()
data.demo.ses <- data.demo %>%
  count(ses, article_year) %>%
  ungroup()

# Compute proportion of articles by year of publication
data.demo.education <- data.demo.education %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(education, add=TRUE) %>%
  mutate(per=round(100*n/total,3),
         variable="edu") %>%
  select(article_year, variable, education, per)
colnames(data.demo.education)[3] <- "report"

data.demo.ethnicity <- data.demo.ethnicity %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(ethnicity, add=TRUE) %>%
  mutate(per=round(100*n/total,3),
         variable="ethni") %>%
  select(article_year, variable, ethnicity, per)
colnames(data.demo.ethnicity)[3] <- "report"

data.demo.profession <- data.demo.profession %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(profession, add=TRUE) %>%
  mutate(per=round(100*n/total,3),
         variable="prof") %>%
  select(article_year, variable, profession, per)
colnames(data.demo.profession)[3] <- "report"

data.demo.ses <- data.demo.ses %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(ses, add=TRUE) %>%
  mutate(per=round(100*n/total,3),
         variable="ses") %>%
  select(article_year, variable, ses, per)
colnames(data.demo.ses)[3] <- "report"

# Merge back the datasets together
merge1 <- rbind(data.demo.education, data.demo.ethnicity)
merge2 <- rbind(merge1, data.demo.profession)
data.demo <- rbind(merge2, data.demo.ses)
data.demo$article_year <- as.numeric(data.demo$article_year)
data.demo$per <- as.numeric(data.demo$per)

# Stacked bar chart  of reporting of other demographic variables by year of publication
figure6.bars <- ggplot(data.demo, aes(x=variable, y=per, fill=report)) +
  geom_bar(stat="identity") +
  facet_grid(~article_year) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Variable reported") +
  scale_fill_manual(values=c("#482677FF","#D8DE29FF")) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"),
        axis.text.x=element_text(size=5, angle=45, hjust=1),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10),
        panel.spacing=unit(0.1, "lines"))

data.demo.reported <- data.demo %>%
  filter(report == "No") %>%
  mutate(per = (100 - per))

figure6.points <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
  geom_point(size=2.5) +
  geom_line(linetype="dashed") +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  scale_color_viridis_d(name="Variable", 
                        labels=c("Education","Ethnicity","Profession","Socio-economic status")) +
  scale_x_continuous(limits=c(1979,2019), breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(limits=c(0,100)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))

# Compute percentage of articles that report education
data.education <- data %>%
  filter(education_reported_mean_yn == 1 | education_reported_yn == 1)
percent.education <- round(nrow(data.education)/nrow(data) * 100, 2)

# Compute distribution of the number of categories reported for education
data.education$education_reported_numcategory <- as.numeric(data.education$education_reported_numcategory)
figure7A <- ggplot(data=data.education, aes(x=education_reported_numcategory)) +
  geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
  geom_vline(aes(xintercept=mean(education_reported_numcategory, na.rm=TRUE)), 
             color="#D8DE29FF", size=1) +
  theme_apa() +
  ylab("Density") +
  xlab("Education") +
  scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))

# Compute percentage of articles that report ethnicity
data.ethnicity <- data %>%
  filter(ethnicity_reported_yn == 1)
percent.ethnicity <- round(nrow(data.ethnicity)/nrow(data) * 100, 2)

# Compute distribution of the number of categories reported for ethnicity
data.ethnicity$ethnicity_reported_numcategory <- as.numeric(data.ethnicity$ethnicity_reported_numcategory)
figure7B <- ggplot(data=data.ethnicity, aes(x=ethnicity_reported_numcategory)) +
  geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
  geom_vline(aes(xintercept=mean(ethnicity_reported_numcategory, na.rm=TRUE)), 
             color="#D8DE29FF", size=1) +
  theme_apa() +
  ylab("Density") +
  xlab("Ethnicity") +
  scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))

# Compute percentage of articles that report profession
data.profession <- data %>%
  filter(profession_reported_yn == 1)
percent.profession <- round(nrow(data.profession)/nrow(data) * 100, 2)

# Compute distribution of the number of categories reported for profession
data.profession$profession_reported_numcategory <- as.numeric(data.profession$profession_reported_numcategory)
figure7C <- ggplot(data=data.profession, aes(x=profession_reported_numcategory)) +
  geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
  geom_vline(aes(xintercept=mean(profession_reported_numcategory, na.rm=TRUE)), 
             color="#D8DE29FF", size=1) +
  theme_apa() +
  ylab("Density") +
  xlab("Profession") +
  scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))

# Compute percentage of articles that report ses
data.ses <- data %>%
  filter(ses_reported_yn ==1)
percent.ses <- round(nrow(data.ses)/nrow(data) * 100, 2)

# Compute distribution of the number of categories reported for ses
data.ses$ses_reported_numcategory <- as.numeric(data.ses$ses_reported_numcategory)
figure7D <- ggplot(data=data.ses, aes(x=ses_reported_numcategory)) +
  geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
  geom_vline(aes(xintercept=mean(ses_reported_numcategory, na.rm=TRUE)), 
             color="#D8DE29FF", size=1) +
  theme_apa() +
  ylab("Density") +
  xlab("SES") +
  scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"),
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8)) 
```

```{r Focus, echo=FALSE, message=FALSE, warning=FALSE}
## Sex feature
# Compute percentage of articles focusing on a sex dependent feature
data.focus.sex <- data %>%
  filter(sample_focus_sex_dependent_yn == 1)
percent.focus.sex <- round(nrow(data.focus.sex)/nrow(data) * 100, 2)

# Compute number of articles per sex dependent feature 
sex.dependent <- data.focus.sex %>%
  group_by(sample_focus_sex_dependent) %>%
  tally() %>%
  mutate(n=round((n/nrow(data.focus.sex)*100),1))

# Pie chart of the reported sex dependent features 
pc.sex.dependent <- ggplot(data=sex.dependent, aes(x="", y=n, fill=sample_focus_sex_dependent)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  geom_text(aes(label=paste0(n, "%")), position=position_stack(vjust=0.5)) +
  ggtitle("Articles focused on a sex dependent feature") +
  labs(fill="Sex dependent feature") +
  scale_fill_brewer(palette="RdBu") +
  theme_apa()

## Clinical feature
# Compute percentage of articles focusing on a clinical feature
data.focus.clinical <- data %>%
  filter(sample_focus_clinical_yn == 1)
percent.focus.clinical <- round(nrow(data.focus.clinical)/nrow(data) * 100, 2)

# Compute number of articles per clinical feature 
clinical <- data.focus.clinical %>%
  group_by(sample_focus_clinical) %>%
  tally() %>%
  mutate(n=round((n/nrow(data.focus.clinical)*100),1))

# Pie chart of the reported clinical features 
pc.clinical <- ggplot(data=clinical, aes(x="", y=n, fill=sample_focus_clinical)) +
  geom_bar(stat="identity", width=1, color="white") +
  coord_polar("y", start=0) +
  geom_text(aes(label=paste0(n, "%")), position=position_stack(vjust=0.5)) +
  ggtitle("Articles focused on a clinical feature") +
  labs(fill="Clinical feature") +
  scale_fill_brewer(palette="RdBu") +
  theme_void()

## Twins
# Compute percentage of articles focusing on twins
data.focus.twins <- data %>%
  filter(sample_focus_twins == 1)
percent.focus.twins <- round(nrow(data.focus.twins)/nrow(data) * 100, 2)

## University students
# Compute percentage of articles focusing on university students
data.focus.students <- data %>%
  filter(sample_focus_university_students == 1)
percent.focus.students <- round(nrow(data.focus.students)/nrow(data) * 100, 2)

## Shift workers
# Compute percentage of articles focusing on shift workers
data.focus.shift <- data %>%
  filter(sample_focus_shift_workers == 1)
percent.focus.shift <- round(nrow(data.focus.shift)/nrow(data) * 100, 2)

## Pregnant women
# Compute percentage of articles focusing on pregnant women
data.focus.pregnant <- data %>%
  filter(sample_focus_pregnant_women == 1)
percent.focus.pregnant <- round(nrow(data.focus.pregnant)/nrow(data) * 100, 2)
```

```{r Disaggregation, echo=FALSE, warning=FALSE, message=FALSE}
# Select articles that report no difference in results based on age vs use age as a covariate
data.diff.age <- data %>%
  filter(analysis_no_difference_age == 1)
data.dis.age <- data %>%
  filter(analysis_disaggregation_age == 1)

# Select articles that report no difference in results based on sex vs use sex as a covariate
data.diff.sex <- data %>%
  filter(analysis_no_difference_sex == 1)
data.dis.sex <- data %>%
  filter(analysis_disaggregation_sex == 1)

# Select articles that report no difference in results based on ethnicity vs use ethnicity as a covariate
data.diff.ethnicity <- data %>%
  filter(analysis_no_difference_ethnicity == 1)
data.dis.ethnicity <- data %>%
  filter(analysis_disaggregation_ethnicity == 1)

# Select articles that report no difference in results based on education vs use education as a covariate
data.diff.education <- data %>%
  filter(analysis_no_difference_education == 1)
data.dis.education <- data %>%
  filter(analysis_disaggregation_education == 1)

# Select articles that report no difference in results based on profession vs use profession as a covariate
data.diff.profession <- data %>%
  filter(analysis_no_difference_profession == 1)
data.dis.profession <- data %>%
  filter(analysis_disaggregation_profession == 1)

# Select articles that report no difference in results based on ses vs use ses as a covariate
data.diff.ses <- data %>%
  filter(analysis_no_difference_ses == 1)
data.dis.ses <- data %>%
  filter(analysis_disaggregation_ses == 1)

# Select single studies and disaggregation variables
data.analysis <- data %>%
  select(article_year, analysis_disaggregation_age, analysis_disaggregation_sex, analysis_disaggregation_ethnicity, 
         analysis_disaggregation_education, analysis_disaggregation_profession, analysis_disaggregation_ses)

# Compute whether articles use demographic variables in the analysis
data.analysis <- data.analysis %>%
  mutate(report=if_else(analysis_disaggregation_age == 1 & analysis_disaggregation_sex == 0 & 
                            analysis_disaggregation_ethnicity == 0 & analysis_disaggregation_education == 0 &
                            analysis_disaggregation_profession == 0 & analysis_disaggregation_ses == 0 , "Age", 
                          if_else(analysis_disaggregation_age == 0 & analysis_disaggregation_sex == 1 & 
                            analysis_disaggregation_ethnicity == 0 & analysis_disaggregation_education == 0 &
                            analysis_disaggregation_profession == 0 & analysis_disaggregation_ses == 0 , "Sex",
                          if_else(analysis_disaggregation_age == 1 & analysis_disaggregation_sex == 1 & 
                            analysis_disaggregation_ethnicity == 0 & analysis_disaggregation_education == 0 &
                            analysis_disaggregation_profession == 0 & analysis_disaggregation_ses == 0 , "Age & Sex",
                          if_else(analysis_disaggregation_age == 1 & analysis_disaggregation_sex == 0 & (
                            analysis_disaggregation_ethnicity == 1 | analysis_disaggregation_education == 1 |
                            analysis_disaggregation_profession == 1 | analysis_disaggregation_ses == 1), "Age +1",
                          if_else(analysis_disaggregation_age == 0 & analysis_disaggregation_sex == 1 & (
                            analysis_disaggregation_ethnicity == 1 | analysis_disaggregation_education == 1 |
                            analysis_disaggregation_profession == 1 | analysis_disaggregation_ses == 1), "Sex +1",
                          if_else(analysis_disaggregation_age == 1 & analysis_disaggregation_sex == 1 & (
                            analysis_disaggregation_ethnicity == 1 | analysis_disaggregation_education == 1 |
                            analysis_disaggregation_profession == 1 | analysis_disaggregation_ses == 1), "Age & Sex +1",
                          if_else((analysis_disaggregation_age == 0 | analysis_disaggregation_sex == 0) & (
                            analysis_disaggregation_ethnicity == 1 | analysis_disaggregation_education == 1 |
                            analysis_disaggregation_profession == 1 | analysis_disaggregation_ses == 1), "Other demographics",
                            "No disaggregation", ))))))))

# Count the number of articles for each condition by year of publication
data.analysis <- data.analysis %>%
  count(report, article_year) %>%
  ungroup()

# Compute the proportion of articles per condition by year of publication
data.analysis <- data.analysis %>%
  group_by(article_year) %>%
  mutate(total=sum(n)) %>%
  group_by(report, add=TRUE) %>%
  mutate(per=round(100*n/total,3))
data.analysis$article_year <- as.numeric(data.analysis$article_year)
data.analysis$per <- as.numeric(data.analysis$per)

# Stacked bar chart of the use of demographic variables in the analysis by year of publication
figure8 <- ggplot(data.analysis, aes(x=article_year, y=per)) +
  geom_col(aes(fill=report)) +
  theme_apa() +
  xlab("Year of publication") +
  ylab("Percentage of articles") +
  labs(fill="Variable reported") +
  scale_fill_viridis_d() +
  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
  scale_y_continuous(expand=c(0,0)) +
  theme(axis.title=element_text(face="bold", size=12), 
        axis.text=element_text(colour="black", face="bold"), 
        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
        legend.title=element_text(face="bold", size=12),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))
```

# Results

### Inclusion & Exclusion

  `r round(percent.included)`% of the sampled articles met eligibility criteria during the review.

```{r}
figure1A
```

```{r, fig.cap="Inclusion and exclusion of articles. A. Number of articles included by journal and year of publication. B. Proportion of articles included and excluded by reason for exclusion and year of publication.", warning=FALSE, message=FALSE}
figure1B
```

### Funding

  Funding sources were reported by `r round(percent.funding.name)`% of studies, while funding number was also reported in `r round(percent.funding.num)`% of these cases. Overall, funding by the United States' National Institutes of Health (NIH) represented `r round(percent.funding.NI)`% of the reported funding agencies. `r round(percent.funding.NI.num)`% of the studies funded by the NIH also reported funding number.

### Geographical location

  `r round(100-percent.country.multiple)`% of articles were conducted in a single country. The geographical location of the study was explicitly reported in `r round(percent.country.reported)`% of studies. The country of study was inferred for the rest of the sampled articles. Inference was primarily based on the first author's affiliation. Overall, `r nrow(country.single)` countries were represented. `r nrow(country.multiple)`% of articles reported multiple countries of study. Figure 2B shows the distribution of study location across time with the eight most represented countries.

```{r, fig.cap="Geographical location of the studies. The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
figure2B
```

### Sample size

  Sample size was reported in `r round(percent.ss)`% of studies, while `r round(percent.study.single)`% investigated a single sampled population.

```{r, fig.cap="Sample size of the recruited volunteers. Numbers are computed with a log-10 scale.", warning=FALSE, message=FALSE}
figure3
```

### Age

  `r round(percent.age)`% of articles reported a variable describing age. Figure 4A and 4B show the use of various age variables by year and journal of publication. Overall, the mean age of the study populations was `r round(mean(data.age.mean$age_reported_mean_value))` years old.

```{r, warning=FALSE, message=FALSE}
figure4A
figure4B
```

```{r, fig.cap="Reporting of age variables. Correlation between the reporting of the mean, standrad deviation of the mean, median, minimum, maximum, interquartile range and other age variables by A. year of publication and B. journal of publication. C. Mean age by year of publication.", message=FALSE, warning=FALSE}
figure4C
#grid.arrange(figure4A, figure4C, figure4B, nrow=2)
```

### Sex

  Sex was reported in `r round(percent.sex)`% of the studies. Figure 5 displays the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants. `r round(percent.sex.male)`% of the studies reporting sex only recruited male participants, while `r round(percent.sex.female)`% only employed females. Out of the studies focusing on a single gender, `r round(percent.sex.male.focus)`% of the male studies focused on a sex dependent feature, while `r round(percent.sex.female.focus)`% of the female studies did. `r round(percent.sex.age)`% of studies reported age by sex.

```{r, fig.cap="Sex inclusion. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
figure5A
```

###  Ethnicity, education, profession and socio-economic status

  Other demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status. Figure 6 shows the distribution of this reporting across the years. Figures 7 show the number of categories reported for each variable.

```{r, fig.cap="Reporting of education, ethinicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
#figure6.bars
figure6.points
```

```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
grid.arrange(figure7A, figure7B, figure7C, figure7D, nrow=2)
```

### Study focus

  `r round(percent.focus.sex)`% of articles focused on a sex dependent feature, while `r round(percent.focus.clinical)`% investigated a clinical feature. `r round(percent.focus.twins)`% of studies focused on twins, `r round(percent.focus.pregnant)`% on pregnant women, `r round(percent.focus.shift)`% on shift workers and `r round(percent.focus.students)`% on university students.

### Analysis disaggregation

  A

```{r, fig.cap="Use of study population characteristics as variables in the analysis."}
figure8
```

# Discussion

# Conclusion

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

### 2021-04-30 Demographics Review R markdown.Rmd
New version of R markdown
Added new reference file and modified figures.

@@
 
-bibliography      : ["r-references.bib"]
+bibliography      : Demographics Review References.bib
 
@@
 
-r_refs("r-references.bib")
+r_refs("Demographics Review References.bib")
 ```
@@
 In this systematic review, we evaluated the study sample characteristics in the published sleep and chronobiology research over the past 40 years.
+
+  A recent study reviewed the reporting and analysis of sex in the biological sciences research [@RN1].
+The authors found that while sex inclusion has significantly increased over the past 10 years, sex-based analysis has not improved, despite recent policies.
 
@@
   xlab("Year of publication") +
-  ylab("Percentage of articles") +
+  ylab("Proportion of articles") +
   labs(fill="Eligibility") +
@@
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(expand=c(0,0)) +
-  scale_fill_manual(values=c("#29AF7FFF","#482677FF","#D8DE29FF","#2D708EFF")) +
+  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
+  scale_fill_manual(values=c("#29AF7FFF","#2D708EFF","#D8DE29FF","#482677FF")) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
   xlab("Year of publication") +
-  ylab("Percentage of articles") +
+  ylab("Proportion of articles") +
   labs(fill="Country") +
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(expand=c(0,0)) +
+  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
   scale_fill_viridis_d(direction=-1) +
@@
 
-```{r Age, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Age, message=FALSE, warning=FALSE, include=FALSE}
 # Select articles reporting age
@@
 
+# Check regression statistics
+summary(lm(age_reported_mean_value ~ article_year, data=data.age.mean))
+
+# Violin plot with regression line and equation for mean age by year of publication
 figure4C <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
   geom_smooth(method="lm", color="#440154FF") +
-  theme_apa() +
-  ylab("Mean age") +
+  geom_point(size=1, color="#33638DFF") +
+  geom_text(x=1988, y=10, size=3.3,
+            label="Mean age = 281.5 - 0.1 * Year of publication") + # edit manually
+  geom_text(x=1988, y=7, size=3.1, label="R-Sq = 0.3%") + # edit manually
+  theme_apa() +
+  ylab("Mean age (years)") +
   xlab("Year") +
   scale_x_continuous(limits=c(1979,2019),breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(limits=c(20,60), expand=c(0,0)) +
+  scale_y_continuous(limits=c(0,60), expand=c(0,0)) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
   xlab("Year of publication") +
-  ylab("Percentage of articles") +
+  ylab("Proportion of articles") +
   labs(fill="Sex") +
@@
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(expand=c(0,0)) +
+  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
   xlab("Year of publication") +
-  ylab("Percentage of articles") +
+  ylab("Proportion of articles") +
   labs(fill="Variable reported") +
@@
   xlab("Year of publication") +
-  ylab("Percentage of articles") +
+  ylab("Proportion of articles") +
   scale_color_viridis_d(name="Variable", 
@@
   scale_x_continuous(limits=c(1979,2019), breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(limits=c(0,100)) +
+  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), limits=c(0,100)) +
   theme(axis.title=element_text(face="bold", size=12), 

### 2021-04-30 Demographics Review R markdown.Rmd
Updates

@@
 # Set working directory
-setwd("~/Demographics Review")
+#setwd("~/Demographics Review")
+setwd("~/Projects/sleep_circadian_demographics_data")
 
 # Download scoring sheet 
-data.original <- read_excel("scoring_rhiannon.xlsx")
+data.original <- read_excel("scoring.xlsx")
 ```
@@
 
-  A recent study reviewed the reporting and analysis of sex in the biological sciences research [@RN1].
+  A recent study reviewed the reporting and analysis of sex in the biological sciences research [@woitowich_a_2020, @mallampalli_exploring_2014].
 The authors found that while sex inclusion has significantly increased over the past 10 years, sex-based analysis has not improved, despite recent policies.

### 2021-04-30 Demographics Review R markdown.Rmd
Updates
Updates R markdown file with final scoring sheet. Edits to mean age plot and addition of funding plot.

--- 2021-04-30 Demographics Review R markdown.Rmd
@@
 # Download scoring sheet 
-data.original <- read_excel("scoring_rhiannon.xlsx")
+data.original <- read_excel("scoring.xlsx")
 ```
@@
 percent.funding.num <- round(nrow(data.funding.num)/as.numeric(nrow(data.funding.name)) * 100, 2)
-list.funding <- tolower(data.included$funding_names)
-
-# Split the category combinations and convert to lower case letters
+
+# Compute whether articles report funding and funding number
+data.funding <- data.included %>%
+  mutate(funding=if_else(!is.na(funding_names) & funding_refnum_yn==0, 
+                         "Funding Reported", 
+                         if_else(!is.na(funding_names) & funding_refnum_yn==1, 
+                                 "Funding and Grant Reported", 
+                                 "Funding Not Reported")))
+
+data.funding$funding <- as.factor(data.funding$funding)
+
+# Count number of articles for funding condition by year of publication
+data.funding <- data.funding %>%
+  count(funding, article_year) %>%
+  ungroup()
+
+# Compute proportion of articles per sex category by year of publication
+data.funding <- data.funding %>%
+  group_by(article_year) %>%
+  mutate(total=sum(n)) %>%
+  group_by(funding, add=TRUE) %>%
+  mutate(per=round(100*n/total,3))
+data.funding$article_year <- as.numeric(data.funding$article_year)
+data.funding$per <- as.numeric(data.funding$per)
+
+# Stacked bar chart of reporting of funding and funding number by year of publication
+figure2A <- ggplot(data.funding, aes(x=article_year, y=per)) +
+  geom_col(aes(fill=factor(funding, levels=c("Funding Not Reported",
+                                            "Funding Reported",
+                                            "Funding and Grant Reported")))) +
+  theme_apa() +
+  xlab("Year of publication") +
+  ylab("Proportion of articles") +
+  labs(fill="") +
+  scale_fill_manual(values=c("#482677FF","#DCE319FF","#FDE725FF")) +
+  scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
+  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
+  theme(axis.title=element_text(face="bold", size=12), 
+        axis.text=element_text(colour="black", face="bold"), 
+        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
+        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))
+
+# Change funding names to lower case
+data.funding.name$funding_names <- tolower(data.funding.name$funding_names)
+
+# Create list of funding sources
+list.funding <- data.funding.name$funding_names
+
+# Split combinations of funding sources (multiple funding names)
+# which are delimited by commas
 list.funding <- as.list(strsplit(list.funding, ","))
@@
 
-# Count number of occurrences for each category
+# Count number of occurrences for each funding name
 table.list.funding <- as.data.frame(table(list.funding))
-n.funding.nih <- sum(str_count(list.funding, "nih"), str_count(list.funding, "national institute"))
+
+# Combine funding names that fall under the same agency
+table.list.funding$list.funding <- as.character(table.list.funding$list.funding)
+
+# NIH (USA)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("nih", table.list.funding$list.funding)== TRUE | grepl("national institute", table.list.funding$list.funding) == TRUE | grepl("nimh", table.list.funding$list.funding)== TRUE | grepl("nhlbi", table.list.funding$list.funding)== TRUE | grepl("national heart lung", table.list.funding$list.funding)== TRUE | grepl("national cancer institute", table.list.funding$list.funding)== TRUE | grepl("nia", table.list.funding$list.funding)== TRUE | grepl("ninds", table.list.funding$list.funding)== TRUE | grepl("ninr", table.list.funding$list.funding)== TRUE | grepl("ninda", table.list.funding$list.funding)== TRUE,
+                              "NI", table.list.funding$list.funding))
+# NHMRC (AUS)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("nhmrc", table.list.funding$list.funding)== TRUE | grepl("national health and medical research council", table.list.funding$list.funding)== TRUE,
+                              "NHMRC", table.list.funding$list.funding))
+
+# Wellcome Trust (GBR)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("wellcome", table.list.funding$list.funding)== TRUE,
+                              "Wellcome Trust", table.list.funding$list.funding))
+
+# CIHR (CAN)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("canadian institute", table.list.funding$list.funding)== TRUE,
+                              "CIHR", table.list.funding$list.funding))
+
+# NSFC (CHN)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("national natural science", table.list.funding$list.funding)== TRUE,
+                              "NSFC", table.list.funding$list.funding))
+
+# Academy of Finland (FIN)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("academy of finland", table.list.funding$list.funding)== TRUE,
+                              "Academy of Finland", table.list.funding$list.funding))
+
+# Inserm (FRA)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("inserm", table.list.funding$list.funding)== TRUE,
+                              "Inserm", table.list.funding$list.funding))
+
+# SNSF (CHE)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("swiss national", table.list.funding$list.funding)== TRUE,
+                              "SNSF", table.list.funding$list.funding))
+
+# Department of Veterans Affairs (USA)
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("veterans affairs", table.list.funding$list.funding)== TRUE,
+                              "Dept Vet Affairs", table.list.funding$list.funding))
+
+# European Commission
+table.list.funding <- table.list.funding %>%
+  mutate(list.funding=if_else(grepl("european commission", table.list.funding$list.funding)== TRUE,
+                              "EC", table.list.funding$list.funding))
+
+# Group by funding agency
+table.list.funding <- table.list.funding %>%
+  group_by(list.funding) %>%
+  summarise(N = sum(Freq))
 
@@
 data.funding.NI <- data.funding.name %>%
-  filter(grepl("NI", funding_names) | grepl("National Institute", funding_names))
+  filter(grepl("nih", funding_names) | grepl("national institute", funding_names) | grepl("nimh", funding_names) | grepl("nhlbi", funding_names) | grepl("national heart lung", funding_names) | grepl("national cancer institute", funding_names) | grepl("nia", funding_names) | grepl("ninr", funding_names) | grepl("ninda", funding_names) | grepl("ninds", funding_names))
+
 percent.funding.NI <- round(nrow(data.funding.NI)/as.numeric(nrow(data.included)) * 100, 2)
@@
 # Select articles that report funding number and funding from National Institutes in the US
-data.funding.NI.num <- data.funding.name %>%
-  filter(grepl("NI", funding_names) | grepl("National Institute", funding_names)) %>%
+data.funding.NI.num <- data.funding.NI %>%
   filter(funding_refnum_yn == 1)
+
 percent.funding.NI.num <- round(nrow(data.funding.NI.num)/as.numeric(nrow(data.funding.NI)) * 100, 2)
@@
   xlab("Age variable") +
-  scale_fill_viridis_c(end=0.6, direction=-1, labels=c("0.0","0.25","0.5","0.75","1")) +
+  scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
+                       limits=c(-1,1), labels=c("0.0","0.25","0.5","0.75","1")) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
         axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8), 
-        legend.position="none", legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_blank())
+        legend.position="right", legend.margin=margin(0,0,0,0),
+        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_text(face="bold"))
 
@@
   xlab("Age variable") +
-  scale_fill_viridis_c(end=0.6, direction=-1, labels=c("0.0","0.25","0.5","0.75","1")) +
+  scale_fill_viridis_c(end=0.6, direction=-1) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
         axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
-        legend.position="right", legend.margin=margin(0,0,0,0),
+        legend.position="none", legend.margin=margin(0,0,0,0),
         legend.box.margin=margin(-10,0,-10,-10), legend.title=element_blank())
@@
 
-# Change age from weeks to years, with 1 month = 4.3 weeks, and from months to years
+# Change age from days / weeks / months to years, with 1 month = 4.3 weeks = 30.4 days
 # age not in years is entered with "months" or "weeks"
@@
   mutate(age_reported_mean_value=ifelse(text=="weeks", 
-                                        age_reported_mean_value / 4.3 /12,
-                                        age_reported_mean_value / 12)) %>%
+                                        age_reported_mean_value / 4.3 / 12,
+                                        if_else(text=="days",
+                                                age_reported_mean_value / 30.4 / 12,
+                                                age_reported_mean_value / 12))) %>%
   select(article_pmid, article_year, age_reported_mean_value)# else case is in months
@@
   geom_point(size=1, color="#33638DFF") +
-  geom_text(x=1988, y=10, size=3.3,
-            label="Mean age = 281.5 - 0.1 * Year of publication") + # edit manually
-  geom_text(x=1988, y=7, size=3.1, label="R-Sq = 0.3%") + # edit manually
+  #geom_text(x=1988, y=10, size=3.3,
+            #label="Mean age = 281.5 - 0.1 * Year of publication") + # edit manually
+  #geom_text(x=1988, y=7, size=3.1, label="R-Sq = 0.3%") + # edit manually
   theme_apa() +
   ylab("Mean age (years)") +
-  xlab("Year") +
+  xlab("Year of publication") +
   scale_x_continuous(limits=c(1979,2019),breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(limits=c(0,60), expand=c(0,0)) +
+  scale_y_continuous(limits=c(0,80), expand=c(0,0)) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
 
-# Stacked bar chart  of reporting of other demographic variables by year of publication
-figure6.bars <- ggplot(data.demo, aes(x=variable, y=per, fill=report)) +
-  geom_bar(stat="identity") +
-  facet_grid(~article_year) +
-  theme_apa() +
-  xlab("Year of publication") +
-  ylab("Proportion of articles") +
-  labs(fill="Variable reported") +
-  scale_fill_manual(values=c("#482677FF","#D8DE29FF")) +
-  theme(axis.title=element_text(face="bold", size=12), 
-        axis.text=element_text(colour="black", face="bold"),
-        axis.text.x=element_text(size=5, angle=45, hjust=1),
-        axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
-        legend.title=element_text(face="bold", size=12),
-        legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10),
-        panel.spacing=unit(0.1, "lines"))
-
+# Scatter plot of reporting of other demographic variables by year of publication
 data.demo.reported <- data.demo %>%
@@
 
-figure6.points <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
+figure6 <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
   geom_point(size=2.5) +
@@
 data.education$education_reported_numcategory <- as.numeric(data.education$education_reported_numcategory)
+
 figure7A <- ggplot(data=data.education, aes(x=education_reported_numcategory)) +
@@
 `r round(percent.funding.NI.num)`% of the studies funded by the NIH also reported funding number.
+The second most represented funding agencies were the Australian National Health and Medical Research Council (NHMRC) and the Canadian Institutes of Health Research (CIHR).
+
+```{r}
+figure2A
+```
 
@@
 
-```{r, fig.cap="Geographical location of the studies. The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
+```{r, fig.cap="Funding sources and geographical location of the studies.
+A.
+Reporting of funding across the years.
+B.
+Distribution of articles by study location.
+The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
 figure2B
@@
 ```{r, fig.cap="Reporting of education, ethinicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
-#figure6.bars
-figure6.points
+figure6
 ```

### 2021-05-01 Demographics Review R markdown.Rmd
Updates
Edited graphs format and style.

@@
   theme_apa() +
-  ylab(label="Year") +
+  ylab(label="Year of publication") +
   xlab("Age variable") +
+  scale_y_continuous(breaks=c(1979,1989,1999,2009,2019), expand=c(0.02,0),
+                     labels=c("                   1979", "1989","1999","2009","2019")) +
   scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
@@
   theme_apa() +
-  ylab(label="Journal") +
+  ylab(label="Journal of publication") +
   xlab("Age variable") +
-  scale_fill_viridis_c(end=0.6, direction=-1) +
+  scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
+                       limits=c(-1,1), labels=c("0.0","0.25","0.5","0.75","1")) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
         axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
-        legend.position="none", legend.margin=margin(0,0,0,0),
-        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_blank())
+        legend.position="right", legend.margin=margin(0,0,0,0),
+        legend.box.margin=margin(-10,0,-10,-10), legend.title=element_text(face="bold"))
 
@@
   geom_point(size=1, color="#33638DFF") +
-  #geom_text(x=1988, y=10, size=3.3,
-            #label="Mean age = 281.5 - 0.1 * Year of publication") + # edit manually
-  #geom_text(x=1988, y=7, size=3.1, label="R-Sq = 0.3%") + # edit manually
   theme_apa() +
@@
         axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8),
+        legend.title=element_text(face="bold"),
         legend.margin=margin(0,0,0,0), legend.box.margin=margin(-10,0,-10,-10))
@@
   ylab("Proportion of articles") +
-  scale_color_viridis_d(name="Variable", 
+  scale_color_viridis_d(name="Variable reported", 
                         labels=c("Education","Ethnicity","Profession","Socio-economic status")) +
@@
 figure4C
-#grid.arrange(figure4A, figure4C, figure4B, nrow=2)
 ```

### 2021-05-08 article.Rmd
Update

--- 2021-04-30 Demographics Review R markdown.Rmd
@@
 ---
-title             : "Study Sample Characteristics in the Sleep and Chronobiology Research: A Systematic Review"
-shorttitle        : "Demographics analysis of sleep and chronobiology research"
+title             : "Study sample characteristics in chronobiology and sleep research"
+shorttitle        : "Study sample characteristics in chronobiology and sleep research"
 
@@
     role:
-      - wrote the original draft protocol, analyzed data, reviewed and edited the manuscript
+      - Data curation
+      - Investigation
+      - Software
+      - Visualization
+      - Writing – original draft
+      - Writing – review & editing
   - name          : "Rhiannon White"
@@
     role:         
-      - reviewed and scored articles, reviewed and edited the manuscript
+      - Investigation
+      - Writing – original draft
+      - Writing – review & editing
   - name          : "Manuel Spitschan"
@@
     role:
-      - wrote the original draft protocol, reviewed and edited the manuscript
+      - Conceptualization
+      - Data curation
+      - Funding acquisition
+      - Investigation
+      - Methodology
+      - Project administration
+      - Software
+      - Writing – original draft
+      - Writing – review & editing
 
@@
   - id            : "1"
-    institution   : "Department of Experimental Psychology, University of Oxford, Anna Watts Building, Radcliffe Observatory Quarter, Oxford, OX2 6GG, United Kingdom"
+    institution   : "Department of Experimental Psychology, University of Oxford"
   - id            : "2"
-    institution   : "Nuffield Department of Clinical Neurosciences, University of Oxford"
+    institution   : "Sleep and Circadian Neuroscience Institute, Nuffield Department of Clinical Neurosciences, University of Oxford"
   - id            : "3"
@@
 authornote: |
-  This systematic review is funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
+  Corresponding author: Manuel Spitschan (manuel.spitschan@psy.ox.ac.uk).
+This work ws funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
 abstract: |
-  Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex.
+  Many aspects of sleep and circadian physiology appear to be sensitive to participant-level characteristics.
 While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature.
-In this systematic review, we evaluated the study sample characteristics in the published sleep and chronobiology research over the past 40 years.
-
-  Articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals, identified by their five-year Impact Factor, were found through MEDLINE.
-6,777 articles were identified and a random sample of 20% was initially included for screening.
-Inclusion requirements included conducting original research, reporting human data, and recruiting volunteers.
-The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0=not reported), and any reported aggregate summary statistics for these variables were recorded.
-Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed including any of the demographic variables as covariates.
-
-  ~75% of screened articles met inclusion criteria.
-While >90% of studies reported age or sex, all other variables were reported in <10% of cases.
-We found that sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s.
-Overall, ~75% of studies recruited both male and female participants.
-Of studies recruiting a single sex, ~50% all-female studies focused on a sex-dependent feature, compared to <5% in all-male studies.
-
-  In this comprehensive review, we found that the majority of studies report at least sex or age, while many other important variables are typically not reported.
-Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate disaggregated data analyses.
-
-  <!-- https://tinyurl.com/ybremelq -->
-
-keywords          : "Study Sample Characteristics, Demographics, Sex, Systematic Review"
+Here, we investigated study sample characteristics in the published sleep and chronobiology research over the past 40 years.
+6,777 articles were identified and a random sample of 20% was included.
+The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored, and any reported aggregate summary statistics for these variables were recorded.
+We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases.
+Sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s.
+We found that the majority of studies report at least sex or age, while other variables are typically not reported.
+Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate meta analyses.
+
+keywords          : "demographics, ethnicity, sex, research participants, reporting, publishing, meta-science"
 wordcount         : "X"
 
-bibliography      : Demographics Review References.bib
+bibliography      : references.bib
 
 floatsintext      : yes
-figurelist        : no
+figurelist        : yes
 tablelist         : no
@@
 
-r_refs("Demographics Review References.bib")
+r_refs("references_r.bib")
 ```
@@
 # Set working directory
-setwd("~/Demographics Review")
+#setwd("~/Demographics Review")
+setwd("~/Projects/sleep_circadian_demographics_data")
 
 # Download scoring sheet 
-data.original <- read_excel("scoring.xlsx")
+data.original <- read_excel("data.xlsx")
 ```
@@
 
-  Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex.
-While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature.
-In this systematic review, we evaluated the study sample characteristics in the published sleep and chronobiology research over the past 40 years.
-
-  A recent study reviewed the reporting and analysis of sex in the biological sciences research [@RN1].
-The authors found that while sex inclusion has significantly increased over the past 10 years, sex-based analysis has not improved, despite recent policies.
+  Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex [@redline_effects_2004; @santhi_sex_2016; @cain_sex_2010; @anderson_sexual_2020; @mong_sleep_2011], age [@redline_effects_2004; @espiritu_aging-related_2008; @mander_sleep_2017; @desforges_sleep_1990; @li_sleep_2018; @bliwise_sleep_1993; @benloucif_responsiveness_2006; @duffy_aging_2015] and ethnicity [@ahn_scoping_2021; @goldstein_sleep_2020; @eastman_blacks_2012; @eastman_circadian_2016].
+More generally, there is a large literature on individual differences on sleep and circadian physiology [@van_dongen_individual_2005; @kerkhof_inter-individual_1985; @tankova_circadian_1994; @horne_individual_1977; @baehr_individual_2000; @burgess_individual_2008; @santhi_spectral_2012; @chellappa_individual_2021], demonstrating the need to consider participant-level data.
+
+The extent to which a scientific field's findings are generalisable depend very much on the representativeness of a given study sample.
+As an example, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020].
+The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014].
+
+While research findings converge on participant-level demographic characteristics affecting outcomes, it is not clear to what extent this information is available in the large body of already published literature, nor to what extent it is even reported.
+Here, we address the question of participant-level demographic characteristics (age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population) and reporting thereof in chronobiology and sleep research.
+Here, we extracted the study sample characteristics in a total of 1356 randomly sampled publications across the 8 top (ranked by the Journal Impact Factor) chronobiology and sleep research and subjected them to a comprehensive analysis.
 
@@
 
-  Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals  were found through MEDLINE by MS.
-The eight journals were selected based on their five-year Impact Factor, and included Journal of Pineal Research, Sleep, Journal of Sleep Research, Sleep Medicine, Journal of Clinical Medicine, Journal of Biological Rhythms, Behavioral Sleep Medicine, and Chronobiology International.
-6,777 articles were identified and a random sample of 20% was initially selected for screening.
+*Procedure*.
+Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals were considered.
+The listing of possible of target journals was based on a previously established list of journals implementing a hybrid strategy by consulting the Web of Science Master Journal List, domain-relevant expertise in sleep and chronobiology and consulting with a senior researcher with >25 years of experience in the field [@spitschan_transparency_2020].
+From this previously derived list, we selected eight journals were selected based on their five-year Impact Factor, and included *Journal of Pineal Research* (ISSN: 0742-3098 / 1600-079X; 2018 5-year IF: 12.197), *Sleep*  (0161-8105 / 1550-9109; 5.588), *Journal of Sleep Research* (0962-1105 / 1365-2869; 3.951), *Sleep Medicine* (1389-9457 / 1878-5506; 3.934), *Journal of Clinical Sleep Medicine* (1550-9389 / 1550-9397; 3.855), *Journal of Biological Rhythms* (0748-7304 / 1552-4531; 3.349), Behavioral Sleep Medicine (1540-2002 / 1540-2010; 3.162), and *Chronobiology International* (0742-0528 / 1525-6073; 2.998).
+While *Sleep Medicine Reviews* also features in the list of journals, we did not include it as it primarily publishes reviews.
+
+*Article inclusion*.
+6,777 articles were identified through a MEDLINE search by the journal and including odd years.
+A random sample of 20% was initially selected for screening.
 Inclusion requirements included conducting original research in the English language, reporting human data, and recruiting volunteers.
-As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, erratums and corrigendums were excluded.
-All articles were reviewed for eligibility and coded by ST.
-The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded.
+As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, errata and corrigenda were excluded.
+
+*Review and article extraction*.
+All included articles were reviewed for eligibility and coded by RW.
+The reporting of sample size, age, sex, gender, ethnicity, level of education, socioeconomic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded.
 Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.
@@
 # Stacked bar chart of included articles by journal and year of publication
-figure1A <- ggplot(data.included, aes(x=article_year)) +
+figure1 <- ggplot(data.included, aes(x=article_year)) +
   geom_bar(stat="count", aes(fill=article_journal)) +
@@
   labs(fill="Journal") +
-  ggtitle("A") +
+  ggtitle("Included and analysed articles by journal and year") +
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
@@
 # Stacked bar chart of excluded articles by reason for exclusion and year of publication
-figure1B <- ggplot(data.eligibility, aes(x=article_year, y=per)) +
+figure2 <- ggplot(data.eligibility, aes(x=article_year, y=per)) +
   geom_col(aes(fill=factor(eligibility, levels=c("Not an original research article",
@@
   labs(fill="Eligibility") +
-  ggtitle("B") +
+  ggtitle("Inclusion and reasons for exclusion") +
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
@@
 ```
+
+  Data were coded in an Excel Spreadsheet and analyzed in R Studio (version 4.0.5).
+Reporting of funding, geographical location, and number of sub-studies for each article were investigated for the sample of articles that passed all eligibility criteria.
+
+*Pre-registration*.
+We pre-registered our protocol (specified using the PRISMA-P template [@shamseer_preferred_2015; @prisma-p_group_preferred_2015]) on the Open Science Framework (https://osf.io/cu3we/).
+
+*Materials, data and code availability*.
+All data underlying this manuscript are available on a public GitHub repository (https://github.com/hcvnl/sleep_circadian_demographics_data).
+The article was written in R [@R-base] using RMarkdown and papaja [@R-papaja], employing a series of additional R packages [@R-corrplot2017; @R-dplyr; @R-ggExtra; @R-ggformula; @R-ggplot2; @R-ggridges; @R-ggsci; @R-ggstance; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lattice; @R-latticeExtra; @R-lemon; @R-Matrix; @R-mosaic; @R-mosaicData; @R-pracma; @R-readr; @R-readxl; @R-reshape2; @R-stringr; @R-tibble; @R-tidyr; @R-BayesFactor] and is fully reproducible.
 
@@
   mutate(funding=if_else(!is.na(funding_names) & funding_refnum_yn==0, 
-                         "Funding Reported", 
+                         "Funding reported", 
                          if_else(!is.na(funding_names) & funding_refnum_yn==1, 
-                                 "Funding and Grant Reported", 
-                                 "Funding Not Reported")))
+                                 "Funding and grant reported", 
+                                 "Funding not reported")))
 
@@
 # Stacked bar chart of reporting of funding and funding number by year of publication
-figure2A <- ggplot(data.funding, aes(x=article_year, y=per)) +
-  geom_col(aes(fill=factor(funding, levels=c("Funding Not Reported",
-                                            "Funding Reported",
-                                            "Funding and Grant Reported")))) +
+figure3 <- ggplot(data.funding, aes(x=article_year, y=per)) +
+  geom_col(aes(fill=factor(funding, levels=c("Funding not reported",
+                                            "Funding reported",
+                                            "Funding and grant reported")))) +
   theme_apa() +
@@
 # Stacked bar chart of 8 most represented study countries by year of publication
-figure2B <- ggplot(data.country, aes(x=article_year, y=per)) +
+figure4 <- ggplot(data.country, aes(x=article_year, y=per)) +
   geom_col(aes(fill=factor(study_country_values, 
@@
 
-  Data were coded in an Excel Spreadsheet and analyzed in R Studio (version 4.0.5).
-Reporting of funding, geographical location, and number of sub-studies for each article were investigated for the sample of articles that passed all eligibility criteria.
-The rest of the analysis was conducted on the subset of articles that reported a single study (`r round(percent.study.single)`% of articles), i.e.
-a single sample population.
-
 ```{r Sample size, echo=FALSE, message=FALSE, warning=FALSE}
@@
 
-figure3 <- ggplot(data.ss, aes(x=article_year, y=sample_size_total_reported_values)) +
+figure5 <- ggplot(data.ss, aes(x=article_year, y=sample_size_total_reported_values)) +
   geom_violin(fill="#440154FF") +
@@
 # Heat map of reporting of age variables by year of publication
-figure4A <- ggplot(data=age.corr.year, aes(x=variable, y=article_year, fill=value)) +
+figure6 <- ggplot(data=age.corr.year, aes(x=variable, y=article_year, fill=value)) +
   geom_tile() +
   theme_apa() +
-  ylab(label="Year") +
+  ylab(label="Year of publication") +
   xlab("Age variable") +
+  scale_y_continuous(breaks=c(1979,1989,1999,2009,2019), expand=c(0.02,0),
+                     labels=c("                   1979", "1989","1999","2009","2019")) +
   scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
@@
 # Heat map of reporting of age variables by journal of publication
-figure4B <- ggplot(data=age.corr.journal, 
+figure7 <- ggplot(data=age.corr.journal, 
                    aes(x=variable, y=article_journal, fill=value)) +
@@
 # Violin plot with regression line and equation for mean age by year of publication
-figure4C <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
+figure8 <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
   geom_smooth(method="lm", color="#440154FF") +
   geom_point(size=1, color="#33638DFF") +
-  #geom_text(x=1988, y=10, size=3.3,
-            #label="Mean age = 281.5 - 0.1 * Year of publication") + # edit manually
-  #geom_text(x=1988, y=7, size=3.1, label="R-Sq = 0.3%") + # edit manually
   theme_apa() +
@@
 # Stacked bar chart of sex reporting by year of publication
-figure5A <- ggplot(data.sex, aes(x=article_year, y=per)) +
+figure9 <- ggplot(data.sex, aes(x=article_year, y=per)) +
   geom_col(aes(fill=factor(report, levels=c("Unspecified",
@@
 
-figure6 <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
+figure10 <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
   geom_point(size=2.5) +
@@
 
-figure7A <- ggplot(data=data.education, aes(x=education_reported_numcategory)) +
+figure11A <- ggplot(data=data.education, aes(x=education_reported_numcategory)) +
   geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
@@
 data.ethnicity$ethnicity_reported_numcategory <- as.numeric(data.ethnicity$ethnicity_reported_numcategory)
-figure7B <- ggplot(data=data.ethnicity, aes(x=ethnicity_reported_numcategory)) +
+figure11B <- ggplot(data=data.ethnicity, aes(x=ethnicity_reported_numcategory)) +
   geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
@@
 data.profession$profession_reported_numcategory <- as.numeric(data.profession$profession_reported_numcategory)
-figure7C <- ggplot(data=data.profession, aes(x=profession_reported_numcategory)) +
+figure11C <- ggplot(data=data.profession, aes(x=profession_reported_numcategory)) +
   geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
@@
 data.ses$ses_reported_numcategory <- as.numeric(data.ses$ses_reported_numcategory)
-figure7D <- ggplot(data=data.ses, aes(x=ses_reported_numcategory)) +
+figure11D <- ggplot(data=data.ses, aes(x=ses_reported_numcategory)) +
   geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
@@
 # Stacked bar chart of the use of demographic variables in the analysis by year of publication
-figure8 <- ggplot(data.analysis, aes(x=article_year, y=per)) +
+figure12 <- ggplot(data.analysis, aes(x=article_year, y=per)) +
   geom_col(aes(fill=report)) +
@@
 
-```{r}
-figure1A
-```
-
-```{r, fig.cap="Inclusion and exclusion of articles.
+```{r, fig.cap="Funding sources and geographical location of the studies.
 A.
-Number of articles included by journal and year of publication.
+Reporting of funding across the years.
 B.
-Proportion of articles included and excluded by reason for exclusion and year of publication.", warning=FALSE, message=FALSE}
-figure1B
+Distribution of articles by study location.
+The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
+figure1
+```
+
+```{r, fig.cap="Funding sources and geographical location of the studies.
+A.
+Reporting of funding across the years.
+B.
+Distribution of articles by study location.
+The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
+figure2
 ```
@@
 
-```{r}
-figure2A
+```{r, fig.cap="Funding sources and geographical location of the studies.
+A.
+Reporting of funding across the years.
+B.
+Distribution of articles by study location.
+The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
+figure3
 ```
@@
 The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
-figure2B
+figure4
 ```
@@
 ```{r, fig.cap="Sample size of the recruited volunteers. Numbers are computed with a log-10 scale.", warning=FALSE, message=FALSE}
-figure3
+figure5
 ```
@@
 
-```{r, warning=FALSE, message=FALSE}
-figure4A
-figure4B
+```{r, fig.cap="To be completed.", warning=FALSE, message=FALSE}
+figure6
+```
+```{r, fig.cap="To be completed.", warning=FALSE, message=FALSE}
+figure7
 ```
@@
 Mean age by year of publication.", message=FALSE, warning=FALSE}
-figure4C
-#grid.arrange(figure4A, figure4C, figure4B, nrow=2)
+figure8
 ```
@@
 ```{r, fig.cap="Sex inclusion. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
-figure5A
+figure9
 ```
@@
 ```{r, fig.cap="Reporting of education, ethinicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
-figure6
+figure10
 ```
@@
 ```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
-grid.arrange(figure7A, figure7B, figure7C, figure7D, nrow=2)
+grid.arrange(figure11A, figure11B, figure11C, figure11D, nrow=2)
 ```
@@
 ```{r, fig.cap="Use of study population characteristics as variables in the analysis."}
-figure8
+figure12
 ```
@@
 # Discussion
+
+In the clinical domain, the need to time therapy based on a patient's individual circadian rhythm has more recently become the focus of the emerging field of chronotherapy or chronotherapeutics [@adam_core_2019; @greco_personalized_2020; @dijk_novel_2020; @hill_optimizing_2020].
+
+The ability to generalise findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories.
+The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020].
 

### 2021-05-08 article.tex
Updates

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  jou,floatsintext]{apa6}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Study sample characteristics in chronobiology and sleep research},
  pdfauthor={Selma Tir1,2, Rhiannon White1,3, \& Manuel Spitschan1,4,5},
  pdflang={en-EN},
  pdfkeywords={demographics, ethnicity, sex, research participants, reporting, publishing, meta-science},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\begin{center}\begin{threeparttable}}
%   {\end{threeparttable}\end{center}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\begin{center}\begin{ThreePartTable}}{\end{ThreePartTable}\end{center}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Study sample characteristics in chronobiology and sleep research}
\keywords{demographics, ethnicity, sex, research participants, reporting, publishing, meta-science\newline\indent Word count: X}
\usepackage{dblfloatfix}

\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage[titles]{tocloft}
\cftpagenumbersoff{figure}
\renewcommand{\cftfigpresnum}{\itshape\figurename\enspace}
\renewcommand{\cftfigaftersnum}{.\space}
\setlength{\cftfigindent}{0pt}
\setlength{\cftafterloftitleskip}{0pt}
\settowidth{\cftfignumwidth}{Figure 10.\qquad}
\ifXeTeX
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\title{Study sample characteristics in chronobiology and sleep research}
\author{Selma Tir\textsuperscript{1,2}, Rhiannon White\textsuperscript{1,3}, \& Manuel Spitschan\textsuperscript{1,4,5}}
\date{}

\authornote{

Corresponding author: Manuel Spitschan (\href{mailto:manuel.spitschan@psy.ox.ac.uk}{\nolinkurl{manuel.spitschan@psy.ox.ac.uk}}). This work ws funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment -- Diversity \& Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).

The authors made the following contributions. Selma Tir: Data curation, Investigation, Software, Visualization, Writing -- original draft, Writing -- review \& editing; Rhiannon White: Investigation, Writing -- original draft, Writing -- review \& editing; Manuel Spitschan: Conceptualization, Data curation, Funding acquisition, Investigation, Methodology, Project administration, Software, Writing -- original draft, Writing -- review \& editing.

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Department of Experimental Psychology, University of Oxford\\\textsuperscript{2} Sleep and Circadian Neuroscience Institute, Nuffield Department of Clinical Neurosciences, University of Oxford\\\textsuperscript{3} Warwick Medical School, University of Warwick, United Kingdom\\\textsuperscript{4} Centre for Chronobiology, Psychiatric Hospital of the University of Basel, Switzerland\\\textsuperscript{5} Transfaculty Research Platform Molecular and Cognitive Neurosciences, University of Basel, Switzerland}

\abstract{
Many aspects of sleep and circadian physiology appear to be sensitive to participant-level characteristics. While recent research robustly highlights the importance of considering participant-level demographic information, it is not clear to what extent this information is available in the large body of already published literature. Here, we investigated study sample characteristics in the published sleep and chronobiology research over the past 40 years. 6,777 articles were identified and a random sample of 20\% was included. The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored, and any reported aggregate summary statistics for these variables were recorded. We found that while \textgreater90\% of studies reported age or sex, all other variables were reported in \textless10\% of cases. Sex balance greatly changed over the years, from a \textasciitilde3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s. We found that the majority of studies report at least sex or age, while other variables are typically not reported. Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate meta analyses.
}

\begin{document}
\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex (Anderson \& FitzGerald, 2020; Cain et al., 2010; Mong et al., 2011; Redline et al., 2004; Santhi et al., 2016), age (Benloucif et al., 2006; Bliwise, 1993; Desforges, Prinz, Vitiello, Raskind, \& Thorpy, 1990; Duffy, Zitting, \& Chinoy, 2015; Espiritu, 2008; Li, Vitiello, \& Gooneratne, 2018; Mander, Winer, \& Walker, 2017; Redline et al., 2004) and ethnicity (Ahn et al., 2021; Eastman, Molina, Dziepak, \& Smith, 2012; Eastman, Tomaka, \& Crowley, 2016; Goldstein, Gaston, McGrath, \& Jackson, 2020). More generally, there is a large literature on individual differences on sleep and circadian physiology (Baehr, Revelle, \& Eastman, 2000; Burgess \& Fogg, 2008; Chellappa, 2021; Dongen, Vitellaro, \& Dinges, 2005; Horne \& Östberg, 1977; Kerkhof, 1985; Santhi et al., 2012; Tankova, Adan, \& Buela-Casal, 1994), demonstrating the need to consider participant-level data.

The extent to which a scientific field's findings are generalisable depend very much on the representativeness of a given study sample. As an example, a recent study reviewed the reporting and analysis of sex in biological sciences research (Woitowich, Beery, \& Woodruff, 2020). The authors found that while sex inclusion has significantly increased over the past 10 years (Beery \& Zucker, 2011), sex-based analysis has not improved, despite recent policies and funder mandates (Clayton \& Collins, 2014).

While research findings converge on participant-level demographic characteristics affecting outcomes, it is not clear to what extent this information is available in the large body of already published literature, nor to what extent it is even reported. Here, we address the question of participant-level demographic characteristics (age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population) and reporting thereof in chronobiology and sleep research. Here, we extracted the study sample characteristics in a total of 1356 randomly sampled publications across the 8 top (ranked by the Journal Impact Factor) chronobiology and sleep research and subjected them to a comprehensive analysis.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\emph{Procedure}. Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals were considered. The listing of possible of target journals was based on a previously established list of journals implementing a hybrid strategy by consulting the Web of Science Master Journal List, domain-relevant expertise in sleep and chronobiology and consulting with a senior researcher with \textgreater25 years of experience in the field (Spitschan, Schmidt, \& Blume, 2020). From this previously derived list, we selected eight journals were selected based on their five-year Impact Factor, and included \emph{Journal of Pineal Research} (ISSN: 0742-3098 / 1600-079X; 2018 5-year IF: 12.197), \emph{Sleep} (0161-8105 / 1550-9109; 5.588), \emph{Journal of Sleep Research} (0962-1105 / 1365-2869; 3.951), \emph{Sleep Medicine} (1389-9457 / 1878-5506; 3.934), \emph{Journal of Clinical Sleep Medicine} (1550-9389 / 1550-9397; 3.855), \emph{Journal of Biological Rhythms} (0748-7304 / 1552-4531; 3.349), Behavioral Sleep Medicine (1540-2002 / 1540-2010; 3.162), and \emph{Chronobiology International} (0742-0528 / 1525-6073; 2.998). While \emph{Sleep Medicine Reviews} also features in the list of journals, we did not include it as it primarily publishes reviews.

\emph{Article inclusion}. 6,777 articles were identified through a MEDLINE search by the journal and including odd years. A random sample of 20\% was initially selected for screening. Inclusion requirements included conducting original research in the English language, reporting human data, and recruiting volunteers. As such, animal studies, bibliographies, case reports, comments, conference proceedings, editorials, guidelines, letters, retracted publications, reviews, errata and corrigenda were excluded.

\emph{Review and article extraction}. All included articles were reviewed for eligibility and coded by RW. The reporting of sample size, age, sex, gender, ethnicity, level of education, socioeconomic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded. Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.

Data were coded in an Excel Spreadsheet and analyzed in R Studio (version 4.0.5). Reporting of funding, geographical location, and number of sub-studies for each article were investigated for the sample of articles that passed all eligibility criteria.

\emph{Pre-registration}. We pre-registered our protocol (specified using the PRISMA-P template (PRISMA-P Group et al., 2015; Shamseer et al., 2015)) on the Open Science Framework (\url{https://osf.io/cu3we/}).

\emph{Materials, data and code availability}. All data underlying this manuscript are available on a public GitHub repository (\url{https://github.com/hcvnl/sleep_circadian_demographics_data}). The article was written in R (R Core Team, 2021) using RMarkdown and papaja (Aust \& Barth, 2020), employing a series of additional R packages (Arnold, 2021; Attali \& Baker, 2019; Auguie, 2017; Bates \& Maechler, 2021; Borchers, 2021; Edwards, 2020; Henry, Wickham, \& Chang, 2020; Kaplan \& Pruim, 2021; Morey \& Rouder, 2018; Müller \& Wickham, 2021; Pruim, Kaplan, \& Horton, 2021; Pruim, Kaplan, \& Horton, 2017; Sarkar, 2008; Sarkar \& Andrews, 2019; Wei \& Simko, 2017; Wickham, 2007, 2016, 2019, 2021; Wickham \& Bryan, 2019; Wickham, François, Henry, \& Müller, 2021; Wickham \& Hester, 2020; Wilke, 2021; Xiao, 2018; Xie, 2015) and is fully reproducible.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{inclusion-exclusion}{%
\subsubsection{Inclusion \& Exclusion}\label{inclusion-exclusion}}

85\% of the sampled articles met eligibility criteria during the review.

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-1-1.pdf}
\caption{\label{fig:unnamed-chunk-1}Funding sources and geographical location of the studies. A. Reporting of funding across the years. B. Distribution of articles by study location. The eight most represented countries are individually shown.}
\end{figure}

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{\label{fig:unnamed-chunk-2}Funding sources and geographical location of the studies. A. Reporting of funding across the years. B. Distribution of articles by study location. The eight most represented countries are individually shown.}
\end{figure}

\hypertarget{funding}{%
\subsubsection{Funding}\label{funding}}

Funding sources were reported by 62\% of studies, while funding number was also reported in 69\% of these cases. Overall, funding by the United States' National Institutes of Health (NIH) represented 19\% of the reported funding agencies. 92\% of the studies funded by the NIH also reported funding number. The second most represented funding agencies were the Australian National Health and Medical Research Council (NHMRC) and the Canadian Institutes of Health Research (CIHR).

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{\label{fig:unnamed-chunk-3}Funding sources and geographical location of the studies. A. Reporting of funding across the years. B. Distribution of articles by study location. The eight most represented countries are individually shown.}
\end{figure}

\hypertarget{geographical-location}{%
\subsubsection{Geographical location}\label{geographical-location}}

93\% of articles were conducted in a single country. The geographical location of the study was explicitly reported in 57\% of studies. The country of study was inferred for the rest of the sampled articles. Inference was primarily based on the first author's affiliation. Overall, 53 countries were represented. 77\% of articles reported multiple countries of study. Figure 2B shows the distribution of study location across time with the eight most represented countries.

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-4-1.pdf}
\caption{\label{fig:unnamed-chunk-4}Funding sources and geographical location of the studies. A. Reporting of funding across the years. B. Distribution of articles by study location. The eight most represented countries are individually shown.}
\end{figure}

\hypertarget{sample-size}{%
\subsubsection{Sample size}\label{sample-size}}

Sample size was reported in 92\% of studies, while 98\% investigated a single sampled population.

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-5-1.pdf}
\caption{\label{fig:unnamed-chunk-5}Sample size of the recruited volunteers. Numbers are computed with a log-10 scale.}
\end{figure}

\hypertarget{age}{%
\subsubsection{Age}\label{age}}

93\% of articles reported a variable describing age. Figure 4A and 4B show the use of various age variables by year and journal of publication. Overall, the mean age of the study populations was 39 years old.

\includegraphics{article_files/figure-latex/unnamed-chunk-6-1.pdf}
\includegraphics{article_files/figure-latex/unnamed-chunk-7-1.pdf}

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-8-1.pdf}
\caption{\label{fig:unnamed-chunk-8}Reporting of age variables. Correlation between the reporting of the mean, standrad deviation of the mean, median, minimum, maximum, interquartile range and other age variables by A. year of publication and B. journal of publication. C. Mean age by year of publication.}
\end{figure}

\hypertarget{sex}{%
\subsubsection{Sex}\label{sex}}

Sex was reported in 89\% of the studies. Figure 5 displays the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants. 13\% of the studies reporting sex only recruited male participants, while 10\% only employed females. Out of the studies focusing on a single gender, 1\% of the male studies focused on a sex dependent feature, while 2\% of the female studies did. 4\% of studies reported age by sex.

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-9-1.pdf}
\caption{\label{fig:unnamed-chunk-9}Sex inclusion. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.}
\end{figure}

\hypertarget{ethnicity-education-profession-and-socio-economic-status}{%
\subsubsection{Ethnicity, education, profession and socio-economic status}\label{ethnicity-education-profession-and-socio-economic-status}}

Other demographics variables were reported in 12\% of studies for education, 15\% for ethnicity, 2\% for profession, and 4\% for socio-economic status. Figure 6 shows the distribution of this reporting across the years. Figures 7 show the number of categories reported for each variable.

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-10-1.pdf}
\caption{\label{fig:unnamed-chunk-10}Reporting of education, ethinicity, profession and socio-economic status.}
\end{figure}

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-11-1.pdf}
\caption{\label{fig:unnamed-chunk-11}Number of categories reported for education, ethnicity, profession and socio-economic status.}
\end{figure}

\hypertarget{study-focus}{%
\subsubsection{Study focus}\label{study-focus}}

3\% of articles focused on a sex dependent feature, while 50\% investigated a clinical feature. 1\% of studies focused on twins, 1\% on pregnant women, 2\% on shift workers and 4\% on university students.

\hypertarget{analysis-disaggregation}{%
\subsubsection{Analysis disaggregation}\label{analysis-disaggregation}}

A

\begin{figure}
\centering
\includegraphics{article_files/figure-latex/unnamed-chunk-12-1.pdf}
\caption{\label{fig:unnamed-chunk-12}Use of study population characteristics as variables in the analysis.}
\end{figure}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In the clinical domain, the need to time therapy based on a patient's individual circadian rhythm has more recently become the focus of the emerging field of chronotherapy or chronotherapeutics (Adam, 2019; Dijk \& Duffy, 2020; Greco \& Sassone-Corsi, 2020; Hill, Innominato, Lévi, \& Ballesta, 2020).

The ability to generalise findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories. The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts (Henrich, Heine, \& Norenzayan, 2010; Muthukrishna et al., 2020).

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-adam_core_2019}{}}%
Adam, D. (2019). Core {Concept}: {Emerging} science of chronotherapy offers big opportunities to optimize drug delivery. \emph{Proceedings of the National Academy of Sciences}, \emph{116}(44), 21957--21959. \url{https://doi.org/10.1073/pnas.1916118116}

\leavevmode\vadjust pre{\hypertarget{ref-ahn_scoping_2021}{}}%
Ahn, S., Lobo, J. M., Logan, J. G., Kang, H., Kwon, Y., \& Sohn, M.-W. (2021). A scoping review of racial/ethnic disparities in sleep. \emph{Sleep Medicine}, \emph{81}, 169--179. \url{https://doi.org/10.1016/j.sleep.2021.02.027}

\leavevmode\vadjust pre{\hypertarget{ref-anderson_sexual_2020}{}}%
Anderson, S. T., \& FitzGerald, G. A. (2020). Sexual dimorphism in body clocks. \emph{Science (New York, N.Y.)}, \emph{369}(6508), 1164--1165. \url{https://doi.org/10.1126/science.abd4964}

\leavevmode\vadjust pre{\hypertarget{ref-R-ggthemes}{}}%
Arnold, J. B. (2021). \emph{Ggthemes: Extra themes, scales and geoms for 'ggplot2'}. Retrieved from \url{https://CRAN.R-project.org/package=ggthemes}

\leavevmode\vadjust pre{\hypertarget{ref-R-ggExtra}{}}%
Attali, D., \& Baker, C. (2019). \emph{ggExtra: Add marginal histograms to 'ggplot2', and more 'ggplot2' enhancements}. Retrieved from \url{https://CRAN.R-project.org/package=ggExtra}

\leavevmode\vadjust pre{\hypertarget{ref-R-gridExtra}{}}%
Auguie, B. (2017). \emph{gridExtra: Miscellaneous functions for "grid" graphics}. Retrieved from \url{https://CRAN.R-project.org/package=gridExtra}

\leavevmode\vadjust pre{\hypertarget{ref-R-papaja}{}}%
Aust, F., \& Barth, M. (2020). \emph{{papaja}: {Create} {APA} manuscripts with {R Markdown}}. Retrieved from \url{https://github.com/crsh/papaja}

\leavevmode\vadjust pre{\hypertarget{ref-baehr_individual_2000}{}}%
Baehr, E. K., Revelle, W., \& Eastman, C. I. (2000). Individual differences in the phase and amplitude of the human circadian temperature rhythm: With an emphasis on morningness-eveningness: {Phase} and amplitude of the temperature rhythm. \emph{Journal of Sleep Research}, \emph{9}(2), 117--127. \url{https://doi.org/10.1046/j.1365-2869.2000.00196.x}

\leavevmode\vadjust pre{\hypertarget{ref-R-Matrix}{}}%
Bates, D., \& Maechler, M. (2021). \emph{Matrix: Sparse and dense matrix classes and methods}. Retrieved from \url{https://CRAN.R-project.org/package=Matrix}

\leavevmode\vadjust pre{\hypertarget{ref-beery_sex_2011}{}}%
Beery, A. K., \& Zucker, I. (2011). Sex bias in neuroscience and biomedical research. \emph{Neuroscience and Biobehavioral Reviews}, \emph{35}(3), 565--572. \url{https://doi.org/10.1016/j.neubiorev.2010.07.002}

\leavevmode\vadjust pre{\hypertarget{ref-benloucif_responsiveness_2006}{}}%
Benloucif, S., Green, K., L'Hermite-Balériaux, M., Weintraub, S., Wolfe, L. F., \& Zee, P. C. (2006). Responsiveness of the aging circadian clock to light. \emph{Neurobiology of Aging}, \emph{27}(12), 1870--1879. \url{https://doi.org/10.1016/j.neurobiolaging.2005.10.011}

\leavevmode\vadjust pre{\hypertarget{ref-bliwise_sleep_1993}{}}%
Bliwise, D. L. (1993). Sleep in {Normal} {Aging} and {Dementia}. \emph{Sleep}, \emph{16}(1), 40--81. \url{https://doi.org/10.1093/sleep/16.1.40}

\leavevmode\vadjust pre{\hypertarget{ref-R-pracma}{}}%
Borchers, H. W. (2021). \emph{Pracma: Practical numerical math functions}. Retrieved from \url{https://CRAN.R-project.org/package=pracma}

\leavevmode\vadjust pre{\hypertarget{ref-burgess_individual_2008}{}}%
Burgess, H. J., \& Fogg, L. F. (2008). Individual {Differences} in the {Amount} and {Timing} of {Salivary} {Melatonin} {Secretion}. \emph{PLoS ONE}, \emph{3}(8), e3055. \url{https://doi.org/10.1371/journal.pone.0003055}

\leavevmode\vadjust pre{\hypertarget{ref-cain_sex_2010}{}}%
Cain, S. W., Dennison, C. F., Zeitzer, J. M., Guzik, A. M., Khalsa, S. B. S., Santhi, N., \ldots{} Duffy, J. F. (2010). Sex differences in phase angle of entrainment and melatonin amplitude in humans. \emph{Journal of Biological Rhythms}, \emph{25}(4), 288--296. \url{https://doi.org/10.1177/0748730410374943}

\leavevmode\vadjust pre{\hypertarget{ref-chellappa_individual_2021}{}}%
Chellappa, S. L. (2021). Individual differences in light sensitivity affect sleep and circadian rhythms. \emph{Sleep}, \emph{44}(2), zsaa214. \url{https://doi.org/10.1093/sleep/zsaa214}

\leavevmode\vadjust pre{\hypertarget{ref-clayton_policy_2014}{}}%
Clayton, J. A., \& Collins, F. S. (2014). Policy: {NIH} to balance sex in cell and animal studies. \emph{Nature}, \emph{509}(7500), 282--283. \url{https://doi.org/10.1038/509282a}

\leavevmode\vadjust pre{\hypertarget{ref-desforges_sleep_1990}{}}%
Desforges, J. F., Prinz, P. N., Vitiello, M. V., Raskind, M. A., \& Thorpy, M. J. (1990). Sleep {Disorders} and {Aging}. \emph{New England Journal of Medicine}, \emph{323}(8), 520--526. \url{https://doi.org/10.1056/NEJM199008233230805}

\leavevmode\vadjust pre{\hypertarget{ref-dijk_novel_2020}{}}%
Dijk, D.-J., \& Duffy, J. F. (2020). Novel {Approaches} for {Assessing} {Circadian} {Rhythmicity} in {Humans}: {A} {Review}. \emph{Journal of Biological Rhythms}, \emph{35}(5), 421--438. \url{https://doi.org/10.1177/0748730420940483}

\leavevmode\vadjust pre{\hypertarget{ref-van_dongen_individual_2005}{}}%
Dongen, H. P. A. van, Vitellaro, K. M., \& Dinges, D. F. (2005). Individual {Differences} in {Adult} {Human} {Sleep} and {Wakefulness}: {Leitmotif} for a {Research} {Agenda}. \emph{Sleep}, \emph{28}(4), 479--498. \url{https://doi.org/10.1093/sleep/28.4.479}

\leavevmode\vadjust pre{\hypertarget{ref-duffy_aging_2015}{}}%
Duffy, J. F., Zitting, K.-M., \& Chinoy, E. D. (2015). Aging and {Circadian} {Rhythms}. \emph{Sleep Medicine Clinics}, \emph{10}(4), 423--434. \url{https://doi.org/10.1016/j.jsmc.2015.08.002}

\leavevmode\vadjust pre{\hypertarget{ref-eastman_blacks_2012}{}}%
Eastman, C. I., Molina, T. A., Dziepak, M. E., \& Smith, M. R. (2012). Blacks ({African} {Americans}) have shorter free-running circadian periods than whites ({Caucasian} {Americans}). \emph{Chronobiology International}, \emph{29}(8), 1072--1077. \url{https://doi.org/10.3109/07420528.2012.700670}

\leavevmode\vadjust pre{\hypertarget{ref-eastman_circadian_2016}{}}%
Eastman, C. I., Tomaka, V. A., \& Crowley, S. J. (2016). Circadian rhythms of {European} and {African}-{Americans} after a large delay of sleep as in jet lag and night work. \emph{Scientific Reports}, \emph{6}(1), 36716. \url{https://doi.org/10.1038/srep36716}

\leavevmode\vadjust pre{\hypertarget{ref-R-lemon}{}}%
Edwards, S. M. (2020). \emph{Lemon: Freshing up your 'ggplot2' plots}. Retrieved from \url{https://CRAN.R-project.org/package=lemon}

\leavevmode\vadjust pre{\hypertarget{ref-espiritu_aging-related_2008}{}}%
Espiritu, J. R. D. (2008). Aging-{Related} {Sleep} {Changes}. \emph{Clinics in Geriatric Medicine}, \emph{24}(1), 1--14. \url{https://doi.org/10.1016/j.cger.2007.08.007}

\leavevmode\vadjust pre{\hypertarget{ref-goldstein_sleep_2020}{}}%
Goldstein, S. J., Gaston, S. A., McGrath, J. A., \& Jackson, C. L. (2020). Sleep {Health} and {Serious} {Psychological} {Distress}: {A} {Nationally} {Representative} {Study} of the {United} {States} among {White}, {Black}, and {Hispanic}/{Latinx} {Adults}. \emph{Nature and Science of Sleep}, \emph{12}, 1091--1104. \url{https://doi.org/10.2147/NSS.S268087}

\leavevmode\vadjust pre{\hypertarget{ref-greco_personalized_2020}{}}%
Greco, C. M., \& Sassone-Corsi, P. (2020). Personalized medicine and circadian rhythms: {Opportunities} for modern society. \emph{Journal of Experimental Medicine}, \emph{217}(6), e20200702. \url{https://doi.org/10.1084/jem.20200702}

\leavevmode\vadjust pre{\hypertarget{ref-henrich_weirdest_2010}{}}%
Henrich, J., Heine, S. J., \& Norenzayan, A. (2010). The weirdest people in the world? \emph{Behavioral and Brain Sciences}, \emph{33}(2-3), 61--83. \url{https://doi.org/10.1017/S0140525X0999152X}

\leavevmode\vadjust pre{\hypertarget{ref-R-ggstance}{}}%
Henry, L., Wickham, H., \& Chang, W. (2020). \emph{Ggstance: Horizontal 'ggplot2' components}. Retrieved from \url{https://CRAN.R-project.org/package=ggstance}

\leavevmode\vadjust pre{\hypertarget{ref-hill_optimizing_2020}{}}%
Hill, R. J. W., Innominato, P. F., Lévi, F., \& Ballesta, A. (2020). Optimizing circadian drug infusion schedules towards personalized cancer chronotherapy. \emph{PLOS Computational Biology}, \emph{16}(1), e1007218. \url{https://doi.org/10.1371/journal.pcbi.1007218}

\leavevmode\vadjust pre{\hypertarget{ref-horne_individual_1977}{}}%
Horne, J. A., \& Östberg, O. (1977). Individual differences in human circadian rhythms. \emph{Biological Psychology}, \emph{5}(3), 179--190. \url{https://doi.org/10.1016/0301-0511(77)90001-1}

\leavevmode\vadjust pre{\hypertarget{ref-R-ggformula}{}}%
Kaplan, D., \& Pruim, R. (2021). \emph{Ggformula: Formula interface to the grammar of graphics}. Retrieved from \url{https://CRAN.R-project.org/package=ggformula}

\leavevmode\vadjust pre{\hypertarget{ref-kerkhof_inter-individual_1985}{}}%
Kerkhof, G. A. (1985). Inter-individual differences in the human circadian system: {A} review. \emph{Biological Psychology}, \emph{20}(2), 83--112. \url{https://doi.org/10.1016/0301-0511(85)90019-5}

\leavevmode\vadjust pre{\hypertarget{ref-li_sleep_2018}{}}%
Li, J., Vitiello, M. V., \& Gooneratne, N. S. (2018). Sleep in {Normal} {Aging}. \emph{Sleep Medicine Clinics}, \emph{13}(1), 1--11. \url{https://doi.org/10.1016/j.jsmc.2017.09.001}

\leavevmode\vadjust pre{\hypertarget{ref-mander_sleep_2017}{}}%
Mander, B. A., Winer, J. R., \& Walker, M. P. (2017). Sleep and {Human} {Aging}. \emph{Neuron}, \emph{94}(1), 19--36. \url{https://doi.org/10.1016/j.neuron.2017.02.004}

\leavevmode\vadjust pre{\hypertarget{ref-mong_sleep_2011}{}}%
Mong, J. A., Baker, F. C., Mahoney, M. M., Paul, K. N., Schwartz, M. D., Semba, K., \& Silver, R. (2011). Sleep, rhythms, and the endocrine brain: Influence of sex and gonadal hormones. \emph{The Journal of Neuroscience: The Official Journal of the Society for Neuroscience}, \emph{31}(45), 16107--16116. \url{https://doi.org/10.1523/JNEUROSCI.4175-11.2011}

\leavevmode\vadjust pre{\hypertarget{ref-R-BayesFactor}{}}%
Morey, R. D., \& Rouder, J. N. (2018). \emph{BayesFactor: Computation of bayes factors for common designs}. Retrieved from \url{https://CRAN.R-project.org/package=BayesFactor}

\leavevmode\vadjust pre{\hypertarget{ref-muthukrishna_beyond_2020}{}}%
Muthukrishna, M., Bell, A. V., Henrich, J., Curtin, C. M., Gedranovich, A., McInerney, J., \& Thue, B. (2020). Beyond {Western}, {Educated}, {Industrial}, {Rich}, and {Democratic} ({WEIRD}) {Psychology}: {Measuring} and {Mapping} {Scales} of {Cultural} and {Psychological} {Distance}. \emph{Psychological Science}, \emph{31}(6), 678--701. \url{https://doi.org/10.1177/0956797620916782}

\leavevmode\vadjust pre{\hypertarget{ref-R-tibble}{}}%
Müller, K., \& Wickham, H. (2021). \emph{Tibble: Simple data frames}. Retrieved from \url{https://CRAN.R-project.org/package=tibble}

\leavevmode\vadjust pre{\hypertarget{ref-prisma-p_group_preferred_2015}{}}%
PRISMA-P Group, Moher, D., Shamseer, L., Clarke, M., Ghersi, D., Liberati, A., \ldots{} Stewart, L. A. (2015). Preferred reporting items for systematic review and meta-analysis protocols ({PRISMA}-{P}) 2015 statement. \emph{Systematic Reviews}, \emph{4}(1), 1. \url{https://doi.org/10.1186/2046-4053-4-1}

\leavevmode\vadjust pre{\hypertarget{ref-R-mosaicData}{}}%
Pruim, R., Kaplan, D., \& Horton, N. (2021). \emph{mosaicData: Project MOSAIC data sets}. Retrieved from \url{https://CRAN.R-project.org/package=mosaicData}

\leavevmode\vadjust pre{\hypertarget{ref-R-mosaic}{}}%
Pruim, R., Kaplan, D. T., \& Horton, N. J. (2017). The mosaic package: Helping students to 'think with data' using r. \emph{The R Journal}, \emph{9}(1), 77--102. Retrieved from \url{https://journal.r-project.org/archive/2017/RJ-2017-024/index.html}

\leavevmode\vadjust pre{\hypertarget{ref-R-base}{}}%
R Core Team. (2021). \emph{R: A language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing. Retrieved from \url{https://www.R-project.org/}

\leavevmode\vadjust pre{\hypertarget{ref-redline_effects_2004}{}}%
Redline, S., Kirchner, H. L., Quan, S. F., Gottlieb, D. J., Kapur, V., \& Newman, A. (2004). The effects of age, sex, ethnicity, and sleep-disordered breathing on sleep architecture. \emph{Archives of Internal Medicine}, \emph{164}(4), 406--418. \url{https://doi.org/10.1001/archinte.164.4.406}

\leavevmode\vadjust pre{\hypertarget{ref-santhi_sex_2016}{}}%
Santhi, N., Lazar, A. S., McCabe, P. J., Lo, J. C., Groeger, J. A., \& Dijk, D.-J. (2016). Sex differences in the circadian regulation of sleep and waking cognition in humans. \emph{Proceedings of the National Academy of Sciences of the United States of America}, \emph{113}(19), E2730--2739. \url{https://doi.org/10.1073/pnas.1521637113}

\leavevmode\vadjust pre{\hypertarget{ref-santhi_spectral_2012}{}}%
Santhi, N., Thorne, H. C., Veen, D. R. van der, Johnsen, S., Mills, S. L., Hommes, V., \ldots{} Dijk, D.-J. (2012). The spectral composition of evening light and individual differences in the suppression of melatonin and delay of sleep in humans: {Artificial} evening light suppresses melatonin and delays sleep. \emph{Journal of Pineal Research}, \emph{53}(1), 47--59. \url{https://doi.org/10.1111/j.1600-079X.2011.00970.x}

\leavevmode\vadjust pre{\hypertarget{ref-R-lattice}{}}%
Sarkar, D. (2008). \emph{Lattice: Multivariate data visualization with r}. New York: Springer. Retrieved from \url{http://lmdvr.r-forge.r-project.org}

\leavevmode\vadjust pre{\hypertarget{ref-R-latticeExtra}{}}%
Sarkar, D., \& Andrews, F. (2019). \emph{latticeExtra: Extra graphical utilities based on lattice}. Retrieved from \url{https://CRAN.R-project.org/package=latticeExtra}

\leavevmode\vadjust pre{\hypertarget{ref-shamseer_preferred_2015}{}}%
Shamseer, L., Moher, D., Clarke, M., Ghersi, D., Liberati, A., Petticrew, M., \ldots{} the PRISMA-P Group. (2015). Preferred reporting items for systematic review and meta-analysis protocols ({PRISMA}-{P}) 2015: Elaboration and explanation. \emph{BMJ}, \emph{349}(jan02 1), g7647--g7647. \url{https://doi.org/10.1136/bmj.g7647}

\leavevmode\vadjust pre{\hypertarget{ref-spitschan_transparency_2020}{}}%
Spitschan, M., Schmidt, M. H., \& Blume, C. (2020). Transparency and open science principles in reporting guidelines in sleep research and chronobiology journals. \emph{Wellcome Open Research}, \emph{5}, 172. \url{https://doi.org/10.12688/wellcomeopenres.16111.1}

\leavevmode\vadjust pre{\hypertarget{ref-tankova_circadian_1994}{}}%
Tankova, I., Adan, A., \& Buela-Casal, G. (1994). Circadian typology and individual differences. {A} review. \emph{Personality and Individual Differences}, \emph{16}(5), 671--684. \url{https://doi.org/10.1016/0191-8869(94)90209-7}

\leavevmode\vadjust pre{\hypertarget{ref-R-corrplot2017}{}}%
Wei, T., \& Simko, V. (2017). \emph{R package "corrplot": Visualization of a correlation matrix}. Retrieved from \url{https://github.com/taiyun/corrplot}

\leavevmode\vadjust pre{\hypertarget{ref-R-reshape2}{}}%
Wickham, H. (2007). Reshaping data with the {reshape} package. \emph{Journal of Statistical Software}, \emph{21}(12), 1--20. Retrieved from \url{http://www.jstatsoft.org/v21/i12/}

\leavevmode\vadjust pre{\hypertarget{ref-R-ggplot2}{}}%
Wickham, H. (2016). \emph{ggplot2: Elegant graphics for data analysis}. Springer-Verlag New York. Retrieved from \url{https://ggplot2.tidyverse.org}

\leavevmode\vadjust pre{\hypertarget{ref-R-stringr}{}}%
Wickham, H. (2019). \emph{Stringr: Simple, consistent wrappers for common string operations}. Retrieved from \url{https://CRAN.R-project.org/package=stringr}

\leavevmode\vadjust pre{\hypertarget{ref-R-tidyr}{}}%
Wickham, H. (2021). \emph{Tidyr: Tidy messy data}. Retrieved from \url{https://CRAN.R-project.org/package=tidyr}

\leavevmode\vadjust pre{\hypertarget{ref-R-readxl}{}}%
Wickham, H., \& Bryan, J. (2019). \emph{Readxl: Read excel files}. Retrieved from \url{https://CRAN.R-project.org/package=readxl}

\leavevmode\vadjust pre{\hypertarget{ref-R-dplyr}{}}%
Wickham, H., François, R., Henry, L., \& Müller, K. (2021). \emph{Dplyr: A grammar of data manipulation}. Retrieved from \url{https://CRAN.R-project.org/package=dplyr}

\leavevmode\vadjust pre{\hypertarget{ref-R-readr}{}}%
Wickham, H., \& Hester, J. (2020). \emph{Readr: Read rectangular text data}. Retrieved from \url{https://CRAN.R-project.org/package=readr}

\leavevmode\vadjust pre{\hypertarget{ref-R-ggridges}{}}%
Wilke, C. O. (2021). \emph{Ggridges: Ridgeline plots in 'ggplot2'}. Retrieved from \url{https://CRAN.R-project.org/package=ggridges}

\leavevmode\vadjust pre{\hypertarget{ref-woitowich_10-year_2020}{}}%
Woitowich, N. C., Beery, A., \& Woodruff, T. (2020). A 10-year follow-up study of sex inclusion in the biological sciences. \emph{eLife}, \emph{9}, e56344. \url{https://doi.org/10.7554/eLife.56344}

\leavevmode\vadjust pre{\hypertarget{ref-R-ggsci}{}}%
Xiao, N. (2018). \emph{Ggsci: Scientific journal and sci-fi themed color palettes for 'ggplot2'}. Retrieved from \url{https://CRAN.R-project.org/package=ggsci}

\leavevmode\vadjust pre{\hypertarget{ref-R-knitr}{}}%
Xie, Y. (2015). \emph{Dynamic documents with {R} and knitr} (2nd ed.). Boca Raton, Florida: Chapman; Hall/CRC. Retrieved from \url{https://yihui.org/knitr/}

\end{CSLReferences}

\endgroup

\clearpage
\renewcommand{\listfigurename}{Figure captions}

\end{document}

### 2021-05-08 article.Rmd
Updates

@@
 ---
-title             : "Study sample characteristics in chronobiology and sleep research"
+title             : "Inclusion, reporting and analysis of demographic variables in chronobiology and sleep research"
 shorttitle        : "Study sample characteristics in chronobiology and sleep research"
@@
 The extent to which a scientific field's findings are generalisable depend very much on the representativeness of a given study sample.
-As an example, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020].
+A recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020].
 The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014].
+The term "gender data gap" has recently been introduced, demonstrating that women have historically been excluded from biomedical research [@criado-perez_invisible_2020].
+
+The lack of diversity in biomedical and clinical research and understudy of minorities in the presence of existing health inequities exacerbates these inequities [@oh_diversity_2015].
 
@@
 Here, we address the question of participant-level demographic characteristics (age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population) and reporting thereof in chronobiology and sleep research.
-Here, we extracted the study sample characteristics in a total of 1356 randomly sampled publications across the 8 top (ranked by the Journal Impact Factor) chronobiology and sleep research and subjected them to a comprehensive analysis.
+Here, we extracted the study sample characteristics in a total of 1355 randomly sampled publications across the 8 top (ranked by the Journal Impact Factor) chronobiology and sleep research and subjected them to a comprehensive analysis.
 
@@
 
-*Procedure*.
+### Procedure
+
 Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals were considered.
@@
 
-*Article inclusion*.
+### Article inclusion
+
 6,777 articles were identified through a MEDLINE search by the journal and including odd years.
@@
 
-*Review and article extraction*.
+### Review and article extraction
+
 All included articles were reviewed for eligibility and coded by RW.
@@
 
-*Pre-registration*.
+### Pre-registration
+
 We pre-registered our protocol (specified using the PRISMA-P template [@shamseer_preferred_2015; @prisma-p_group_preferred_2015]) on the Open Science Framework (https://osf.io/cu3we/).
 
-*Materials, data and code availability*.
+### Materials, data and code availability
+
 All data underlying this manuscript are available on a public GitHub repository (https://github.com/hcvnl/sleep_circadian_demographics_data).
-The article was written in R [@R-base] using RMarkdown and papaja [@R-papaja], employing a series of additional R packages [@R-corrplot2017; @R-dplyr; @R-ggExtra; @R-ggformula; @R-ggplot2; @R-ggridges; @R-ggsci; @R-ggstance; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lattice; @R-latticeExtra; @R-lemon; @R-Matrix; @R-mosaic; @R-mosaicData; @R-pracma; @R-readr; @R-readxl; @R-reshape2; @R-stringr; @R-tibble; @R-tidyr; @R-BayesFactor] and is fully reproducible.
+The article was written in R [@R-base] using RMarkdown and papaja [@R-papaja], employing a series of additional R packages [@R-corrplot2017; @R-dplyr; @R-ggExtra; @R-ggformula; @R-ggplot2; @R-ggridges; @R-ggsci; @R-ggstance; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lattice; @R-latticeExtra; @R-lemon; @R-Matrix; @R-mosaic; @R-mosaicData; @R-pracma; @R-readr; @R-readxl; @R-reshape2; @R-stringr; @R-tibble; @R-tidyr] and is fully reproducible.
 
@@
   ylab("Proportion of articles") +
+  ggtitle("Funding reporting") +
   labs(fill="") +
@@
   ylab("Proportion of articles") +
+  ggtitle("Geographical location of included articles") +
   labs(fill="Country") +
@@
   ylab("Sample size, log-10 scaling") +
+  ggtitle("Evolution of sample size") +
   annotation_logticks(sides="l") +
@@
   xlab("Age variable") +
+  ggtitle("Age reporting by year") +
   scale_y_continuous(breaks=c(1979,1989,1999,2009,2019), expand=c(0.02,0),
@@
   xlab("Age variable") +
+  ggtitle("Age reporting by journal") +
   scale_fill_viridis_c(end=0.6, direction=-1, name="Reporting",
@@
   xlab("Year of publication") +
+  ggtitle("Evolution of mean age by year") +
   scale_x_continuous(limits=c(1979,2019),breaks=c(1979,1989,1999,2009,2019)) +
@@
   ylab("Proportion of articles") +
+  ggtitle("Sex inclusion by year") +
   labs(fill="Sex") +
@@
   ylab("Proportion of articles") +
+  ggtitle("Reporting of other demographiic variables") +
   scale_color_viridis_d(name="Variable reported", 
@@
   ylab("Proportion of articles") +
+  ggtitle("Data analysis by demographic variables") +
   labs(fill="Variable reported") +
@@
 
-### Inclusion & Exclusion
-
-  `r round(percent.included)`% of the sampled articles met eligibility criteria during the review.
-
-```{r, fig.cap="Funding sources and geographical location of the studies.
-A.
-Reporting of funding across the years.
-B.
-Distribution of articles by study location.
-The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
+### Number of analysed articles
+
+Out of `r as.numeric(nrow(data.original))` articles, we included and extracted data from `r as.numeric(nrow(data.included))` (`r round(percent.included)`%).
+The distribution of years in which the articles were published is non-uniform and we included and extracted data from more recent articles (Fig.
+1).
+In addition, the representation of journals in the final list of articles is non-uniform, not least as the included journals will have not have been available from the entire data collection period (1979 and onwards).
+We also examined reasons for exclusion amongst the articles that we did not include and extract data from.
+These are given in Figure 2.
+
+```{r, fig.cap="Included and analysed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
 figure1
@@
 
-```{r, fig.cap="Funding sources and geographical location of the studies.
-A.
-Reporting of funding across the years.
-B.
-Distribution of articles by study location.
-The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
+```{r, fig.cap="Normalised distributions of included and excluded articles.", warning=FALSE, message=FALSE}
 figure2
@@
 
-  Funding sources were reported by `r round(percent.funding.name)`% of studies, while funding number was also reported in `r round(percent.funding.num)`% of these cases.
+  Funding sources were reported by `r round(percent.funding.name)`% of studies, while funding number was also reported in `r round(percent.funding.num)`% of these cases (Fig.
+3).
 Overall, funding by the United States' National Institutes of Health (NIH) represented `r round(percent.funding.NI)`% of the reported funding agencies.
@@
 
-```{r, fig.cap="Funding sources and geographical location of the studies.
-A.
-Reporting of funding across the years.
-B.
-Distribution of articles by study location.
-The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
+```{r, fig.cap="Reporting of funding across years.", warning=FALSE, message=FALSE}
 figure3
@@
 `r nrow(country.multiple)`% of articles reported multiple countries of study.
-Figure 2B shows the distribution of study location across time with the eight most represented countries.
-
-```{r, fig.cap="Funding sources and geographical location of the studies.
-A.
-Reporting of funding across the years.
-B.
-Distribution of articles by study location.
-The eight most represented countries are individually shown.", warning=FALSE, message=FALSE}
+Figure 4 shows the distribution of study location across time with the eight most represented countries.
+
+```{r, fig.cap="Gegraphical location of the studies. The eight most represented countries across the entire dataset are individually shown.", warning=FALSE, message=FALSE}
 figure4
@@
 
-```{r, fig.cap="Sample size of the recruited volunteers. Numbers are computed with a log-10 scale.", warning=FALSE, message=FALSE}
+```{r, fig.cap="Sample size of the recruited volunteers as a function of publication year. Numbers are computed on a log-10 scale.", warning=FALSE, message=FALSE}
 figure5
@@
   `r round(percent.age)`% of articles reported a variable describing age.
-Figure 4A and 4B show the use of various age variables by year and journal of publication.
-Overall, the mean age of the study populations was `r round(mean(data.age.mean$age_reported_mean_value))` years old.
-
-```{r, fig.cap="To be completed.", warning=FALSE, message=FALSE}
-figure6
-```
-```{r, fig.cap="To be completed.", warning=FALSE, message=FALSE}
-figure7
-```
-
-```{r, fig.cap="Reporting of age variables.
-Correlation between the reporting of the mean, standrad deviation of the mean, median, minimum, maximum, interquartile range and other age variables by A.
-year of publication and B.
-journal of publication.
-C.
-Mean age by year of publication.", message=FALSE, warning=FALSE}
+Overall, the average mean age of the study populations was `r round(mean(data.age.mean$age_reported_mean_value))` years old.
+We examined the extent to which the mean age across studies differed widely as a function of publication year (Fig.
+6), and found that the mean age is much more widely varied in later years, likely reflecting the extent of considering study samples that are more varied in age.
+
+```{r, fig.cap="Evolution of mean age in included studies as a function of publication year. Fit shown is a linear fit.", warning=FALSE, message=FALSE}
 figure8
@@
   Sex was reported in `r round(percent.sex)`% of the studies.
-Figure 5 displays the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants.
+In Figure 7, we show the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants.
 `r round(percent.sex.male)`% of the studies reporting sex only recruited male participants, while `r round(percent.sex.female)`% only employed females.
@@
 
-```{r, fig.cap="Sex inclusion. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
+```{r, fig.cap="Sex inclusion by year. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
 figure9
@@
 
+  We also examined the reporting of other demographic variables, including ethnicity, education, profession and socio-economic status.
 Other demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status.
-Figure 6 shows the distribution of this reporting across the years.
-Figures 7 show the number of categories reported for each variable.
+Figure 8 shows the distribution of this reporting across the years.
+Qualitatively, we see an increase of reporting additional demographic variables over time.
+In Figure 9, we show the number of categories reported for each variable amongst those articles that included it in a histogram.
 
@@
 ```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
-grid.arrange(figure11A, figure11B, figure11C, figure11D, nrow=2)
+grid.arrange(figure11A, figure11B, figure11C, figure11D, nrow=2, top="Number of reported categories")
 ```
@@
 
-  A
+  We also examined the extent to which articles reported subgroup analyses of the data based on one or more of the reported demographic variables.
+Over time (Fig.
+10), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed.
+The most common subgroup analyses involve disaggregting by sex and age, or both.
 
@@
 
-In the clinical domain, the need to time therapy based on a patient's individual circadian rhythm has more recently become the focus of the emerging field of chronotherapy or chronotherapeutics [@adam_core_2019; @greco_personalized_2020; @dijk_novel_2020; @hill_optimizing_2020].
+## Taking an inventory of represented study samples reveals the representativeness of our collective knowledge
 
@@
 The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020].
+In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at these question of participant demographics in chronobiology and sleep research.
+
+##
+
+The ability to generalise findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories.
+The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020].
+
+## The need to consider individual differences
+
+In the clinical domain, the need to time therapy based on a patient's individual circadian rhythm has more recently become the focus of the emerging field of chronotherapy or chronotherapeutics [@adam_core_2019; @greco_personalized_2020; @dijk_novel_2020; @hill_optimizing_2020].
+
+## Limitations of the current review
+
+We now turn to possible limitations of this review and the included analyses and discuss how they might introduce bias in our findings.
+First, we consider the possibility that the article selection procedure may have introduced biases.
+Our review only concerned articles from a subset of eight specialized journals.
+As a consequence, the included articles were necessarily published in these journals, ignoring relevant articles published in other specialised journals (such as those included in the list of candidate journals), and articles published in other, including interdisciplinary journals.
+This raises the question to what extent we may have missed a section of the literature that would have been relevant to include here.
+As an alternative strategy, we considered randomly sampling a subset of chronobiology and sleep research articles produced by a general search (e.g.
+on search from "sleep OR chronobiology" on MEDLINE), but considered this to be too permissive.
+Our strategy of selecting a subset of candidate journals provided a reasonable trade-off, as well as sampling from a range of field-specific outlets.
+
+Due to the non-uniform distribution of publication years of the included articles (Fig.
+1), variables derived from published papers and visualized and/or analyzed by year will have varying uncertainty, with reported percentages from publications of the earlier years being most uncertain.
+The fact that early years are represented with fewer articles, however, is a not a function of our data set, but of the exponential groth of scientific output [@powell_introduction_2017; @bornmann_growth_2015; @parolo_attention_2015].
+
+## Towards standardised reporting of demographic variables: From checklists to schemas?
+
+There are guidelines and/or checklists for standardizing reporting of participant characteristics, such as CONSORT [@schulz_consort_2010] or STROBE [@von_elm_strengthening_2007] (an extensive data base for health research reporting guidelines is provided by the Equator Network, https://www.equator-network.org/).
+Some biomedical journals [e.g.
+@robinson_transparent_2017] specifically state demographic reporting requirements in the author instructions.
+Similarly, some organisations may make recommendations of specific reporting items for specific types of study [@veitch_cie_2020].
+
+These guidelines and/or checklists are largely focused on *what* is reported and not *how* it is reported.
+There is, *a priori*, however, no reason to not develop and use a standardized and machine-readable schema for reporting participant characteristics.
+The FAIR principles state that data should be findable, accessible, interoperable and reusable [@wilkinson_fair_2016], and one way of realising these criteria is the use of data schemas which could prescribe categories of data and common naming schemes for reporting participant characteristics.
+Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extend such any data schema may be exclusionary (e.g.
+by enforcing gender binaries), and whether any specific demographic variable is truly important (following the principle of data minimization).
 

### 2021-05-13 article.Rmd
Updates

@@
 
-  Many aspects of sleep and circadian physiology appear to be sensitive to characteristics of the studied population, most notably sex [@redline_effects_2004; @santhi_sex_2016; @cain_sex_2010; @anderson_sexual_2020; @mong_sleep_2011], age [@redline_effects_2004; @espiritu_aging-related_2008; @mander_sleep_2017; @desforges_sleep_1990; @li_sleep_2018; @bliwise_sleep_1993; @benloucif_responsiveness_2006; @duffy_aging_2015] and ethnicity [@ahn_scoping_2021; @goldstein_sleep_2020; @eastman_blacks_2012; @eastman_circadian_2016].
-More generally, there is a large literature on individual differences on sleep and circadian physiology [@van_dongen_individual_2005; @kerkhof_inter-individual_1985; @tankova_circadian_1994; @horne_individual_1977; @baehr_individual_2000; @burgess_individual_2008; @santhi_spectral_2012; @chellappa_individual_2021], demonstrating the need to consider participant-level data.
-
-The extent to which a scientific field's findings are generalisable depend very much on the representativeness of a given study sample.
-A recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020].
+There are large individual differences in and circadian physiology [@van_dongen_individual_2005; @kerkhof_inter-individual_1985; @tankova_circadian_1994; @horne_individual_1977; @baehr_individual_2000; @burgess_individual_2008; @santhi_spectral_2012; @chellappa_individual_2021], demonstrating the need to consider sleep and circadian physiology at participant-level.
+Some of these characteristics are systematically related to demographic variables, most notably sex [@redline_effects_2004; @santhi_sex_2016; @cain_sex_2010; @anderson_sexual_2020; @mong_sleep_2011], age [@redline_effects_2004; @espiritu_aging-related_2008; @mander_sleep_2017; @desforges_sleep_1990; @li_sleep_2018; @bliwise_sleep_1993; @benloucif_responsiveness_2006; @duffy_aging_2015] and ethnicity [@ahn_scoping_2021; @goldstein_sleep_2020; @eastman_blacks_2012; @eastman_circadian_2016].
+Zooming out from the level of individual studies, there are significant inequities in sleep health depending on ethnicity and socio-economic status [@jackson_sleep_2020; @jackson_racialethnic_2020; @williams_racialethnic_2015; @jackson_workshop_2020; @jean-louis_importance_2016; @laposky_reducing_2016].
+As compromised sleep has many knock-on effects, including negative effects on cardiovascular, metabolic, neurobehavioural and cognitive function, it is a key imperative to understand how demographic variables influence sleep and circadian rhythms.
+
+The extent to which a scientific field's findings are generalisable are a function of the representativeness of a given study sample, and the extent to which findings differ between, e.g., different sexes and ethnicity In the domain of sex, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020].
 The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014].
 The term "gender data gap" has recently been introduced, demonstrating that women have historically been excluded from biomedical research [@criado-perez_invisible_2020].
-
-The lack of diversity in biomedical and clinical research and understudy of minorities in the presence of existing health inequities exacerbates these inequities [@oh_diversity_2015].
+Similarly, the lack of diversity in biomedical and clinical research, and understudy of minorities in the presence of existing health inequities exacerbates these inequities [@oh_diversity_2015].
 
@@
 Here, we address the question of participant-level demographic characteristics (age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population) and reporting thereof in chronobiology and sleep research.
-Here, we extracted the study sample characteristics in a total of 1355 randomly sampled publications across the 8 top (ranked by the Journal Impact Factor) chronobiology and sleep research and subjected them to a comprehensive analysis.
+We extracted the study sample characteristics in a total of 1355 randomly sampled publications across the 8 top (ranked by the Journal Impact Factor) chronobiology and sleep research and subjected them to a comprehensive analysis in terms of the inclusion, reporting and analysis of demographic variables.
 
@@
 Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals were considered.
-The listing of possible of target journals was based on a previously established list of journals implementing a hybrid strategy by consulting the Web of Science Master Journal List, domain-relevant expertise in sleep and chronobiology and consulting with a senior researcher with >25 years of experience in the field [@spitschan_transparency_2020].
-From this previously derived list, we selected eight journals were selected based on their five-year Impact Factor, and included *Journal of Pineal Research* (ISSN: 0742-3098 / 1600-079X; 2018 5-year IF: 12.197), *Sleep*  (0161-8105 / 1550-9109; 5.588), *Journal of Sleep Research* (0962-1105 / 1365-2869; 3.951), *Sleep Medicine* (1389-9457 / 1878-5506; 3.934), *Journal of Clinical Sleep Medicine* (1550-9389 / 1550-9397; 3.855), *Journal of Biological Rhythms* (0748-7304 / 1552-4531; 3.349), Behavioral Sleep Medicine (1540-2002 / 1540-2010; 3.162), and *Chronobiology International* (0742-0528 / 1525-6073; 2.998).
+The list of possible target journals was based on a previously established list of journals implementing a hybrid strategy by consulting the Web of Science Master Journal List, domain-relevant expertise in sleep and chronobiology and consulting with a senior researcher with >25 years of experience in the field [@spitschan_transparency_2020].
+From this previously derived list, we selected eight journals based on their five-year Impact Factor, and included *Journal of Pineal Research* (ISSN: 0742-3098 / 1600-079X; 2018 5-year IF: 12.197), *Sleep*  (0161-8105 / 1550-9109; 5.588), *Journal of Sleep Research* (0962-1105 / 1365-2869; 3.951), *Sleep Medicine* (1389-9457 / 1878-5506; 3.934), *Journal of Clinical Sleep Medicine* (1550-9389 / 1550-9397; 3.855), *Journal of Biological Rhythms* (0748-7304 / 1552-4531; 3.349), Behavioral Sleep Medicine (1540-2002 / 1540-2010; 3.162), and *Chronobiology International* (0742-0528 / 1525-6073; 2.998).
 While *Sleep Medicine Reviews* also features in the list of journals, we did not include it as it primarily publishes reviews.
@@
 All included articles were reviewed for eligibility and coded by RW.
-The reporting of sample size, age, sex, gender, ethnicity, level of education, socioeconomic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded.
+The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded.
 Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.
+Data were coded in an Excel Spreadsheet and analyzed in R Studio (version 4.0.5).
+Reporting of funding, geographical location, and number of sub-studies for each article were investigated for the sample of articles that passed all eligibility criteria.
 
@@
 
-  Data were coded in an Excel Spreadsheet and analyzed in R Studio (version 4.0.5).
-Reporting of funding, geographical location, and number of sub-studies for each article were investigated for the sample of articles that passed all eligibility criteria.
-
 ### Pre-registration
@@
 data.sex <- data.sex %>%
-  mutate(report=if_else(male == "Yes" & female == "Yes", "Both Sexes", 
+  mutate(report=if_else(male == "Yes" & female == "Yes", "Both sexes", 
                           if_else(male == "Yes" & female == "No", "Male",
@@
                                             "Female",
-                                            "Both Sexes",
+                                            "Both sexes",
                                             "Male")))) +
@@
   ylab("Proportion of articles") +
-  ggtitle("Reporting of other demographiic variables") +
+  ggtitle("Reporting of other demographic variables") +
   scale_color_viridis_d(name="Variable reported", 
@@
 
-Out of `r as.numeric(nrow(data.original))` articles, we included and extracted data from `r as.numeric(nrow(data.included))` (`r round(percent.included)`%).
+Out of `r as.numeric(nrow(data.original))` identified and pre-screened articles, we included and extracted data from `r as.numeric(nrow(data.included))` (`r round(percent.included)`%).
 The distribution of years in which the articles were published is non-uniform and we included and extracted data from more recent articles (Fig.
@@
 In addition, the representation of journals in the final list of articles is non-uniform, not least as the included journals will have not have been available from the entire data collection period (1979 and onwards).
+
+```{r, fig.cap="Included and analysed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
+figure1
+```
+
 We also examined reasons for exclusion amongst the articles that we did not include and extract data from.
 These are given in Figure 2.
-
-```{r, fig.cap="Included and analysed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
-figure1
-```
+Reasons for exclusion vary somewhat between different years, with articles not reporting original research being excluded at the highest rate on average.
 
@@
 
+  We examined the reporting of funding sources in included articles.
 Funding sources were reported by `r round(percent.funding.name)`% of studies, while funding number was also reported in `r round(percent.funding.num)`% of these cases (Fig.
@@
 Overall, `r nrow(country.single)` countries were represented.
-`r nrow(country.multiple)`% of articles reported multiple countries of study.
+`r round(percent.country.multiple)`% of articles reported multiple countries of study.
 Figure 4 shows the distribution of study location across time with the eight most represented countries.
+Across all years, the US is the most represented country.
 
@@
 
-  Sample size was reported in `r round(percent.ss)`% of studies, while `r round(percent.study.single)`% investigated a single sampled population.
+  We examined whether sample size reported in studies, and the sample sizes reported.
+Overall, sample size was reported in `r round(percent.ss)`% of studies.
+We furthermore examined the distribution of samples sizes as a function of the publication year of the article (Fig.
+5), showing a wider distribution of sample sizes in more recent articles.
 
@@
 
-```{r, fig.cap="Evolution of mean age in included studies as a function of publication year. Fit shown is a linear fit.", warning=FALSE, message=FALSE}
+```{r, fig.cap="Evolution of mean age in included studies as a function of publication year. Fit shown is a linear fit (±95%CI).", warning=FALSE, message=FALSE}
 figure8
@@
 
-  We also examined the reporting of other demographic variables, including ethnicity, education, profession and socio-economic status.
-Other demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status.
+  We examined the reporting of the demographic variables ethnicity, education, profession and socio-economic status.
+These demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status.
 Figure 8 shows the distribution of this reporting across the years.
-Qualitatively, we see an increase of reporting additional demographic variables over time.
-In Figure 9, we show the number of categories reported for each variable amongst those articles that included it in a histogram.
-
-```{r, fig.cap="Reporting of education, ethinicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
+Qualitatively, there is a clear increase of reporting additional demographic variables over time, with ethnicity being the most reported demographic variables across articles in more recent years.
+
+```{r, fig.cap="Reporting of education, ethnicity, profession and socio-economic status (SES).", warning=FALSE, message=FALSE}
 figure10
 ```
-
-```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status.", warning=FALSE, message=FALSE}
+  We furthermore examined the number of categories included for each of these demographic variables.
+In Figure 9, we show the number of categories reported for each variable among those articles that included it in a histogram.
+On overage, we find that the coding scheme for each of these variables in the included articles comprised between 3 and 4 variables.
+
+```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status (SES). Yellow vertical line corresponds to the median.", warning=FALSE, message=FALSE}
 grid.arrange(figure11A, figure11B, figure11C, figure11D, nrow=2, top="Number of reported categories")
@@
 
-  `r round(percent.focus.sex)`% of articles focused on a sex dependent feature, while `r round(percent.focus.clinical)`% investigated a clinical feature.
+  We considered whether a given article reported on a specific, pre-defined group of people.
+We find `r round(percent.focus.sex)`% of articles focused on a sex dependent feature, while `r round(percent.focus.clinical)`% investigated a clinical feature.
 `r round(percent.focus.twins)`% of studies focused on twins, `r round(percent.focus.pregnant)`% on pregnant women, `r round(percent.focus.shift)`% on shift workers and `r round(percent.focus.students)`% on university students.
@@
 
-  We also examined the extent to which articles reported subgroup analyses of the data based on one or more of the reported demographic variables.
+  We examined the extent to which articles reported subgroup analyses of the data based on one or more of the reported demographic variables.
 Over time (Fig.
-10), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed.
+9), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed.
 The most common subgroup analyses involve disaggregting by sex and age, or both.
@@
 
+## Summary of main findings
+
+Here, we considered the inclusion, reporting of `r as.numeric(nrow(data.included))` articles in chronobiology and sleep research, sampled from eight journals.
+More articles from more recent years are included, with articles originating in the US representing the large group.
+Over time, we found an increase in the range of sample sizes.
+We furthermore find a wider spread of average ages included in articles.
+Across the included articles, we observe an increase in reporting of demographic variables (ethnicity, education, professional and socioeconomic status) over time, with ethnicity being the most-reported variable in recent years.
+Of studies included these demographic variables, the number of categories varies widely between studies.
+
 ## Taking an inventory of represented study samples reveals the representativeness of our collective knowledge
@@
 The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020].
-In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at these question of participant demographics in chronobiology and sleep research.
-
-##
-
-The ability to generalise findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories.
-The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020].
+In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at the inclusion, reporting and analysis participant demographics in chronobiology and sleep research.
 
@@
 
+There is convergent knowledge that sleep and circadian physiology and health are subject to large individual differences.
+Demographic variables provide on lens through which to understand individual differences, and importantly can provide insights into systemic disadvantages and inequities.
 In the clinical domain, the need to time therapy based on a patient's individual circadian rhythm has more recently become the focus of the emerging field of chronotherapy or chronotherapeutics [@adam_core_2019; @greco_personalized_2020; @dijk_novel_2020; @hill_optimizing_2020].
+Understanding interindividual variability needs to become a key research area to understand the circadian and sleep physiology in the face of human diversity.
 
@@
 
-We now turn to possible limitations of this review and the included analyses and discuss how they might introduce bias in our findings.
+We turn to possible limitations of this review and the included analyses and discuss how they might introduce bias in our findings.
 First, we consider the possibility that the article selection procedure may have introduced biases.
@@
 1), variables derived from published papers and visualized and/or analyzed by year will have varying uncertainty, with reported percentages from publications of the earlier years being most uncertain.
-The fact that early years are represented with fewer articles, however, is a not a function of our data set, but of the exponential groth of scientific output [@powell_introduction_2017; @bornmann_growth_2015; @parolo_attention_2015].
+The fact that early years are represented with fewer articles, however, is a not a function of our data set, but of the exponential growth of scientific output [@powell_introduction_2017; @bornmann_growth_2015; @parolo_attention_2015].
 
@@
 @robinson_transparent_2017] specifically state demographic reporting requirements in the author instructions.
-Similarly, some organisations may make recommendations of specific reporting items for specific types of study [@veitch_cie_2020].
+Similarly, some organisations may make recommendations of specific reporting items for specific types of study questions [@veitch_cie_2020].
 
@@
 The FAIR principles state that data should be findable, accessible, interoperable and reusable [@wilkinson_fair_2016], and one way of realising these criteria is the use of data schemas which could prescribe categories of data and common naming schemes for reporting participant characteristics.
-Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extend such any data schema may be exclusionary (e.g.
-by enforcing gender binaries), and whether any specific demographic variable is truly important (following the principle of data minimization).
+Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extend such any data schema may be exclusionary, e.g., by enforcing gender binaries [@hyde_future_2019], and whether any specific demographic variable is truly important (following the principle of data minimization).
+A further central point to consider is the extent to which disaggregation by demographic variables could be used to do harm in some way [@dignazio_data_2020].
 
 # Conclusion
+
+This review provided a first look at the inclusion, reporting and analysis of demographic variables in the chronobiology and sleep research literature, considering >1000 articles across eight journals.
+We addressed the need to consider individual differences, as well as the dependence of sleep and circadian rhythms ond demographic variables.
+Additionally, we we outline an opportunity to improve reporting of participant-level characteristics using formalised data schemas.
 

### 2021-05-19 article.Rmd
Updates

@@
     email         : "manuel.spitschan@psy.ox.ac.uk"
+    address       : Department of Experimental Psychology, University of Oxford, Anna Watts Building, Radcliffe Observatory Quarter, Woodstock Rd, Oxford, OX2 6GG, United Kingdom
     role:
@@
 classoption       : "jou"
-output            : papaja::apa6_pdf
+output            : papaja::apa6_docx
 ---

### 2021-05-20 article.Rmd
updates

--- 2021-05-13 article.Rmd
@@
     email         : "manuel.spitschan@psy.ox.ac.uk"
+    address       : Department of Experimental Psychology, University of Oxford, Anna Watts Building, Radcliffe Observatory Quarter, Woodstock Rd, Oxford, OX2 6GG, United Kingdom
+    telephone     : +44 1865 281861
     role:
@@
 authornote: |
-  Corresponding author: Manuel Spitschan (manuel.spitschan@psy.ox.ac.uk).
-This work ws funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
+ This work was funded by the Wellcome Trust (Sir Henry Wellcome Fellowship to MS 204686/Z/16/Z; Research Enrichment – Diversity & Inclusion WT 204686/Z/16/A) and Linacre College (Biomedical Sciences Junior Research Fellowship to MS).
 abstract: |
@@
 keywords          : "demographics, ethnicity, sex, research participants, reporting, publishing, meta-science"
-wordcount         : "X"
+wordcount         : "3200"
 
 bibliography      : references.bib
+csl               : vancouver.csl
 
@@
 
+# Summary
+
+In this article, we addressed the question how representiative, diverse and inclusive published research articles in sleep and chronobiology are.
+We analysed a sample of >1,300 articles published in sleep research and chronobiology journals between 1979 and 2019 for its inclusion, reporting analysis of demographic variables of study participants, including age, sex, gender, ethnicity, level of education, socio-economic status, and professio.
+We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases, with the frequency of reporting changing over time.
+We identify opportunities for improving the reporting of demographic variables.
+
+# Research Agenda
+
+Future research needs to identify:
+1. Schemas for reporting demographic variables in a harmonised way across geographical and culutral contexts;
+2. Gaps in the sleep and chronobiology literature with respect to understudied populations;
+3. Understand the extent to which research practices run counter the inclusion of diverse populations in research in all stages of the research cycle, and how this can exacerbate health inequities.
+
+# Practice Points
+
+1. Published studies on circadian and sleep physiology should be carefully examined.
+1. Reporting of demographic variables should be done carefully and systematically.
+2. Inclusion and diversity of different populations across the field needs to be ensured.
+
 # Introduction
@@
 
-### Procedure
+## Procedure
 
@@
 
-### Article inclusion
+## Article inclusion
 
@@
 
-### Review and article extraction
+## Review and article extraction
 
@@
 
-### Pre-registration
+## Pre-registration
 
@@
 
-### Materials, data and code availability
+## Materials, data and code availability
 
@@
 
-### The rest  of the analysis will be conducted on included single studies
+## The rest  of the analysis will be conducted on included single studies
 # Filter single studies
@@
 
-### Number of analysed articles
+## Number of analysed articles
 
@@
 
-### Funding
+## Funding
 
@@
 
-### Geographical location
+## Geographical location
 
@@
 
-### Sample size
+## Sample size
 
@@
 
-### Age
+## Age
 
@@
 
-### Sex
+## Sex
 
@@
 
-###  Ethnicity, education, profession and socio-economic status
+##  Ethnicity, education, profession and socio-economic status
 
@@
 
-```{r, fig.cap="Reporting of education, ethnicity, profession and socio-economic status (SES).", warning=FALSE, message=FALSE}
+```{r, fig.cap="test", warning=FALSE, message=FALSE}
 figure10
@@
 
-### Study focus
+## Study focus
 
@@
 
-### Analysis disaggregation
+## Analysis disaggregation
 

### 2021-06-14 article.Rmd
Update article.Rmd
Edit typos.

@@
 The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored, and any reported aggregate summary statistics for these variables were recorded.
-We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases.
+We found that while >90% of studies reported age or sex, all other variables were reported in <15% of cases.
 Sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s.
@@
 
-In this article, we addressed the question how representiative, diverse and inclusive published research articles in sleep and chronobiology are.
-We analysed a sample of >1,300 articles published in sleep research and chronobiology journals between 1979 and 2019 for its inclusion, reporting analysis of demographic variables of study participants, including age, sex, gender, ethnicity, level of education, socio-economic status, and professio.
+In this article, we addressed the question of how representative, diverse and inclusive published research articles in sleep and chronobiology are.
+We analyzed a sample of >1,300 articles published in sleep research and chronobiology journals between 1979 and 2019 for its inclusion, reporting analysis of demographic variables of study participants, including age, sex, gender, ethnicity, level of education, socio-economic status, and profession.
 We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases, with the frequency of reporting changing over time.
@@
 Future research needs to identify:
-1. Schemas for reporting demographic variables in a harmonised way across geographical and culutral contexts;
+1. Schemas for reporting demographic variables in a harmonized way across geographical and cultural contexts;
 2. Gaps in the sleep and chronobiology literature with respect to understudied populations;
@@
 
-There are large individual differences in and circadian physiology [@van_dongen_individual_2005; @kerkhof_inter-individual_1985; @tankova_circadian_1994; @horne_individual_1977; @baehr_individual_2000; @burgess_individual_2008; @santhi_spectral_2012; @chellappa_individual_2021], demonstrating the need to consider sleep and circadian physiology at participant-level.
+There are large individual differences in sleep and circadian physiology [@van_dongen_individual_2005; @kerkhof_inter-individual_1985; @tankova_circadian_1994; @horne_individual_1977; @baehr_individual_2000; @burgess_individual_2008; @santhi_spectral_2012; @chellappa_individual_2021], demonstrating the need to consider sleep and circadian physiology at a participant-level.
 Some of these characteristics are systematically related to demographic variables, most notably sex [@redline_effects_2004; @santhi_sex_2016; @cain_sex_2010; @anderson_sexual_2020; @mong_sleep_2011], age [@redline_effects_2004; @espiritu_aging-related_2008; @mander_sleep_2017; @desforges_sleep_1990; @li_sleep_2018; @bliwise_sleep_1993; @benloucif_responsiveness_2006; @duffy_aging_2015] and ethnicity [@ahn_scoping_2021; @goldstein_sleep_2020; @eastman_blacks_2012; @eastman_circadian_2016].
 Zooming out from the level of individual studies, there are significant inequities in sleep health depending on ethnicity and socio-economic status [@jackson_sleep_2020; @jackson_racialethnic_2020; @williams_racialethnic_2015; @jackson_workshop_2020; @jean-louis_importance_2016; @laposky_reducing_2016].
-As compromised sleep has many knock-on effects, including negative effects on cardiovascular, metabolic, neurobehavioural and cognitive function, it is a key imperative to understand how demographic variables influence sleep and circadian rhythms.
-
-The extent to which a scientific field's findings are generalisable are a function of the representativeness of a given study sample, and the extent to which findings differ between, e.g., different sexes and ethnicity In the domain of sex, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020].
+As compromised sleep has many knock-on effects, including negative effects on cardiovascular, metabolic, neurobehavioral and cognitive function, it is a key imperative to understand how demographic variables influence sleep and circadian rhythms.
+
+The extent to which a scientific field's findings are generalizable is a function of the representativeness of a given study sample, and the degree to which findings differ between, e.g., different sexes and ethnicity.
+In the domain of sex, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020].
 The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014].
@@
 
-6,777 articles were identified through a MEDLINE search by the journal and including odd years.
+6,777 articles were identified through a MEDLINE search and filtering by journal and odd years.
 A random sample of 20% was initially selected for screening.
@@
 
-## Number of analysed articles
+## Number of analyzed articles
 
@@
 1).
-In addition, the representation of journals in the final list of articles is non-uniform, not least as the included journals will have not have been available from the entire data collection period (1979 and onwards).
-
-```{r, fig.cap="Included and analysed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
+In addition, the representation of journals in the final list of articles is non-uniform, not least as the included journals will not have been available for the entire data collection period (1979 and onwards).
+
+```{r, fig.cap="Included and analyzed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
 figure1
@@
 
-We also examined reasons for exclusion amongst the articles that we did not include and extract data from.
+We also examined reasons for exclusion among the articles that we did not include and extract data from.
 These are given in Figure 2.
@@
 
-  We examined whether sample size reported in studies, and the sample sizes reported.
+  We examined whether sample size was reported in studies, and the sample sizes reported.
 Overall, sample size was reported in `r round(percent.ss)`% of studies.
@@
 Figure 8 shows the distribution of this reporting across the years.
-Qualitatively, there is a clear increase of reporting additional demographic variables over time, with ethnicity being the most reported demographic variables across articles in more recent years.
+Qualitatively, there is a clear increase in the reporting of additional demographic variables over time, with ethnicity being the most reported demographic variable across articles in more recent years.
 
@@
 In Figure 9, we show the number of categories reported for each variable among those articles that included it in a histogram.
-On overage, we find that the coding scheme for each of these variables in the included articles comprised between 3 and 4 variables.
+On overage, we find that the coding scheme for each of these variables in the included articles comprised between three and four variables.
 
@@
 9), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed.
-The most common subgroup analyses involve disaggregting by sex and age, or both.
+The most common subgroup analyses involve disaggregating by sex, age, or both.
 
@@
 Over time, we found an increase in the range of sample sizes.
-We furthermore find a wider spread of average ages included in articles.
-Across the included articles, we observe an increase in reporting of demographic variables (ethnicity, education, professional and socioeconomic status) over time, with ethnicity being the most-reported variable in recent years.
-Of studies included these demographic variables, the number of categories varies widely between studies.
+We furthermore found a wider spread of average ages included in articles.
+Across the included articles, we observed an increase in the reporting of demographic variables (ethnicity, education, professional and socioeconomic status) over time, with ethnicity being the most-reported variable in recent years.
+Of studies included these demographic variables, the number of categories varied widely between studies.
 
@@
 
-The ability to generalise findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories.
+The ability to generalize findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories.
 The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020].
-In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at the inclusion, reporting and analysis participant demographics in chronobiology and sleep research.
+In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at the inclusion, reporting and analysis of participant demographics in chronobiology and sleep research.
 
@@
 
-There is convergent knowledge that sleep and circadian physiology and health are subject to large individual differences.
+There is convergent knowledge that health and sleep and circadian physiology are subject to large individual differences.
 Demographic variables provide on lens through which to understand individual differences, and importantly can provide insights into systemic disadvantages and inequities.
@@
 Our review only concerned articles from a subset of eight specialized journals.
-As a consequence, the included articles were necessarily published in these journals, ignoring relevant articles published in other specialised journals (such as those included in the list of candidate journals), and articles published in other, including interdisciplinary journals.
+As a consequence, the included articles were necessarily published in these journals, ignoring relevant articles published in other specialized journals (such as those included in the list of candidate journals), and articles published in other, including interdisciplinary journals.
 This raises the question to what extent we may have missed a section of the literature that would have been relevant to include here.
@@
 1), variables derived from published papers and visualized and/or analyzed by year will have varying uncertainty, with reported percentages from publications of the earlier years being most uncertain.
-The fact that early years are represented with fewer articles, however, is a not a function of our data set, but of the exponential growth of scientific output [@powell_introduction_2017; @bornmann_growth_2015; @parolo_attention_2015].
-
-## Towards standardised reporting of demographic variables: From checklists to schemas?
-
-There are guidelines and/or checklists for standardizing reporting of participant characteristics, such as CONSORT [@schulz_consort_2010] or STROBE [@von_elm_strengthening_2007] (an extensive data base for health research reporting guidelines is provided by the Equator Network, https://www.equator-network.org/).
+The fact that early years are represented with fewer articles, however, is not a function of our data set, but of the exponential growth of scientific output [@powell_introduction_2017; @bornmann_growth_2015; @parolo_attention_2015].
+
+## Towards standardized reporting of demographic variables: From checklists to schemas?
+
+There are guidelines and/or checklists for standardizing reporting of participant characteristics, such as CONSORT [@schulz_consort_2010] or STROBE [@von_elm_strengthening_2007] (an extensive database for health research reporting guidelines is provided by the Equator Network, https://www.equator-network.org/).
 Some biomedical journals [e.g.
 @robinson_transparent_2017] specifically state demographic reporting requirements in the author instructions.
-Similarly, some organisations may make recommendations of specific reporting items for specific types of study questions [@veitch_cie_2020].
+Similarly, some organizations may make recommendations of specific reporting items for specific types of study questions [@veitch_cie_2020].
 
@@
 There is, *a priori*, however, no reason to not develop and use a standardized and machine-readable schema for reporting participant characteristics.
-The FAIR principles state that data should be findable, accessible, interoperable and reusable [@wilkinson_fair_2016], and one way of realising these criteria is the use of data schemas which could prescribe categories of data and common naming schemes for reporting participant characteristics.
-Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extend such any data schema may be exclusionary, e.g., by enforcing gender binaries [@hyde_future_2019], and whether any specific demographic variable is truly important (following the principle of data minimization).
+The FAIR principles state that data should be findable, accessible, interoperable and reusable [@wilkinson_fair_2016], and one way of realizing these criteria is the use of data schemas which could prescribe categories of data and common naming schemes for reporting participant characteristics.
+Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extent such data schema may be exclusionary, e.g., by enforcing gender binaries [@hyde_future_2019], and whether any specific demographic variable is truly important (following the principle of data minimization).
 A further central point to consider is the extent to which disaggregation by demographic variables could be used to do harm in some way [@dignazio_data_2020].
@@
 This review provided a first look at the inclusion, reporting and analysis of demographic variables in the chronobiology and sleep research literature, considering >1000 articles across eight journals.
-We addressed the need to consider individual differences, as well as the dependence of sleep and circadian rhythms ond demographic variables.
-Additionally, we we outline an opportunity to improve reporting of participant-level characteristics using formalised data schemas.
+We addressed the need to consider individual differences, as well as the dependence of sleep and circadian rhythms and demographic variables.
+Additionally, we outline an opportunity to improve reporting of participant-level characteristics using formalized data schemas.
 

### 2021-08-31 article.Rmd
Update article.Rmd
Update article.Rmd
Edit syntax.

@@
 1. Published studies on circadian and sleep physiology should be carefully examined.
-1. Reporting of demographic variables should be done carefully and systematically.
-2. Inclusion and diversity of different populations across the field needs to be ensured.
+2. Reporting of demographic variables should be done carefully and systematically.
+3. Inclusion and diversity of different populations across the field needs to be ensured.
 
@@
 1).
-In addition, the representation of journals in the final list of articles is non-uniform, not least as the included journals will not have been available for the entire data collection period (1979 and onwards).
+In addition, the representation of journals in the final list of articles is non-uniform, as the included journals were not all available from the start date of our data collection (1979 and onwards).
 

### 2022-01-30 article.Rmd
rmd update

@@
 The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored, and any reported aggregate summary statistics for these variables were recorded.
-We found that while >90% of studies reported age or sex, all other variables were reported in <15% of cases.
-Sex balance greatly changed over the years, from a ~3:1 male to female ratio in the 1990s to a near-equal representation in the 2010s.
-We found that the majority of studies report at least sex or age, while other variables are typically not reported.
+We found that while >90% of studies reported age or sex, all other variables were reported in <25% of cases.
 Reporting quality is highly variable, indicating an opportunity to standardize reporting guidelines for participant-level characteristics to facilitate meta analyses.
@@
 library(ggplot2)
-library(ggthemes)
-library(ggExtra)
-library(ggsci)
-library(graphics)
 library(gridExtra)
@@
 library(pracma)
-library(readr)
 library(readxl)
@@
 library(stringr)
-library(tibble)
 library(tidyr)
-library(lemon)
 
@@
 # Set working directory
-#setwd("~/Demographics Review")
-setwd("~/Projects/sleep_circadian_demographics_data")
+setwd("~/Sleep Review/sleep_circadian_demographics_data-main")
 
@@
 In this article, we addressed the question of how representative, diverse and inclusive published research articles in sleep and chronobiology are.
-We analyzed a sample of >1,300 articles published in sleep research and chronobiology journals between 1979 and 2019 for its inclusion, reporting analysis of demographic variables of study participants, including age, sex, gender, ethnicity, level of education, socio-economic status, and profession.
-We found that while >90% of studies reported age or sex, all other variables were reported in <10% of cases, with the frequency of reporting changing over time.
+We analyzed a sample of >1300 articles published in sleep research and chronobiology journals between 1979 and 2019 for its inclusion, reporting and analysis of study population characteristics, including age, sex, gender, ethnicity, level of education, socio-economic status, and profession.
+We found that while >90% of studies reported age or sex, all other variables were reported in <25% of cases, with the frequency of reporting changing over time.
 We identify opportunities for improving the reporting of demographic variables.
@@
 
-Future research needs to identify:
-1. Schemas for reporting demographic variables in a harmonized way across geographical and cultural contexts;
-2. Gaps in the sleep and chronobiology literature with respect to understudied populations;
-3. Understand the extent to which research practices run counter the inclusion of diverse populations in research in all stages of the research cycle, and how this can exacerbate health inequities.
+Future research needs to:
+1. Establish schemas for reporting demographic variables in a harmonized way across geographical and cultural contexts;
+2. Identify gaps in the sleep and chronobiology literature with respect to understudied populations;
+3. Understand the extent to which research practices allow for the inclusion of diverse populations in all stages of the research cycle, and how this can exacerbate health inequities.
 
@@
 
-The extent to which a scientific field's findings are generalizable is a function of the representativeness of a given study sample, and the degree to which findings differ between, e.g., different sexes and ethnicity.
+The extent to which a scientific field's findings are generalizable is a function of the representativeness of a given study sample, and the degree to which findings differ between, e.g., different sexes and ethnicities.
 In the domain of sex, a recent study reviewed the reporting and analysis of sex in biological sciences research [@woitowich_10-year_2020].
 The authors found that while sex inclusion has significantly increased over the past 10 years [@beery_sex_2011], sex-based analysis has not improved, despite recent policies and funder mandates [@clayton_policy_2014].
+Indeed, in 2015 the United States' National Institutes of Health (NIH) issued a notice requiring grant holders to factor sex in the design, analysis and reporting of vertebrate and human studies, or to provide subtsantial justification for the study of a single sex [@nih-2015notice].
 The term "gender data gap" has recently been introduced, demonstrating that women have historically been excluded from biomedical research [@criado-perez_invisible_2020].
 Similarly, the lack of diversity in biomedical and clinical research, and understudy of minorities in the presence of existing health inequities exacerbates these inequities [@oh_diversity_2015].
+Indeed, a recent review of contemporary dementia research reported a lack of demographic, racial, and geographic diversity [@Mooldijk2021].
 
@@
 
-Journal articles published between 1979 and 2019 (odd years) in the top eight sleep and chronobiology journals were considered.
+Journal articles published between 1979 and 2019 in the top eight sleep and chronobiology journals were considered.
+For practical reasons, the number of screened articles was reduced by only analyzing those published in odd years.
 The list of possible target journals was based on a previously established list of journals implementing a hybrid strategy by consulting the Web of Science Master Journal List, domain-relevant expertise in sleep and chronobiology and consulting with a senior researcher with >25 years of experience in the field [@spitschan_transparency_2020].
@@
 All included articles were reviewed for eligibility and coded by RW.
-The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded.
-Funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.
+The reporting of sample size, age, sex, gender, ethnicity, level of education, socio-economic status, and profession of the study population was scored binarily (0 = not reported, 1 = reported), and any reported aggregate summary statistics for these variables were recorded (e.g., mean, median, etc.).
+Sample size referred to the total number of participants for a given study.
+If an article reported multiple studies, then it was analyzed for each of its individual study.
+Age was analyzed when recorded in days, weeks or years.
+Sex referred to biological sex, while gender referred to the social construction of sex.
+Since the language and system describing ethnicity, level of education, socio-economic status and profession differ between countries and individuals, all reporting was taken into account as long as there was a clear indication of what the variable represented.
+Additionally, funding source, geographical location and clinical focus of the article were examined, as well as whether data were analyzed by including any of the demographic variables as covariates.
 Data were coded in an Excel Spreadsheet and analyzed in R Studio (version 4.0.5).
-Reporting of funding, geographical location, and number of sub-studies for each article were investigated for the sample of articles that passed all eligibility criteria.
 
@@
   labs(fill="Journal") +
-  ggtitle("Included and analysed articles by journal and year") +
+  ggtitle("Included and analyzed articles by journal and year") +
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
@@
 data.eligibility <- data.eligibility %>%
-  count(eligibility, article_year) %>%
+  plyr::count(vars=c("eligibility", "article_year")) %>%
   ungroup()
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(eligibility) %>%
-  mutate(per=round(100*n/total,3))
+  mutate(per=round(100*freq/total,3))
 
@@
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
+  scale_y_continuous(breaks=c(0,25,50,75,100), labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
   scale_fill_manual(values=c("#29AF7FFF","#2D708EFF","#D8DE29FF","#482677FF")) +
@@
 We pre-registered our protocol (specified using the PRISMA-P template [@shamseer_preferred_2015; @prisma-p_group_preferred_2015]) on the Open Science Framework (https://osf.io/cu3we/).
+For thorough details on the screening and analysis methods, see the protocol.
 
@@
 All data underlying this manuscript are available on a public GitHub repository (https://github.com/hcvnl/sleep_circadian_demographics_data).
-The article was written in R [@R-base] using RMarkdown and papaja [@R-papaja], employing a series of additional R packages [@R-corrplot2017; @R-dplyr; @R-ggExtra; @R-ggformula; @R-ggplot2; @R-ggridges; @R-ggsci; @R-ggstance; @R-ggthemes; @R-gridExtra; @R-knitr; @R-lattice; @R-latticeExtra; @R-lemon; @R-Matrix; @R-mosaic; @R-mosaicData; @R-pracma; @R-readr; @R-readxl; @R-reshape2; @R-stringr; @R-tibble; @R-tidyr] and is fully reproducible.
+The article was written in R [@R-base] using RMarkdown and papaja [@R-papaja], employing a series of additional R packages [@R-corrplot2017; @R-dplyr; @R-ggplot2; @R-gridExtra; @R-knitr; @R-latticeExtra; @R-mosaic; @R-mosaicData; @R-pracma; @R-readxl; @R-reshape2; @R-stringr; @R-tidyr] and is fully reproducible.
 
@@
 data.funding <- data.funding %>%
-  count(funding, article_year) %>%
+  plyr::count(vars=c("funding", "article_year")) %>%
   ungroup()
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(funding, add=TRUE) %>%
-  mutate(per=round(100*n/total,3))
+  mutate(per=round(100*freq/total,3))
 data.funding$article_year <- as.numeric(data.funding$article_year)
@@
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
+  scale_y_continuous(breaks=c(0,25,50,75,100), labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
   filter(!grepl(",", study_country_values)) %>%
-  count(study_country_values, article_year) %>%
+  plyr::count(vars=c("study_country_values", "article_year")) %>%
   ungroup()
@@
 # Sum rows with 'Other' by year
-data.country <- aggregate(cbind(n) ~ study_country_values + article_year, data=data.country, FUN=sum)
+data.country <- aggregate(cbind(freq) ~ study_country_values + article_year, data=data.country, FUN=sum)
 
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(study_country_values) %>%
-  mutate(per=round(100*n/total,3))
+  mutate(per=round(100*freq/total,3))
 data.country$article_year <- as.numeric(data.country$article_year)
@@
   scale_x_continuous(breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
+  scale_y_continuous(breaks=c(0,25,50,75,100), labels=c("0%","25%","50%","75%","100%"), expand=c(0,0)) +
   scale_fill_viridis_d(direction=-1) +
@@
 
+#data.age.year$article_year <- as.factor(data.age.year$article_year)
+
 # Group by year and compute average for each sex variable
@@
 # Heat map of reporting of age variables by year of publication
-figure6 <- ggplot(data=age.corr.year, aes(x=variable, y=article_year, fill=value)) +
+figure6A <- ggplot(data=age.corr.year, aes(x=variable, y=article_year, fill=value)) +
   geom_tile() +
@@
 # Heat map of reporting of age variables by journal of publication
-figure7 <- ggplot(data=age.corr.journal, 
+figure6B <- ggplot(data=age.corr.journal, 
                    aes(x=variable, y=article_journal, fill=value)) +
@@
 # Violin plot with regression line and equation for mean age by year of publication
-figure8 <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
+figure7 <- ggplot(data=data.age.mean, aes(x=article_year, y=age_reported_mean_value)) +
   geom_smooth(method="lm", color="#440154FF") +
@@
 data.sex <- data.sex %>%
-  count(report, article_year) %>%
+  plyr::count(vars=c("report", "article_year")) %>%
   ungroup()
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(report, add=TRUE) %>%
-  mutate(per=round(100*n/total,3))
+  mutate(per=round(100*freq/total,3))
 data.sex$article_year <- as.numeric(data.sex$article_year)
@@
 # Stacked bar chart of sex reporting by year of publication
-figure9 <- ggplot(data.sex, aes(x=article_year, y=per)) +
+figure8 <- ggplot(data.sex, aes(x=article_year, y=per)) +
   geom_col(aes(fill=factor(report, levels=c("Unspecified",
@@
 
-```{r Other demographics, echo=FALSE, message=FALSE, warning=FALSE}
+```{r Gender, echo=FALSE, message=FALSE, warning=FALSE}
+data.gender <- data %>% 
+  filter(gender_reported_yn==1)
+```
+
+```{r Other demographics, message=FALSE, warning=FALSE, include=FALSE}
 # Select single studies and other demographic variables
@@
 data.demo.education <- data.demo %>%
-  count(education, article_year) %>%
+  plyr::count(vars=c("education", "article_year")) %>%
   ungroup()
 data.demo.ethnicity <- data.demo %>%
-  count(ethnicity, article_year) %>%
+  plyr::count(vars=c("ethnicity", "article_year")) %>%
   ungroup()
 data.demo.profession <- data.demo %>%
-  count(profession, article_year) %>%
+  plyr::count(vars=c("profession", "article_year")) %>%
   ungroup()
 data.demo.ses <- data.demo %>%
-  count(ses, article_year) %>%
+  plyr::count(vars=c("ses", "article_year")) %>%
   ungroup()
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(education, add=TRUE) %>%
-  mutate(per=round(100*n/total,3),
+  mutate(per=round(100*freq/total,3),
          variable="edu") %>%
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(ethnicity, add=TRUE) %>%
-  mutate(per=round(100*n/total,3),
+  mutate(per=round(100*freq/total,3),
          variable="ethni") %>%
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(profession, add=TRUE) %>%
-  mutate(per=round(100*n/total,3),
+  mutate(per=round(100*freq/total,3),
          variable="prof") %>%
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(ses, add=TRUE) %>%
-  mutate(per=round(100*n/total,3),
+  mutate(per=round(100*freq/total,3),
          variable="ses") %>%
@@
 
-figure10 <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
-  geom_point(size=2.5) +
-  geom_line(linetype="dashed") +
+figure9 <- ggplot(data.demo.reported, aes(x=article_year, y=per, color=factor(variable))) +
+  geom_point(size=1.2) +
+  geom_line() +
   theme_apa() +
@@
   scale_x_continuous(limits=c(1979,2019), breaks=c(1979,1989,1999,2009,2019)) +
-  scale_y_continuous(labels=c("0%","25%","50%","75%","100%"), limits=c(0,100)) +
+  scale_y_continuous(breaks=c(0,25,50,75,100), labels=c("0%","25%","50%","75%","100%"), limits=c(0,100)) +
   theme(axis.title=element_text(face="bold", size=12), 
@@
 
+#Education####
 # Compute percentage of articles that report education
@@
 
-figure11A <- ggplot(data=data.education, aes(x=education_reported_numcategory)) +
+figure10A <- ggplot(data=data.education, aes(x=education_reported_numcategory)) +
   geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
@@
   theme_apa() +
-  ylab("Density") +
+  ylab("Proportion of articles") +
   xlab("Education") +
   scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
-  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
-  theme(axis.title=element_text(face="bold", size=12), 
-        axis.text=element_text(colour="black", face="bold"),
+  scale_y_continuous(limits=c(0,0.5), expand=c(0,0), breaks=c(0,0.2,0.4), labels=c("0%","20%","40%")) +
+  theme(axis.title=element_text(face="bold", size=11), 
+        axis.text=element_text(colour="black", face="bold", size=10),
         axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
 
+# compute the education category names into a list
+education.category.list <- tolower(data.education$education_reported_categories)
+
+# check the category combination that is most frequent 
+table(education.category.list)
+
+# compute the number of articles reporting education as a mean
+data.education.mean <- data.education %>%
+  filter(education_reported_mean_yn==1)
+percent.education.mean <- round(nrow(data.education.mean)/nrow(data.education) * 100, 2)
+
+data.education.mean$education_reported_mean <- as.numeric(data.education.mean$education_reported_mean)
+data.education.mean <- data.education.mean %>%
+  filter(!is.na(education_reported_mean))
+mean.education <- mean(data.education.mean$education_reported_mean)
+
+#Ethnicity####
 # Compute percentage of articles that report ethnicity
@@
 
+# Compute percentage of articles where people can select multiple ethnicities out of those that report ethnicity
+data.ethnicity.multiple <- data %>%
+  filter(ethnicity_multiple_categories_yn == 1)
+
+percent.ethnicity.multiple <- round(nrow(data.ethnicity.multiple)/nrow(data.ethnicity) * 100, 2)
+
 # Compute distribution of the number of categories reported for ethnicity
 data.ethnicity$ethnicity_reported_numcategory <- as.numeric(data.ethnicity$ethnicity_reported_numcategory)
-figure11B <- ggplot(data=data.ethnicity, aes(x=ethnicity_reported_numcategory)) +
+figure10B <- ggplot(data=data.ethnicity, aes(x=ethnicity_reported_numcategory)) +
   geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
@@
   theme_apa() +
-  ylab("Density") +
+  ylab("Proportion of articles") +
   xlab("Ethnicity") +
   scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
-  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
-  theme(axis.title=element_text(face="bold", size=12), 
-        axis.text=element_text(colour="black", face="bold"),
+  scale_y_continuous(limits=c(0,0.5), expand=c(0,0), breaks=c(0,0.2,0.4), labels=c("0%","20%","40%")) +
+  theme(axis.title=element_text(face="bold", size=11), 
+        axis.text=element_text(colour="black", face="bold", size=10),
         axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
 
+# compute the ethnicity category names into a list
+ethnicity.category.list <- tolower(data.ethnicity$ethnicity_reported_categories)
+
+# check the category combination that is most frequent 
+table(ethnicity.category.list)
+
+# split the category combinations and convert to lower case letters
+ethnicity.category.list <- as.list(strsplit(ethnicity.category.list, ","))
+ethnicity.category.list <- unlist(ethnicity.category.list)
+
+# count the number of occurences for each category
+table(ethnicity.category.list)
+
+# count the number of occurrences for "white"/"caucasian", "african american"/"black", and "other"
+n.ethnicity.white <- sum(str_count(ethnicity.category.list, "white"),str_count(ethnicity.category.list, "caucasian"),str_count(ethnicity.category.list, "non hispanic white"),str_count(ethnicity.category.list, "white caucasian"),str_count(ethnicity.category.list, "white american"))
+n.ethnicity.black <- sum(str_count(ethnicity.category.list, "black"),str_count(ethnicity.category.list, "african american"),str_count(ethnicity.category.list, "african-american"),str_count(ethnicity.category.list, "non hispanic black"))
+n.ethnicity.other <- sum(str_count(ethnicity.category.list, "other"))
+
+# compute percentage of articles where white vs black is reported as an ethnicity
+percent.ethnicity.white <- round(n.ethnicity.white/nrow(data.ethnicity) * 100, 2)
+percent.ethnicity.black <- round(n.ethnicity.black/nrow(data.ethnicity) * 100, 2)
+percent.ethnicity.other <- round(n.ethnicity.other/nrow(data.ethnicity) * 100, 2)
+
+#Profession####
 # Compute percentage of articles that report profession
@@
 data.profession$profession_reported_numcategory <- as.numeric(data.profession$profession_reported_numcategory)
-figure11C <- ggplot(data=data.profession, aes(x=profession_reported_numcategory)) +
+figure10C <- ggplot(data=data.profession, aes(x=profession_reported_numcategory)) +
   geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
@@
   theme_apa() +
-  ylab("Density") +
+  ylab("Proportion of articles") +
   xlab("Profession") +
   scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
-  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
-  theme(axis.title=element_text(face="bold", size=12), 
-        axis.text=element_text(colour="black", face="bold"),
+  scale_y_continuous(limits=c(0,0.5), expand=c(0,0), breaks=c(0,0.2,0.4), labels=c("0%","20%","40%")) +
+  theme(axis.title=element_text(face="bold", size=11), 
+        axis.text=element_text(colour="black", face="bold", size=10),
         axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
 
+# compute the profession category names into a list
+profession.category.list <- tolower(data.profession$profession_reported_categories)
+
+# check the category combination that is most frequent 
+table(profession.category.list)
+
+# split the category combinations
+profession.category.list <- as.list(strsplit(profession.category.list, ","))
+profession.category.list <- unlist(profession.category.list)
+
+# count the number of occurences for each category
+table(profession.category.list)
+
+# count the number of occurrences for "unemployed"
+n.profession.unemployed <- sum(str_count(profession.category.list, "unemployed"))
+
+# compute percentage of articles where this is reported
+percent.profession.unemployed <- round(n.profession.unemployed/nrow(data.profession) * 100, 2)
+
+#SES####
 # Compute percentage of articles that report ses
 data.ses <- data %>%
-  filter(ses_reported_yn ==1)
+  filter(ses_reported_yn==1)
 percent.ses <- round(nrow(data.ses)/nrow(data) * 100, 2)
@@
 data.ses$ses_reported_numcategory <- as.numeric(data.ses$ses_reported_numcategory)
-figure11D <- ggplot(data=data.ses, aes(x=ses_reported_numcategory)) +
+figure10D <- ggplot(data=data.ses, aes(x=ses_reported_numcategory)) +
   geom_histogram(binwidth=1, aes(y=..density..), colour="black", fill="#2D708EFF") +
@@
   theme_apa() +
-  ylab("Density") +
+  ylab("Proportion of articles") +
   xlab("SES") +
   scale_x_continuous(limits=c(0.5,8.5), breaks=c(1,2,3,4,5,6,7,8)) +
-  scale_y_continuous(limits=c(0,0.5), expand=c(0,0)) +
-  theme(axis.title=element_text(face="bold", size=12), 
-        axis.text=element_text(colour="black", face="bold"),
+  scale_y_continuous(limits=c(0,0.5), expand=c(0,0), breaks=c(0,0.2,0.4), labels=c("0%","20%","40%")) +
+  theme(axis.title=element_text(face="bold", size=11), 
+        axis.text=element_text(colour="black", face="bold", size=10),
         axis.line=element_line(size=0.8), axis.ticks=element_line(size=0.8))
+
+# compute the ses category names into a list
+ses.category.list <- tolower(data.ses$ses_reported_categories)
+
+# check the category combination that is most frequent 
+table(ses.category.list)
+
+# compute percentage of articles where ses is based on income with numbers
+percent.ses.income <- 100 - round(21/nrow(data.ses) * 100, 2)
 ```
@@
 data.analysis <- data.analysis %>%
-  count(report, article_year) %>%
+  plyr::count(vars=c("report", "article_year")) %>%
   ungroup()
@@
   group_by(article_year) %>%
-  mutate(total=sum(n)) %>%
+  mutate(total=sum(freq)) %>%
   group_by(report, add=TRUE) %>%
-  mutate(per=round(100*n/total,3))
+  mutate(per=round(100*freq/total,3))
 data.analysis$article_year <- as.numeric(data.analysis$article_year)
@@
 # Stacked bar chart of the use of demographic variables in the analysis by year of publication
-figure12 <- ggplot(data.analysis, aes(x=article_year, y=per)) +
+figure11 <- ggplot(data.analysis, aes(x=article_year, y=per)) +
   geom_col(aes(fill=report)) +
@@
 Out of `r as.numeric(nrow(data.original))` identified and pre-screened articles, we included and extracted data from `r as.numeric(nrow(data.included))` (`r round(percent.included)`%).
-The distribution of years in which the articles were published is non-uniform and we included and extracted data from more recent articles (Fig.
+The distribution of years in which the articles were published is non-uniform and the proportion of included and extracted data is greater for more recent articles (Fig.
 1).
-In addition, the representation of journals in the final list of articles is non-uniform, as the included journals were not all available from the start date of our data collection (1979 and onwards).
-
-```{r, fig.cap="Included and analyzed articles by year and journal. More recent articles are more represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
+In addition, the representation of journals in the final list of articles is also non-uniform, as the included journals were not all available from the start date of our data collection (1979 onwards).
+
+```{r, fig.cap="Included and analyzed articles by year and journal of publication.
+Recent articles are more widely represented, reflecting an overall increase in scientific output.", warning=FALSE, message=FALSE}
 figure1
@@
 
-```{r, fig.cap="Normalised distributions of included and excluded articles.", warning=FALSE, message=FALSE}
+```{r, fig.cap="Percentage of included and excluded articles by year of publication, with reasons for exclusion.", warning=FALSE, message=FALSE}
 figure2
@@
 Overall, `r nrow(country.single)` countries were represented.
-`r round(percent.country.multiple)`% of articles reported multiple countries of study.
 Figure 4 shows the distribution of study location across time with the eight most represented countries.
-Across all years, the US is the most represented country.
+Across all years, the US is the most represented country and only three studies were primarily conducted in Africa (Nigeria, Senegal and Tunisia).
 
@@
 
-  We examined whether sample size was reported in studies, and the sample sizes reported.
+  We examined whether sample size was reported in studies, and the numbers recorded.
 Overall, sample size was reported in `r round(percent.ss)`% of studies.
@@
 
-  `r round(percent.age)`% of articles reported a variable describing age.
+  `r round(percent.age)`% of articles reported a variable describing age, such as the mean, standard deviation, median, minimum, maximum and interquartile range.
+Overall, the median and interquartile range were the least reported variables for sex (Fig.
+6).
+The minimum and maximum age were the most employed variables in 1979, while their reporting decreased throughout the years compared to other variables.
+On another hand, the standard deviation of the mean was increasingly reported throughout the years.
+Trends for specific journals were also observed, such as the substantial use of the min/max variables in *Journal of Pineal Research*, and the lack of in *Journal of Clinical Sleep Medicine*.
+
+```{r, warning=FALSE, message=FALSE}
+figure6A
+```
+
+```{r, fig.cap="Reporting of various age variables by year (top) and journal (bottom) of publication. SD = standard deviation of the mean, IQR = interquartile range.", warning=FALSE, message=FALSE}
+figure6B
+```
+
   Overall, the average mean age of the study populations was `r round(mean(data.age.mean$age_reported_mean_value))` years old.
-We examined the extent to which the mean age across studies differed widely as a function of publication year (Fig.
-6), and found that the mean age is much more widely varied in later years, likely reflecting the extent of considering study samples that are more varied in age.
+We examined the extent to which the mean age differed across studies as a function of publication year (Fig.
+7), and found that the mean age is much more widely varied in later years, likely reflecting the extent of considering study samples that are more diversified in age.
 
 ```{r, fig.cap="Evolution of mean age in included studies as a function of publication year. Fit shown is a linear fit (±95%CI).", warning=FALSE, message=FALSE}
+figure7
+```
+
+## Sex & Gender
+
+  Sex was reported in `r round(percent.sex)`% of the studies.
+In Figure 8, we show the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants.
+Overall, `r round(percent.sex.male)`% of the studies reporting sex only recruited male participants, while `r round(percent.sex.female)`% only employed females.
+Out of the studies focusing on a single sex, `r round(percent.sex.male.focus)`% of the male studies focused on a sex dependent feature, while `r round(percent.sex.female.focus)`% of the female studies did.
+`r round(percent.sex.age)`% of studies reported age by sex.
+
+  Gender was reported in only one article, reflecting the lack of inclusion for gender identity. In the reporting gender, participants were described as male, female or transgender.
+
+```{r, fig.cap="Sex inclusion by year. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
 figure8
@@
 
-## Sex
-
-  Sex was reported in `r round(percent.sex)`% of the studies.
-In Figure 7, we show the proportion of studies that recruited male subjects, female subjects, both sexes or did not specify the sex of the participants.
-`r round(percent.sex.male)`% of the studies reporting sex only recruited male participants, while `r round(percent.sex.female)`% only employed females.
-Out of the studies focusing on a single gender, `r round(percent.sex.male.focus)`% of the male studies focused on a sex dependent feature, while `r round(percent.sex.female.focus)`% of the female studies did.
-`r round(percent.sex.age)`% of studies reported age by sex.
-
-```{r, fig.cap="Sex inclusion by year. Proportion of studies that recruited male subjects, female subjects, both sexes, or did not specify the sex of the participants.", warning=FALSE, message=FALSE}
+##  Ethnicity, education, profession and socio-economic status
+
+  We examined the reporting of further demographic variables, including ethnicity, education, socio-economic status and profession.
+These demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.ses)`% for socio-economic status and `r round(percent.profession)`% for profession.
+Figure 9 shows the distribution of this reporting across the years.
+Qualitatively, there is a clear increase in the reporting of additional demographic variables over time, with ethnicity being the most reported demographic variable across articles in more recent years.
+Participants could choose multiple categories for ethnicity in `r round(percent.ethnicity.multiple)`% of the studies reporting this variable.
+The mean number of years of education was reported in `r round(percent.education.mean)`% of articles, and was `r round(mean.education)` years of study overall.
+
+```{r, fig.cap="Reporting of education, ethnicity, profession and socio-economic status by year of publication.", warning=FALSE, message=FALSE}
 figure9
@@
 
-##  Ethnicity, education, profession and socio-economic status
-
-  We examined the reporting of the demographic variables ethnicity, education, profession and socio-economic status.
-These demographics variables were reported in `r round(percent.education)`% of studies for education, `r round(percent.ethnicity)`% for ethnicity, `r round(percent.profession)`% for profession, and `r round(percent.ses)`% for socio-economic status.
-Figure 8 shows the distribution of this reporting across the years.
-Qualitatively, there is a clear increase in the reporting of additional demographic variables over time, with ethnicity being the most reported demographic variable across articles in more recent years.
-
-```{r, fig.cap="test", warning=FALSE, message=FALSE}
-figure10
-```
   We furthermore examined the number of categories included for each of these demographic variables.
-In Figure 9, we show the number of categories reported for each variable among those articles that included it in a histogram.
-On overage, we find that the coding scheme for each of these variables in the included articles comprised between three and four variables.
-
-```{r, fig.cap="Number of categories reported for education, ethnicity, profession and socio-economic status (SES). Yellow vertical line corresponds to the median.", warning=FALSE, message=FALSE}
-grid.arrange(figure11A, figure11B, figure11C, figure11D, nrow=2, top="Number of reported categories")
+In Figure 10, we show the number of categories reported for each variable among those articles that included it.
+On overage, we find that the coding scheme for each of these comprised between three and four variables.
+Education was reported as a range of years of study, degree level, or a subjective measure such as "low, medium, high".
+The most common categories for ethnicity were "white/caucasian", "black/african american" and "other", and were mentioned in `r round(percent.ethnicity.white)`%, `r round(percent.ethnicity.black)`% and `r round(percent.ethnicity.other)`% of studies respectively.
+The most common category for profession was "unemployed", mentionned in `r round(percent.profession.unemployed)`% of studies, along with a variety of specific jobs, such as attorney, farmer and astronaut.
+In `r round(percent.ses.income)`% of studies reporting socio-economic status, the categories of SES were based on range of income.
+The rest of the articles reported SES in subjective measures, such as "high, medium, low".
+
+```{r, fig.cap="Distribution of the number of categories reported for education, ethnicity, profession and socio-economic status (SES).
+The yellow vertical line corresponds to the median.", warning=FALSE, message=FALSE}
+grid.arrange(figure10A, figure10B, figure10C, figure10D, nrow=2, top="Number of reported categories")
 ```
@@
   We considered whether a given article reported on a specific, pre-defined group of people.
-We find `r round(percent.focus.sex)`% of articles focused on a sex dependent feature, while `r round(percent.focus.clinical)`% investigated a clinical feature.
+We found that `r round(percent.focus.sex)`% of articles focused on a sex dependent feature, while `r round(percent.focus.clinical)`% investigated a clinical feature.
 `r round(percent.focus.twins)`% of studies focused on twins, `r round(percent.focus.pregnant)`% on pregnant women, `r round(percent.focus.shift)`% on shift workers and `r round(percent.focus.students)`% on university students.
@@
 Over time (Fig.
-9), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed.
+11), we see a distinct evolution of the extent to which subgroup analyses of the study sample were performed.
 The most common subgroup analyses involve disaggregating by sex, age, or both.
@@
 ```{r, fig.cap="Use of study population characteristics as variables in the analysis."}
-figure12
+figure11
 ```
@@
 
-Here, we considered the inclusion, reporting of `r as.numeric(nrow(data.included))` articles in chronobiology and sleep research, sampled from eight journals.
-More articles from more recent years are included, with articles originating in the US representing the large group.
-Over time, we found an increase in the range of sample sizes.
-We furthermore found a wider spread of average ages included in articles.
-Across the included articles, we observed an increase in the reporting of demographic variables (ethnicity, education, professional and socioeconomic status) over time, with ethnicity being the most-reported variable in recent years.
-Of studies included these demographic variables, the number of categories varied widely between studies.
+Here, we considered the inclusion and reporting of `r as.numeric(nrow(data.included))` articles in chronobiology and sleep research, sampled from eight journals.
+Articles from recent years were more widely included, with articles originating in the US representing the largest group.
+While studies conducted in Western countries in North America and central Europe were predominant, those in Africa only represented 0.3% of all screened articles, underlining a gap in geographical inclusion.
+Over time, we found an increase in the range of sample sizes, as well as a wider spread of average ages included in the studies.
+We also observed an increase in the reporting of demographic variables (ethnicity, education, professional and socioeconomic status), with ethnicity being the most-reported variable in recent years.
+A wide variety of reporting strategies, which involve both objective and subjective measures, was also observed.
 
@@
 
-The ability to generalize findings from the scientific literature to wide and diverse populations of people hinges upon the representativeness of the study sample with respect to demographic categories.
-The question to what extent the composition of a given study sample can make the generalisability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020].
-In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020], but to our knowledge, this review represents a first look at the inclusion, reporting and analysis of participant demographics in chronobiology and sleep research.
+The ability to generalize findings from the scientific literature to wide and diverse populations hinges upon the representativeness of the study sample with respect to demographic categories.
+The question to what extent the composition of a given study sample can make the generalizability of findings difficult or impossible has received attention in the field of psychology, where many articles published in prominent journals reflected participants from WEIRD (Western, Educated, Industrialized, Rich, and Democratic) contexts [@henrich_weirdest_2010; @muthukrishna_beyond_2020].
+In other fields, analyses similar to the one in the present review have been published [@obryant_reporting_2004; @sifers_reporting_2002; @jones_reporting_2020;@Mooldijk2021], but to our knowledge, this review represents a first look at the inclusion, reporting and analysis of participant demographics in chronobiology and sleep research.
 
@@
 Our review only concerned articles from a subset of eight specialized journals.
-As a consequence, the included articles were necessarily published in these journals, ignoring relevant articles published in other specialized journals (such as those included in the list of candidate journals), and articles published in other, including interdisciplinary journals.
-This raises the question to what extent we may have missed a section of the literature that would have been relevant to include here.
+As a consequence, the included articles were necessarily published in these journals, ignoring relevant articles published in other specialized journals (such as those included in the list of candidate journals), and articles published in other interdisciplinary journals.
+This raises the question to what extent we may have missed a section of the literature that would have been relevant to include in this study.
 As an alternative strategy, we considered randomly sampling a subset of chronobiology and sleep research articles produced by a general search (e.g.
@@
 The FAIR principles state that data should be findable, accessible, interoperable and reusable [@wilkinson_fair_2016], and one way of realizing these criteria is the use of data schemas which could prescribe categories of data and common naming schemes for reporting participant characteristics.
-Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extent such data schema may be exclusionary, e.g., by enforcing gender binaries [@hyde_future_2019], and whether any specific demographic variable is truly important (following the principle of data minimization).
+Importantly, however, "what gets counted counts" [@dignazio_data_2020], and it will be imperative to understand to what extent such data schema may be exclusionary, e.g., by enforcing sex binaries [@hyde_future_2019], and whether any specific demographic variable is truly important (following the principle of data minimization).
 A further central point to consider is the extent to which disaggregation by demographic variables could be used to do harm in some way [@dignazio_data_2020].
@@
 
-This review provided a first look at the inclusion, reporting and analysis of demographic variables in the chronobiology and sleep research literature, considering >1000 articles across eight journals.
-We addressed the need to consider individual differences, as well as the dependence of sleep and circadian rhythms and demographic variables.
+This review provides a first look at the inclusion, reporting and analysis of demographic variables in the chronobiology and sleep research literature, considering >1000 articles across eight journals.
+We address the need to consider individual differences, as well as the dependence of sleep and circadian rhythms on demographic variables.
 Additionally, we outline an opportunity to improve reporting of participant-level characteristics using formalized data schemas.
